### 2018-04-24 eyeIntegration_app_manuscript.Rmd
first commit

---
title: Test

author: |
  | David M McGaughey 1^[Corresponding author: mcgaugheyd@mail.nih.gov] $^1$, John M Bryan $^1$
  | $^1$OGVFB NEI NIH

abstract: |
  eyeIntegration is an series of open databases with an extensive transcriptomic database containing over 30 GTEx human body tissues as well as major components of the eye: retina, RPE, RPE-choroid, and cornea. eyeIntegration also holds two WGCNA networks for retina and RPE-choroid. areWeb application con

bibliography: Unified Gene Expression.bib
csl: genome-biology.csl

output:
  bookdown::pdf_document2:
  toc: no
fig_caption: yes

always_allow_html: yes

# These are LaTex settings to take care of floating figures/tables, line spacing, etc
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)

# I usually load my libraries up front to keep things organized
library(bookdown)
#library(knitr)
#library(kableExtra)
library(tidyverse)

```

### 2018-04-24 eyeIntegration_app_manuscript.Rmd
test files created

@@
 title: Test
-
 author: |
+  | $^1$OGVFB NEI NIH
   | David M McGaughey 1^[Corresponding author: mcgaugheyd@mail.nih.gov] $^1$, John M Bryan $^1$
-  | $^1$OGVFB NEI NIH
-
 abstract: |
@@
 eyeIntegration also holds two WGCNA networks for retina and RPE-choroid.
-areWeb application con
-
+Web application.
 bibliography: Unified Gene Expression.bib
 csl: genome-biology.csl
-
 output:
   bookdown::pdf_document2:
-  toc: no
+  toc: yes
+  word_document: default
 fig_caption: yes
-
 always_allow_html: yes
-
-# These are LaTex settings to take care of floating figures/tables, line spacing, etc
 header-includes:
@@
 ```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
-knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
+#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
 knitr::opts_chunk$set(out.width = '100%', dpi=300)
@@
 library(tidyverse)
+library(citr)
+```
 
-```
+# Introduction
+
+# Materials and Methods
+
+## 
+## Single Cell RNA-Seq Data
+[@macoskoHighlyParallelGenomewide2015]
+
+# Results
+
+# Conclusions

### 2019-01-17 draft-style.Rmd
template messing around

---
title: 'Eye on a Disk: eyeIntegration version 1.00'
author: "David McGaughey"
date: "1/17/2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-01-17 eyeIntegration_app_manuscript.Rmd
demo table and fig
template messing around

@@
 output:
-  bookdown::pdf_document2:
-  toc: yes
-  word_document: default
+  word_document:
+    reference_docx: word-styles-reference-01.docx 
 fig_caption: yes
@@
 
+```{r, fig.cap=paste('Density of num'), echo=F}
+library(tidyverse)
+rand <- sample(1:100, 100, replace=F)
+rand %>% density() %>% plot()
+```
 # Materials and Methods
 
-## 
+Cell | Number
+----- | -----
+RPE | 200
+Fibroblast | 30
+
 ## Single Cell RNA-Seq Data

### 2019-01-17 manuscript.Rmd
template messing around

---
title: 'Eye in a Disk: eyeIntegration version 1.00'
author: |
  | $^1$OGVFB NEI NIH
  | David M McGaughey 1^[Corresponding author: mcgaugheyd@mail.nih.gov] $^1$, Vinay Swamy $^1$
date: "1/17/2019"
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
---

My report starts here

### 2019-01-17 test2/test2.Rmd
template messing around

---
title: Short Paper
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: Corresponding Author
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
---

_Text based on elsarticle sample manuscript, see [http://www.elsevier.com/author-schemas/latex-instructions#elsarticle](http://www.elsevier.com/author-schemas/latex-instructions#elsarticle)_

The Elsevier article class
==========================

#### Installation

If the document class *elsarticle* is not available on your computer,
you can download and install the system package *texlive-publishers*
(Linux) or install the LaTeX package *elsarticle* using the package
manager of your TeX installation, which is typically TeX Live or MikTeX.

#### Usage

Once the package is properly installed, you can use the document class
*elsarticle* to create a manuscript. Please make sure that your
manuscript follows the guidelines in the Guide for Authors of the
relevant journal. It is not necessary to typeset your manuscript in
exactly the same way as an article, unless you are submitting to a
camera-ready copy (CRC) journal.

#### Functionality

The Elsevier article class is based on the standard article class and
supports almost all of the functionality of that class. In addition, it
features commands and options to format the

-   document style

-   baselineskip

-   front matter

-   keywords and MSC codes

-   theorems, definitions and proofs

-   lables of enumerations

-   citation style and labeling.

Front matter
============

The author names and affiliations could be formatted in two ways:

(1) Group the authors per affiliation.

(2) Use footnotes to indicate the affiliations.

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

Bibliography styles
===================

There are various bibliography styles available. You can select the
style of your choice in the preamble of this document. These styles are
Elsevier styles based on standard styles like Harvard and Vancouver.
Please use BibTeX to generate your bibliography and include DOIs
whenever available.

Here are two sample references: @Feynman1963118 [@Dirac1953888].

References {#references .unnumbered}
==========

### 2019-01-17 test2/test2.tex
template messing around

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{An awesome journal} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short Paper},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Short Paper}
    \author[Some Institute of Technology]{Alice Anonymous\corref{c1}}
   \ead{alice@example.com} 
   \cortext[c1]{Corresponding Author}
    \author[Another University]{Bob Security}
   \ead{bob@example.com}

      \address[Some Institute of Technology]{Department, Street, City, State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}

 \end{frontmatter}

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

\section{The Elsevier article class}\label{the-elsevier-article-class}

\paragraph{Installation}\label{installation}

If the document class \emph{elsarticle} is not available on your
computer, you can download and install the system package
\emph{texlive-publishers} (Linux) or install the LaTeX package
\emph{elsarticle} using the package manager of your TeX installation,
which is typically TeX Live or MikTeX.

\paragraph{Usage}\label{usage}

Once the package is properly installed, you can use the document class
\emph{elsarticle} to create a manuscript. Please make sure that your
manuscript follows the guidelines in the Guide for Authors of the
relevant journal. It is not necessary to typeset your manuscript in
exactly the same way as an article, unless you are submitting to a
camera-ready copy (CRC) journal.

\paragraph{Functionality}\label{functionality}

The Elsevier article class is based on the standard article class and
supports almost all of the functionality of that class. In addition, it
features commands and options to format the

\begin{itemize}
\item
  document style
\item
  baselineskip
\item
  front matter
\item
  keywords and MSC codes
\item
  theorems, definitions and proofs
\item
  lables of enumerations
\item
  citation style and labeling.
\end{itemize}

\section{Front matter}\label{front-matter}

The author names and affiliations could be formatted in two ways:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Group the authors per affiliation.
\item
  Use footnotes to indicate the affiliations.
\end{enumerate}

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

\section{Bibliography styles}\label{bibliography-styles}

There are various bibliography styles available. You can select the
style of your choice in the preamble of this document. These styles are
Elsevier styles based on standard styles like Harvard and Vancouver.
Please use BibTeX~to generate your bibliography and include DOIs
whenever available.

Here are two sample references: Feynman and Vernon Jr. (1963; Dirac
1953).

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-Dirac1953888}{}
Dirac, P.A.M. 1953. ``The Lorentz Transformation and Absolute Time.''
\emph{Physica} 19 (1---12): 888--96.
doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}.

\hypertarget{ref-Feynman1963118}{}
Feynman, R.P, and F.L Vernon Jr. 1963. ``The Theory of a General Quantum
System Interacting with a Linear Dissipative System.'' \emph{Annals of
Physics} 24: 118--73.
doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}.

\end{document}

### 2019-01-17 manuscript.Rmd
lua code to generate author blocks for docx, improved formatting, vinny intro added, start of citations

@@
 title: 'Eye in a Disk: eyeIntegration version 1.00'
-author: |
-  | $^1$OGVFB NEI NIH
-  | David M McGaughey 1^[Corresponding author: mcgaugheyd@mail.nih.gov] $^1$, Vinay Swamy $^1$
+author:
+  - Vinay Swamy:
+      institute: ogvfb
+  - David McGaughey:
+      institute:
+        - ogvfb
+      correspondence: "yes"
+      email: mcgaugheyd@mail.nih.gov
+institute:
+  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
 date: "1/17/2019"
@@
     reference_docx: word-styles-reference-01.docx
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+bibliography: eyeIntegration_v1_app_manuscript.bib
 ---
 
-My report starts here
+# Abstract
+
+# Introduction
+
+  RNA sequencing (RNA-seq) has become a valuable tool for understanding the role of gene expression in biology.
+RNA-seq provides an accurate estimate of the expression of all genes in a given biological sample.
+By examining gene expression in this concerted manner researchers glean new insight into the complex regulatory mechanisms that affect gene expression and better understand the effect disease has on these mechanisms.
+For example, researchers used RNA-seq to establish the effects of the gene NRL on the transcriptomic regulatory network of developing photoreceptors.
+(Swaroop paper).
+Some studies have generated large RNA-seq data sets to accurately determine gene expression specific to individual tissues in the body.
+However, such studies require enormous amounts of funding, due to the large cost of RNA-seq and large number of samples required for these studies.
+Some major institutions like the NIH have performed such studies and made the data publicly available.
+For example, the GTEX project generated a large amount of RNA-seq data in order to determine gene expression ins 55 Human tissues and published the results in an online web-app [@gtexconsortiumGenotypeTissueExpressionGTEx2013].
+This data serves as a valuable tool for researchers across the world, but lacks relevance to eye research as GTEX failed to cover the eye.
+  Another solution to the problem of creating large, high quality RNA-seq data sets is manual curation of data generated by multiple studies.
+There is a vast amount of publicly available RNA-seq data in the Sequence Read Archive (SRA), where investigators across the world deposit their raw sequencing data.
+As of January 17th 2019, the SRA has 493,821 human RNA-seq samples.
+Creating a curated dataset from multiple data sets presents a difficult challenge.
+Samples must be carefully selected and labelled with the proper metadata, evaluated for quality, and normalized to account for technical variation between studies.
+While many fields of research have made such databases (Other curated database), a robust database does not exist in vision science research.
+Previous attempts to make such a database relied on outdated technology like microarrays (iSyte), or had too narrow of focus (TIGEM).
+To remedy this problem, we have created Eyeintegration, the largest curated RNA-seq database for ocular research.
+We provide accurate information about gene expression in multiple human eye tissues using published studies from the SRA.
+Additionally, we incorporate murine single cell RNA-seq data to provide researchers access to information about retinal development.
+The database is built using a highly reproducible and robust pipeline that includes a rigorous quality control procedure and automated publishing of the website.
+The database contains both gene and transcript expression data for the retina, retinal pigmented epithelium (RPE), cornea, lens and 44 other human organs.
+We performed multiple differential expression analysis between tissues types and provide a wide range of data visualizations that can be accessed through an easy-to-use web-app.
+The database can be found at Eyeintegration.nei.nih.gov
+
+# Methods
+
+# Results
+
+# Conclusions
+
+# Bibliography

### 2019-01-17 manuscript.Rmd
pasted more vinny writing in

@@
 
-  RNA sequencing (RNA-seq) has become a valuable tool for understanding the role of gene expression in biology.
-RNA-seq provides an accurate estimate of the expression of all genes in a given biological sample.
+  RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome. Dozens of groups have used this technology in human samples to probe the normal 
   By examining gene expression in this concerted manner researchers glean new insight into the complex regulatory mechanisms that affect gene expression and better understand the effect disease has on these mechanisms.
@@
 # Methods
+## We find X potential eye studies
+  We searched the Sequence Read Archive^(SRA) for eye related tissues using the query ‘cornea|retina|RPE|ocular|eye’.
+We hand selected relevant studies and selected healthy, control/unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids and cell lines.
+In order to compare gene expression in the eye against expression in other body tissues, we obtained samples for 44 different body tissues from the GTEx project, also housed in the SRA.
+Using SRA metadata from each study we extracted sample and run accessions, library type , tissue of origin, and subtissue of origin.
+Any of the preceding information missing from the SRA metadata was added by hand.
+Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.
+## Data Gathering 
+  We downloaded the relevant SRA files for each sample directly from the NCBI ftp server using the file transfer software Aspera.
+SRA files were converted to FASTQ format using the tool fastq-dump from the SRAtoolkit software package^.
+Samples only available in the BAM format were converted to FASTQ format using SAMTOOLS^.
+Samples were quantified using the alignment free quantification software Salmon, using transcriptomic index built from  gencode v28 protein coding transcript sequences^ using the transcriptomic aligner Salmon^.
+Using the resulting expression data, we identified lowly or unused transcripts within annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene.^(sonneson et al).
+Samples were requantified against a transcriptomic index built on the filtered transcript sequences.
+All expression values in length-scaled TPM counts were combined into a single matrix and aggregated to the gene level using tximport^.
 
+## Quality control 
+We began the quality control procedure by removing samples with a Salmon mapping rate less than 50%.
+We further filtered samples by calculating the median gene expression for each sample and only kept sample that had a median count >2 across all genes.
+We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count>200 across all samples.
+To reduce the noise experimental variability between each study, we normalized samples by library using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth^(Hicks et al).
+We generated sub-tissue clusters by first reduced the dimensionality of the data to two dimensions using the t-Stochastic Neighbor Embedding(t-SNE) algorithm, and then calculated the center of the resulting sub-tissue clusters by finding the average coordinate of all samples  of a given sub-tissue type.
+We used these sub-tissues clusters to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type.
+We calculated the distance of each sample from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance.
+
+## Analysis and generation of web app 
+We used the cleaned data to determine differential gene expression between different sub-tissue types.
+First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
+We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test.
+We extracted significant genes (p <.05) for each eye sub-tissue vs body comparisons and used these to calculate GO enrichment for each eye sub-tissue.
+The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler  R-package (q-value < .01).
+
+VS: 
+I’ll leave the webapp and single cell stuff for you since I didn’t really work on that part 
 # Results
 
+## Data collection and Quality control 
+Our initial query to the SRA found 44* potentially relevant studies.
+We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total of <x> studies and 410 samples, with 196 Retina, 143 RPE, 62 Cornea, and 9 lens samples.
+We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake^, a workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the web-app.
+In order to create a robust and clean data set, we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same tissue type.
+We used t-SNE to cluster samples passing quality controls by tissue type to observe variability within samples of the same tissue and determine the strength of our quality control procedure.
+(figure 1A) We found that samples of the same tissue type and origin cluster well together.
+For example, in retina derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal tissue cluster together.
+Our ability to uniformly cluster data by known biological function independent of study origin demonstrates our quality control method’s ability to account for technical variation between studies.
+We provide plots of t-SNE clusters at variable perplexities within the web-app under the ‘2D-clustering’ tab.
+After quality control we found 310 samples to be of sufficient quality, with 180 Retina, 117 RPE, 42 Cornea, and 6 Lens samples included in downstream analysis.
+The 2019 version of the web-app represents a significant change from our previously published Eyeintegration 2017 data set.
+(Table 1a/Figure 1b) We have added 80 more Retina samples, 61 more RPE samples, 22 more cornea samples, and added Lens as a new tissue type, across <x> more studies.
+We have increased the number of sub-tissue types, adding Retinal fetal tissue, retinal organoids and cornea stem cell lines.
+(Fig.
+1A).
+There has been significant change to the types of gene classes.
+We have added miRNA, lincRNA, pseudogenes, and snRNA as gene classes.
+The web app contains 151222 protein coding, 13188 lincRNA, 10202 pseudogenes, 1900 snRNA and 1881 miRNA transcripts, for a net addition of X more genes and Y more transcripts.
+(Table 1b?).
+
+## eyeIntegration web application provides interactive visual portal to all data 
+All data generated by our analysis is presented through the Eyeintegration web-app.
+The web-app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
+The web-app is divided into 3 general categories: gene/transcript expression, differential comparisons, and single cell data.
+
+## Expression 
+The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues.
+(Fig 2) The user first selects which version of the web-app to use data from, as well as genes and tissues to examine.
+A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue.
+The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.
+The web app provides multiple options for visualizing expression data.
+Box plots can be drawn to compare gene expression across tissues.
+Users may select any number of genes and tissues to compare at a given time, and can arrange visualizations in multiple ways.
+Box plots provide the user a measure of confidence in the variability of reported gene expression levels.
+The metadata behind any given samples can be seen by hovering over a point in the box plot, including study of origin, sex of sample, and SRA accession.
+Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues.
+The visualizations can also be generated at the transcript level by selecting pan-transcript expression under the expression tab.
+Finally, a session can be saved or shared by building a custom link for the session with the build URL shortcut button.
+
+## Differential expression and Gene Ontology data. 
+Using our expression data, we performed multiple differential comparisons at the sub-tissue level.
+We performed differential tests of eye tissues against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTeX data set, allowing users to quickly identify eye specific genes.
+We also performed differential tests between individual eye tissues.
+The user can view the results under the ‘differential’ selection under the ‘expression’ tab.
+(fig 3) As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons.
+The user additionally has the option to select different gene classes to examine, protein coding, lincRNA etc.
+The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values.
+The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format.
+Differential expression can be visualized through fold change bar graphs in the ‘pan-tissue plots’ section under expression.
+Additionally, we performed GO enrichment for all differential comparisons.
+Enriched GO terms are presented first as a word cloud, for quick comparison of GO enrichment.
+We provide tables, with similar viewing options as the differential expression table, for enriched GO terms in each class of a given differential comparison
+
+## Single Cell data 
+Single Cell RNA-seq provides a deeper look into the   
+We provide single cell RNA-seq data from murine retina across 2 studies.
+The data is taken directly from the publishing authors.
+Single cell gene expression data is visualized through a heatmap showing the expression of a single gene across multiple retinal cell types and different developmental time points (E11-P14), and a table of expression values is generated containing the expression data used to draw the heatmap.
+(fig?) We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors.
+(fig?) The plots show all cell types present a given developmental stage, and highlights cells expressing a gene at a user determined level.
+
+## Data accessibility 
+Individual data files for gene expression and sample metadata can be downloaded from the ‘Data’ tab.
+All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/Eyeintegration_app’).
+Code used to generate data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+
+Fig 1A – bulk tissue tsne 
+Fig 1b new samples 
+Table 1a counts at 2019/2017 
+Table 1b #tx/genes for each gene class at different times 2019/2017 
+Fig2 screen shots of expression 
+Fig3 screen shots of differential comparisons 
+(sup)Fig 4 screen shot of single cell data. 
 # Conclusions

### 2019-01-17 manuscript.Rmd
date changed to code in yaml

--- 2019-01-17 manuscript.Rmd
@@
   - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
-date: "1/17/2019"
+date: '`r format(Sys.Date(), "%B %d, %Y")`'
 output:

### 2019-01-17 manuscript.Rmd
updates to intro and bib

--- 2019-01-17 manuscript.Rmd
@@
 ---
-title: 'Eye in a Disk: eyeIntegration version 1.00'
+title: 'Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00'
 author:
@@
 
-  RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome. Dozens of groups have used this technology in human samples to probe the normal 
+  From the anterior to the posterior, the human eye is composed of the cornea, lens, retina, RPE, and choroid.
+The development and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, or transcriptome.
+Today, RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome.
+Analysis of gene patterns across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
+To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids.
+These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].
+  While there is a wealth of pan-eye tissue based RNA-seq data, there has not been a way for researchers to efficiently compare gene and transcript expression information across these tissues.
+For most human tissues the Gene Tissue Expression has generated RNA-sesq data across
+
   By examining gene expression in this concerted manner researchers glean new insight into the complex regulatory mechanisms that affect gene expression and better understand the effect disease has on these mechanisms.

### 2019-01-17 manuscript.Rmd
add iovs csl

@@
 bibliography: eyeIntegration_v1_app_manuscript.bib
+csl: investigative-ophthalmology-and-visual-science.csl
 ---
@@
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].
-  While there is a wealth of pan-eye tissue based RNA-seq data, there has not been a way for researchers to efficiently compare gene and transcript expression information across these tissues.
-For most human tissues the Gene Tissue Expression has generated RNA-sesq data across
+
+The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across 53 post-mortem human tissues and 714 unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
+Unfortunately no eye tissues are included.
 

### 2019-01-18 manuscript.Rmd

@@
 
+```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
+#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
+knitr::opts_chunk$set(out.width = '100%', dpi=300)
+
+# I usually load my libraries up front to keep things organized
+#library(bookdown)
+#library(knitr)
+#library(kableExtra)
+library(pool)
+library(RSQLite)
+library(tidyverse)
+library(citr)
+
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
+core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
+core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+
+```
+
 # Abstract
@@
 
-From the anterior to the posterior, the human eye is composed of the cornea, lens, retina, RPE, and choroid.
+From anterior to posterior, the human eye is composed of the cornea, lens, retina, and RPE (choroid).
 The development and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, or transcriptome.
@@
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].
-
-The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across 53 post-mortem human tissues and 714 unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
-Unfortunately no eye tissues are included.
-
-By examining gene expression in this concerted manner researchers glean new insight into the complex regulatory mechanisms that affect gene expression and better understand the effect disease has on these mechanisms.
-For example, researchers used RNA-seq to establish the effects of the gene NRL on the transcriptomic regulatory network of developing photoreceptors.
-(Swaroop paper).
-Some studies have generated large RNA-seq data sets to accurately determine gene expression specific to individual tissues in the body.
-However, such studies require enormous amounts of funding, due to the large cost of RNA-seq and large number of samples required for these studies.
-Some major institutions like the NIH have performed such studies and made the data publicly available.
-For example, the GTEX project generated a large amount of RNA-seq data in order to determine gene expression ins 55 Human tissues and published the results in an online web-app [@gtexconsortiumGenotypeTissueExpressionGTEx2013].
-This data serves as a valuable tool for researchers across the world, but lacks relevance to eye research as GTEX failed to cover the eye.
-  Another solution to the problem of creating large, high quality RNA-seq data sets is manual curation of data generated by multiple studies.
-There is a vast amount of publicly available RNA-seq data in the Sequence Read Archive (SRA), where investigators across the world deposit their raw sequencing data.
-As of January 17th 2019, the SRA has 493,821 human RNA-seq samples.
-Creating a curated dataset from multiple data sets presents a difficult challenge.
-Samples must be carefully selected and labelled with the proper metadata, evaluated for quality, and normalized to account for technical variation between studies.
-While many fields of research have made such databases (Other curated database), a robust database does not exist in vision science research.
-Previous attempts to make such a database relied on outdated technology like microarrays (iSyte), or had too narrow of focus (TIGEM).
-To remedy this problem, we have created Eyeintegration, the largest curated RNA-seq database for ocular research.
-We provide accurate information about gene expression in multiple human eye tissues using published studies from the SRA.
-Additionally, we incorporate murine single cell RNA-seq data to provide researchers access to information about retinal development.
-The database is built using a highly reproducible and robust pipeline that includes a rigorous quality control procedure and automated publishing of the website.
-The database contains both gene and transcript expression data for the retina, retinal pigmented epithelium (RPE), cornea, lens and 44 other human organs.
-We performed multiple differential expression analysis between tissues types and provide a wide range of data visualizations that can be accessed through an easy-to-use web-app.
-The database can be found at Eyeintegration.nei.nih.gov
+The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues and hundreds unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
+Unfortunately, no eye tissues are included.
+Two groups have quantified huge portions of the RNA-seq data, including some human eye tissues, from the Sequence Read Archive (SRA): recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
+However no curation of the sample level metadata was done, so it is very difficult to parse out which eye tissues are exactly present and, more importnatly, to determine whether any samples were chemically or genetically perturbed.
+More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016].
+iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.
+The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provided the largest set of hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
+This dataset was served in a responsive web app allowing for quick comparisons across many eye tissues and dozens of other body tissues.
+In order to better serve the eye research community, we have re-written the bioinformatic pipeline that creates the Eye in a Disk (EiaD) datasets for eyeIntegration to nearly fully automate the EiaD creation, ensure full reproducbility of the results, allow for external data comparison, provide consistent sample quality control, and improve efficiency for future sample updates.
+Additionally, we are prototyping new tools to display murine single cell RNA-seq data to provide researchers access to information about cell type specific gene expression in retinal development.
+The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples.
+The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands.
 
@@
 ## We find X potential eye studies
-  We searched the Sequence Read Archive^(SRA) for eye related tissues using the query ‘cornea|retina|RPE|ocular|eye’.
-We hand selected relevant studies and selected healthy, control/unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids and cell lines.
-In order to compare gene expression in the eye against expression in other body tissues, we obtained samples for 44 different body tissues from the GTEx project, also housed in the SRA.
-Using SRA metadata from each study we extracted sample and run accessions, library type , tissue of origin, and subtissue of origin.
+  We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|ocular|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
+We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids, and immortalized cell lines.
+In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from 44 different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
+Using SRA metadata from each study we extracted sample and run accessions, library type, tissue of origin, and sub-tissue of origin.
 Any of the preceding information missing from the SRA metadata was added by hand.
 Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.
-## Data Gathering 
+
+## Raw Data Download 
   We downloaded the relevant SRA files for each sample directly from the NCBI ftp server using the file transfer software Aspera.
-SRA files were converted to FASTQ format using the tool fastq-dump from the SRAtoolkit software package^.
-Samples only available in the BAM format were converted to FASTQ format using SAMTOOLS^.
-Samples were quantified using the alignment free quantification software Salmon, using transcriptomic index built from  gencode v28 protein coding transcript sequences^ using the transcriptomic aligner Salmon^.
-Using the resulting expression data, we identified lowly or unused transcripts within annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene.^(sonneson et al).
+SRA files were converted to FASTQ format using the tool fastq-dump from the SRAtoolkit software package.
+Samples only available in the BAM format were converted to FASTQ format using SAMTools [@liSequenceAlignmentMap2009].
+Samples were quantified using the alignment free quantification software Salmon, using transcriptomic index built from gencode v28 protein coding transcript sequences using the transcriptomic aligner Salmon [@harrowGENCODEReferenceHuman2012; @patroSalmonProvidesFast2017a].
+Using the resulting expression quantification, we identified lowly or unused transcripts within annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene as per Sonneson et al [@sonesonIsoformPrefilteringImproves2016].
 Samples were requantified against a transcriptomic index built on the filtered transcript sequences.
-All expression values in length-scaled TPM counts were combined into a single matrix and aggregated to the gene level using tximport^.
+All expression values in length-scaled TPM counts were combined into a single matrix and aggregated to the gene level using tximport [@sonesonDifferentialAnalysesRNAseq2016a].
 
 ## Quality control 
-We began the quality control procedure by removing samples with a Salmon mapping rate less than 50%.
+We began the quality control procedure by removing samples with a Salmon calculated mapping rate less than 50%.
 We further filtered samples by calculating the median gene expression for each sample and only kept sample that had a median count >2 across all genes.
-We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count>200 across all samples.
-To reduce the noise experimental variability between each study, we normalized samples by library using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth^(Hicks et al).
-We generated sub-tissue clusters by first reduced the dimensionality of the data to two dimensions using the t-Stochastic Neighbor Embedding(t-SNE) algorithm, and then calculated the center of the resulting sub-tissue clusters by finding the average coordinate of all samples  of a given sub-tissue type.
-We used these sub-tissues clusters to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type.
-We calculated the distance of each sample from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance.
+We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples.
+To reduce the noise experimental variability between each study, we normalized samples by library using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
+We generated sample level two-dimensional groupings by with the t-Stochastic Neighbor Embedding (t-SNE) algorithm, and then calculated the center of the resulting sub-tissue clusters by finding the average coordinate of all samples of a given sub-tissue type [@maatenAcceleratingTSNEUsing2014].
+We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type.
+We calculated the euclidean distance of each sample from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance.
 
@@
 We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test.
-We extracted significant genes (p <.05) for each eye sub-tissue vs body comparisons and used these to calculate GO enrichment for each eye sub-tissue.
-The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler  R-package (q-value < .01).
+We extracted significant genes (FDR p <.05) for each eye sub-tissue vs body comparisons and used these to calculate GO enrichment for each eye sub-tissue.
+The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R  package (q-value < .01).
 
@@
 
+The 2019 Eye in a Disk (EiaD) dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison.
+The resource was originally to
+
 ## Data collection and Quality control 
-Our initial query to the SRA found 44* potentially relevant studies.
-We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total of <x> studies and 410 samples, with 196 Retina, 143 RPE, 62 Cornea, and 9 lens samples.
+Our initial query to the SRA found 44* potentially relevant studies deposited since our last search XXXXXX.
+We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total of `r  studies and 410 samples, with 196 Retina, 143 RPE, 62 Cornea, and 9 lens samples.
 We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake^, a workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the web-app.

### 2019-01-18 src/tsne_and_pca_calcs.R
tsne and pca code

@@ -1,0 +1,98 @@
+library(tidyverse)
+library(pool)
+library(RSQLite)
+library(dbscan)
+library(ggrepel)
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/Desktop/eyeIntegration_human_expression_2019_01.sqlite')
+core_tight <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble() %>% select(-run_accession) %>% unique()
+anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
+tpm_gene <- gene_pool_2019 %>% tbl('lsTPM_gene') %>% as_tibble() %>%  spread(sample_accession, value)
+set.seed(23452345)
+eye_samples <- core_tight %>%
+  filter(sample_accession %in% colnames(tpm_gene)) %>%
+  filter(!Tissue %in% 'ENCODE Cell Line') %>%
+  filter(study_accession!='SRP012682') %>%
+  filter(!sample_accession %in% c('SRS523795','SRS360124','SRS360123')) %>%
+  .[['sample_accession']]
+
+kept_samples <- core_tight %>%
+  filter(Kept == 'Kept') %>% 
+  pull(sample_accession)
+
+
+# only keep eye samples and protein coding genes
+eye_TPM <- tpm_gene[tpm_gene %>% pull(ID) %in% (anno %>% filter(gene_type == 'protein_coding') %>% pull(ID) %>% unique()),unique(eye_samples)] 
+# all that pass QC
+TPM <- tpm_gene[tpm_gene %>% pull(ID) %in% (anno %>% filter(gene_type == 'protein_coding') %>% pull(ID) %>% unique()), kept_samples] 
+
+plot_maker <- function(df, algorithm = 'pca', dbscan_eps = 2.5){
+  # keep only 3000 most variable genes
+  variance <- apply(df, 1, var, na.rm=TRUE) 
+  df$variance <- variance
+  high_var_TPM <- df %>% arrange(variance) %>% tail(3000) %>% select(-variance)
+  
+  variance <- apply(df, 1, var, na.rm=TRUE) 
+  df$variance <- variance
+  high_var_TPM <- df %>% arrange(variance) %>% tail(3000) %>% select(-variance)
+  
+  if (algorithm == 'pca'){
[…]

### 2019-01-18 supplement.Rmd
start of supplementay materials

---
title: 'Supplementary Materials'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::word_document2:
    reference_docx: word-styles-reference-01.docx
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
---

```{r}
load('data/dimRed_plot_data.Rdata')
library(tidyverse)
library(ggrepel)

# eye PCA
eye_pca_plot <- ggplot(dimRed_plot_prep_eye_pca %>% 
         mutate(Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)),
       aes(x=`Dimension 1`,y=`Dimension 2`)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  ggtitle(paste0('Eye Tissue PCA')) +
  geom_point(size=16, alpha=0.1, aes(colour=Tissue)) +
  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
  xlab('Dimension 1') + ylab('Dimension 2')
# eye tSNE
eye_tsne_plot <- ggplot(dimRed_plot_prep_eye_tsne %>% 
                         mutate(Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)),
                       aes(x=`Dimension 1`,y=`Dimension 2`)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  ggtitle(paste0('Eye Tissue t-SNE')) +
  geom_point(size=16, alpha=0.1, aes(colour=Tissue)) +
  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal()
plot_grid(eye_pca_plot, eye_tsne_plot)

# pca
pca_plot <- ggplot(dimRed_plot_prep_pca %>% 
                     mutate(Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)),
                   aes(x=`Dimension 1`,y=`Dimension 2`)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  ggtitle(paste0('All Tissue PCA')) +
  geom_point(size=16, alpha=0.1, aes(colour=Tissue)) +
  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
  xlab('Dimension 1') + ylab('Dimension 2')

# tSNE
tsne_plot <- ggplot(dimRed_plot_prep_tsne %>% 
                         mutate(Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)),
                       aes(x=`Dimension 1`,y=`Dimension 2`)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  ggtitle(paste0('All Tissue PCA')) +
  geom_point(size=16, alpha=0.1, aes(colour=Tissue)) +
  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
  xlab('Dimension 1') + ylab('Dimension 2')

```

### 2019-01-18 manuscript.Rmd
todays edits and figures added

@@
 output:
-  word_document:
+  bookdown::word_document2:
     reference_docx: word-styles-reference-01.docx
@@
 library(citr)
+library(cowplot)
+library(ggrepel)
+library(colorspace)
+library(png)
 
@@
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+load('data/human_tx_studies.Rdata')
+tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
+anno <- gene_pool_2019 %>% tbl('gene_IDs')
 
@@
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].
+
 The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues and hundreds unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
@@
 iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.
+
 The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provided the largest set of hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
@@
 In order to better serve the eye research community, we have re-written the bioinformatic pipeline that creates the Eye in a Disk (EiaD) datasets for eyeIntegration to nearly fully automate the EiaD creation, ensure full reproducbility of the results, allow for external data comparison, provide consistent sample quality control, and improve efficiency for future sample updates.
-Additionally, we are prototyping new tools to display murine single cell RNA-seq data to provide researchers access to information about cell type specific gene expression in retinal development.
+Additionally, we are prototyping new tools to display murine single cell RNA-seq (scRNA-seq) data to provide researchers access to information about cell type specific gene expression in retinal development.
 The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples.
@@
 # Methods
-## We find X potential eye studies
+## Identifying potential eye samples with exhaustive search of SRA
 We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|ocular|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
 We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids, and immortalized cell lines.
-In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from 44 different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
+In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
 Using SRA metadata from each study we extracted sample and run accessions, library type, tissue of origin, and sub-tissue of origin.
-Any of the preceding information missing from the SRA metadata was added by hand.
+Any of the preceding information missing from the SRA metadata was added by hand, when available.
 Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.
 
-## Raw Data Download 
+## Raw data download and quantification
 We downloaded the relevant SRA files for each sample directly from the NCBI ftp server using the file transfer software Aspera.
@@
 Samples only available in the BAM format were converted to FASTQ format using SAMTools [@liSequenceAlignmentMap2009].
-Samples were quantified using the alignment free quantification software Salmon, using transcriptomic index built from gencode v28 protein coding transcript sequences using the transcriptomic aligner Salmon [@harrowGENCODEReferenceHuman2012; @patroSalmonProvidesFast2017a].
+Sample transcriptomes were quantified using the alignment free quantification software Salmon, using transcriptomic index built from gencode v28 protein coding transcript sequences using the transcriptomic aligner Salmon [@harrowGENCODEReferenceHuman2012; @patroSalmonProvidesFast2017a].
 Using the resulting expression quantification, we identified lowly or unused transcripts within annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene as per Sonneson et al [@sonesonIsoformPrefilteringImproves2016].
-Samples were requantified against a transcriptomic index built on the filtered transcript sequences.
-All expression values in length-scaled TPM counts were combined into a single matrix and aggregated to the gene level using tximport [@sonesonDifferentialAnalysesRNAseq2016a].
+Samples were re-quantified against a transcriptomic index built on the filtered transcript sequences.
+The Salmon count values were were quantified as (transcript) length scaled Transcripts Per Million (TPM) to the transcript and gene level using tximport [@sonesonDifferentialAnalysesRNAseq2016a].
 
@@
 To reduce the noise experimental variability between each study, we normalized samples by library using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
+
 We generated sample level two-dimensional groupings by with the t-Stochastic Neighbor Embedding (t-SNE) algorithm, and then calculated the center of the resulting sub-tissue clusters by finding the average coordinate of all samples of a given sub-tissue type [@maatenAcceleratingTSNEUsing2014].
@@
 
-## Analysis and generation of web app 
+## Differential Gene Expression Analysis and GO term enrichment
 We used the cleaned data to determine differential gene expression between different sub-tissue types.
@@
 We extracted significant genes (FDR p <.05) for each eye sub-tissue vs body comparisons and used these to calculate GO enrichment for each eye sub-tissue.
-The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R  package (q-value < .01).
+The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R  package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
 
-VS: 
-I’ll leave the webapp and single cell stuff for you since I didn’t really work on that part 
+## eyeIntegration web app and R package
+The data generated in the above steps is consolidated into a SQLite database, with the original EiaD data from 2017 and the new 2019 EiaD dataset each getting a separated database file.
+The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov).
+The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app'), eyeIntegrationApp::get_eyeIntegration_datasets(), eyeIntegrationApp::run_eyeIntegration()."
+
+## Snakemake reproducible pipeline
+While the sample search and metadata parsing in a semi-curated process, the processing from the raw data to the creation of the SQLite EiaD database underlying eyeIntegration is wrapped in a Snakemake pipeline, which ensures full reproducibility of the results [@kosterSnakemakeScalableBioinformatics2012].
+We make the code for the pipeline available at https://github.com/davemcg/eyeIntegration_data_build.
+
+## scRNA-seq processing
+The eyeIntegration site currently hosts two large scRNA-seq datasets from Macoscko et al.
+and a unpublished set kindly made available to the public from Clark et al [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
+We use the processed gene count data from each, as well as the cluster assignments that specify what cell type each individual cell is.
+The count data is mean averaged to the cell type, age, and gene level.
+We also display t-SNE and UMAP-based two-dimensional visualizations of the Macosko and Clark data, respectively, in the web app.
+For more details see Methods of [@Bryan_2018].
+
 # Results
 
-The 2019 Eye in a Disk (EiaD) dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison.
-The resource was originally to
+## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes
 
-## Data collection and Quality control 
-Our initial query to the SRA found 44* potentially relevant studies deposited since our last search XXXXXX.
-We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total of `r  studies and 410 samples, with 196 Retina, 143 RPE, 62 Cornea, and 9 lens samples.
-We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake^, a workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the web-app.
-In order to create a robust and clean data set, we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same tissue type.
-We used t-SNE to cluster samples passing quality controls by tissue type to observe variability within samples of the same tissue and determine the strength of our quality control procedure.
-(figure 1A) We found that samples of the same tissue type and origin cluster well together.
-For example, in retina derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal tissue cluster together.
-Our ability to uniformly cluster data by known biological function independent of study origin demonstrates our quality control method’s ability to account for technical variation between studies.
-We provide plots of t-SNE clusters at variable perplexities within the web-app under the ‘2D-clustering’ tab.
-After quality control we found 310 samples to be of sufficient quality, with 180 Retina, 117 RPE, 42 Cornea, and 6 Lens samples included in downstream analysis.
-The 2019 version of the web-app represents a significant change from our previously published Eyeintegration 2017 data set.
-(Table 1a/Figure 1b) We have added 80 more Retina samples, 61 more RPE samples, 22 more cornea samples, and added Lens as a new tissue type, across <x> more studies.
-We have increased the number of sub-tissue types, adding Retinal fetal tissue, retinal organoids and cornea stem cell lines.
-(Fig.
-1A).
-There has been significant change to the types of gene classes.
-We have added miRNA, lincRNA, pseudogenes, and snRNA as gene classes.
-The web app contains 151222 protein coding, 13188 lincRNA, 10202 pseudogenes, 1900 snRNA and 1881 miRNA transcripts, for a net addition of X more genes and Y more transcripts.
-(Table 1b?).
+We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data [@kosterSnakemakeScalableBioinformatics2012].
+In order to create a high quality final dataset, we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+To identify possible tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (Figure \@ref(fig:tSNE)).
 
-## eyeIntegration web application provides interactive visual portal to all data 
-All data generated by our analysis is presented through the Eyeintegration web-app.
-The web-app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
-The web-app is divided into 3 general categories: gene/transcript expression, differential comparisons, and single cell data.
+After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together.
+For example, in retina derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina tissue cluster together.
+Our ability to uniformly cluster data by known biological function independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.
 
-## Expression 
-The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues.
-(Fig 2) The user first selects which version of the web-app to use data from, as well as genes and tissues to examine.
+```{r tSNE, fig.height = 15, fig.width = 20, echo=F, message = F, fig.cap="Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing"}
+dbscan_cluster <- dbscan::dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
+tsne_50$Cluster <- dbscan_cluster$cluster
+col1 = 'X1'
+col2 = 'X2'
+# create label points for each cluster
+cluster_centers <- tsne_50 %>%
+  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
+  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))
+
+# samples closest to center in each cluster
+center_samples <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
+  left_join(.,cluster_centers, by=c('Cluster')) %>%
+  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
+  group_by(Cluster) %>%
+  dplyr::slice(which.min(Distance)) %>%
+  dplyr::slice(1) %>% 
+  .[['sample_accession']]
+
+# cluster stats
+cluster_stats <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  group_by(Cluster) %>%
+  summarise(Cluster_Tissues = paste(unique(Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep=''))
+
+# set up for ggplot
+tsne_50_prep <- tsne_50 %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  select(-Cluster_Tissues) %>% 
+  left_join(., cluster_stats, by=c('Cluster')) %>%
+  mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
+  mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
+
+tsne_50_prep %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+                                     TRUE ~ Tissue),
+                   Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
+                   Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+ggplot(aes(x=X1,y=X2)) +
+  scale_shape_manual(values=c(0:20,35:50)) +
+  geom_point(size=16, alpha=0.1, aes(colour=Group)) +
+  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
+  xlab('t-SNE 1') + ylab('t-SNE 2') +
+  geom_label_repel(size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
+  theme_minimal() +
+  theme(text = element_text(family = 'Helvetica', size = 30)) +
+  scale_color_discrete_sequential(palette = 'viridis') + 
+  guides(colour = guide_legend(override.aes = list(alpha = 1)))
+```
+
+## `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
+Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19th, 2017.
+We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison.
+The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD (Figure \@ref(fig:tissueCountsBarPlot)).
+Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
+Another substantial addition to the 2019 EiaD are non-protein coding genes.
+We now quantify expression across `r gene_pool_2019 %>% tbl('gene_IDs') %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r gene_pool_2019 %>% tbl('tx_IDs') %>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
+
+```{r tissueCountsBarPlot, fig.height = 4, fig.cap="Comparison of counts of eye tissue samples from 2017 to 2019 EiaD", echo=F}
+core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
+
+core_both %>% filter(grepl('Retina|Cornea|Lens|RPE|ESC', Tissue)) %>% unique() %>% mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>% group_by(Tissue, Sub_Tissue, Version) %>% summarise(Count=n()) %>%
+  mutate(Dataset = Version) %>% 
+  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
+  theme(text = element_text(family = 'Helvetica', size = 10))
+```
+
+## eyeIntegration web app provides interactive visual portal to all data 
+All data generated by our analyses is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
+The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
+The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.
+
+## Custom gene and tissue expression boxplots
+The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (Figure \@ref(fig:appScreenshots)A).
+The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
+A boxplot is then generated afer hitting the "Re(Draw) Plot" with overlaid individual data points.
 A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue.
 The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.
-The web app provides multiple options for visualizing expression data.
-Box plots can be drawn to compare gene expression across tissues.
-Users may select any number of genes and tissues to compare at a given time, and can arrange visualizations in multiple ways.
-Box plots provide the user a measure of confidence in the variability of reported gene expression levels.
-The metadata behind any given samples can be seen by hovering over a point in the box plot, including study of origin, sex of sample, and SRA accession.
-Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues.
-The visualizations can also be generated at the transcript level by selecting pan-transcript expression under the expression tab.
-Finally, a session can be saved or shared by building a custom link for the session with the build URL shortcut button.
 
-## Differential expression and Gene Ontology data. 
-Using our expression data, we performed multiple differential comparisons at the sub-tissue level.
-We performed differential tests of eye tissues against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTeX data set, allowing users to quickly identify eye specific genes.
-We also performed differential tests between individual eye tissues.
-The user can view the results under the ‘differential’ selection under the ‘expression’ tab.
-(fig 3) As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons.
+```{r appScreenshots, echo = F, fig.cap = 'Various screenshots from eyeIntegration web app.
+A.
+Pan-tissue gene expression box plots with accompanying data tables.
+B.
+Alternative heatmap visualization.
+C.
+CRX gene expression heatmap and table information from Macosko et al.
+P14 scRNA-seq.
+D.
+t-SNE visualization of individual cell gene expression profiles from Macosko et al.
+E.
+Data table export view.
+F.
+Differential gene expression across different tissues and with GO term enrichment.
+', fig.width=12}
+knitr::include_graphics('figures_and_tables/Figure 2.png')
+```
+
+Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues (Figure \@ref(fig:appScreenshots)B).
+Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.
+
+## Differential expression and gene ontology enrichment tests 
+Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes.
+The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (Figure \@ref(fig:appScreenshots)F).
+As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons.
 The user additionally has the option to select different gene classes to examine, protein coding, lincRNA etc.
+
 The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values.
@@
 Enriched GO terms are presented first as a word cloud, for quick comparison of GO enrichment.
-We provide tables, with similar viewing options as the differential expression table, for enriched GO terms in each class of a given differential comparison
+We provide tables, with similar viewing options as the differential expression table, for enriched GO terms in each class of a given differential comparison.
 
-## Single Cell data 
-Single Cell RNA-seq provides a deeper look into the   
-We provide single cell RNA-seq data from murine retina across 2 studies.
-The data is taken directly from the publishing authors.
-Single cell gene expression data is visualized through a heatmap showing the expression of a single gene across multiple retinal cell types and different developmental time points (E11-P14), and a table of expression values is generated containing the expression data used to draw the heatmap.
-(fig?) We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors.
-(fig?) The plots show all cell types present a given developmental stage, and highlights cells expressing a gene at a user determined level.
+## Gene expression in individual retina cells types with scRNA-seq 
+We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a], allowing for examination of gene expression in individual cell types in the retina.
+The count data is taken directly from the publishing authors (see Methods).
+Single cell gene expression data is visualized through a heatmap showing the expression of a single gene across multiple retinal cell types and different developmental time points (E11-P14), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C).
+We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D).
+The plots show all cell types present a given developmental stage, and highlights cells expressing a gene at a user determined level.
 
 ## Data accessibility 
-Individual data files for gene expression and sample metadata can be downloaded from the ‘Data’ tab.
-All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/Eyeintegration_app’).
+Individual data files for gene expression and sample metadata can be downloaded from the ‘Data’ tab on the web app.
+All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’).
 Code used to generate data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
 
-Fig 1A – bulk tissue tsne 
-Fig 1b new samples 
-Table 1a counts at 2019/2017 
-Table 1b #tx/genes for each gene class at different times 2019/2017 
-Fig2 screen shots of expression 
-Fig3 screen shots of differential comparisons 
-(sup)Fig 4 screen shot of single cell data. 
 # Conclusions

### 2019-01-18 supplement.Rmd
pca plot update

@@
 pca_plot <- ggplot(dimRed_plot_prep_pca %>% 
-                     mutate(Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)),
+                     mutate(Origin = case_when(study_accession == 'SRP012682' ~ 'Adult Tissue',
+                                               TRUE ~ as.character(Origin))) %>% 
+                     mutate(Tissue = case_when(Tissue == 'Brain' ~ Tissue,
+                                               study_accession == 'SRP012682' ~ 'GTEx',
+                                               TRUE ~ Tissue)),
                    aes(x=`Dimension 1`,y=`Dimension 2`)) +
@@
   geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
-  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
-  xlab('Dimension 1') + ylab('Dimension 2')
+  #geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
+  xlab('Dimension 1') + ylab('Dimension 2') +
+  theme_minimal()
 

### 2019-01-18 supplement.Rmd
point size change

--- 2019-01-18 supplement.Rmd
@@
   ggtitle(paste0('All Tissue PCA')) +
-  geom_point(size=16, alpha=0.1, aes(colour=Tissue)) +
-  geom_point(size=6, alpha=0.6, position='jitter', aes(shape=Origin)) +
-  geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
-  xlab('Dimension 1') + ylab('Dimension 2')
+  geom_point(size=6, alpha=0.1, aes(colour=Tissue)) +
+  geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+  #geom_label_repel(aes(label=Label), alpha=0.8, size=4, box.padding = unit(0.3, "lines")) + theme_minimal() +
+  xlab('Dimension 1') + ylab('Dimension 2') +
+  theme_minimal()
 

### 2019-01-20 manuscript.Rmd
captioner captions

@@
 output:
-  bookdown::word_document2:
+  word_document:
     reference_docx: word-styles-reference-01.docx
@@
 
-# I usually load my libraries up front to keep things organized
-#library(bookdown)
-#library(knitr)
-#library(kableExtra)
 library(pool)
@@
 library(png)
+library(flextable)
+library(captioner)
 
+# setup caption-ing
+fig_cap <- captioner("Figure")
+supFig_cap <- captioner("Supplemental Figure")
+tab_cap <- captioner("Table")
+supTab_cap <- captioner("Supplemental Table")
+
+# attach data pools
 gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
@@
 To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids.
-These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].
+These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
 
@@
 The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov).
-The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app'), eyeIntegrationApp::get_eyeIntegration_datasets(), eyeIntegrationApp::run_eyeIntegration()."
+The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
 
@@
 and a unpublished set kindly made available to the public from Clark et al [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
-We use the processed gene count data from each, as well as the cluster assignments that specify what cell type each individual cell is.
-The count data is mean averaged to the cell type, age, and gene level.
+We use the processed gene count data from each group, as well as the cluster assignments that specify what cell type each individual cell is.
+The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration.
 We also display t-SNE and UMAP-based two-dimensional visualizations of the Macosko and Clark data, respectively, in the web app.
-For more details see Methods of [@Bryan_2018].
+For detail so the t-SNE processing we did on the Macosko dataset, see the methods of [@Bryan_2018].
 
@@
 ## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes
-
+```{bash, echo = F, eval = F}
+# ran in biowulf2.nih.gov:/data/mcgaugheyd/projects/nei/mcgaughey/auto_eyeIntegration
+# 1314 from running: for i in RE_quant_files/*/lib_format_counts.json; do echo $i; done | wc -l 
+# 46030412878 from running: for i in RE_quant_files/*/lib_format_counts.json; do grep num_compatible_fragments $i; done |  cut -f2 -d":" | cut -f1 -d"," | awk '{sum += $1} END {print sum}'
+```
 We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data [@kosterSnakemakeScalableBioinformatics2012].
-In order to create a high quality final dataset, we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
-To identify possible tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (Figure \@ref(fig:tSNE)).
+In order to create a high quality final dataset across the 1,314 initial samples and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+To identify tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
 
@@
 
-```{r tSNE, fig.height = 15, fig.width = 20, echo=F, message = F, fig.cap="Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing"}
+```{r tsne_fig, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
+tsne_cap <- fig_cap(name='tSNE', caption = 'Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing')
+
 dbscan_cluster <- dbscan::dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
@@
 
-## `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sample along with `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
 Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19th, 2017.
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
-The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison.
-The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD (Figure \@ref(fig:tissueCountsBarPlot)).
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r paste0('Tables ', tab_cap(name = 'tableSampleCounts', display = 'num'), ', ', tab_cap(name = 'tableSampleCountsGTEx', display = 'num'))`.
+The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
 Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
@@
 
-```{r tissueCountsBarPlot, fig.height = 4, fig.cap="Comparison of counts of eye tissue samples from 2017 to 2019 EiaD", echo=F}
+```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
+tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type  from 2017 to 2019 EiaD')
+
 core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
@@
 
+```{r tableSampleCounts, echo = F, results='asis', warning = F, fig.cap = tableSampleCounts_cap}
+tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD and eyeIntegration contain a huge set of eye tissues')
+table <- left_join(core_tight_2019  %>% 
+                     select(-run_accession) %>% 
+                     unique() %>% 
+                     filter(study_accession != 'SRP012682',Tissue %in% c('RPE','Lens','Cornea', 'Retina')) %>% 
+                     group_by(Tissue) %>% 
+                     summarise(`Pre QC Count` = n()),
+
+                   core_tight_2019  %>% 
+                     select(-run_accession) %>% 
+                     unique() %>% 
+                     filter(study_accession != 'SRP012682', Tissue %in% c('RPE','Lens','Cornea', 'Retina'), Kept == 'Kept') %>% 
+                     rowwise() %>% 
+                     mutate(Sub = str_split(Sub_Tissue, ' - ')[[1]][2]) %>% 
+                     mutate(Sub = case_when(is.na(Sub) ~ Sub_Tissue, TRUE ~ Sub)) %>% 
+                     group_by(Tissue, Sub) %>% summarise(Count = n()) %>% 
+                     mutate(Sub = paste0(Sub, ' (', Count, ')')) %>% 
+                     ungroup() %>% 
+                     group_by(Tissue) %>% 
+                     summarise(`Count` = sum(Count), `Sub-Tissue Types (Count)` = paste(sort(unique(trimws(Sub))), collapse=', ')), by = 'Tissue') %>% 
+  as_tibble()
+
+gtex_tissues <- core_tight_2019 %>% 
+  select(-run_accession) %>% 
+  unique() %>%
+  filter(study_accession == 'SRP012682', Kept == 'Kept') %>%
+  group_by(Tissue, study_accession) %>% 
+  summarise(Count = n()) %>% 
+  ungroup() %>% 
+  mutate(Tissue = paste0(Tissue, ' (', Count, ')')) %>% 
+  pull(Tissue)
+
+gtex <- cbind('GTEx', core_tight_2019 %>% 
+                select(-run_accession) %>% 
+                unique() %>%
+                filter(study_accession == 'SRP012682') %>% nrow(),
+              core_tight_2019 %>% 
+                select(-run_accession) %>% 
+                unique() %>%
+                filter(study_accession == 'SRP012682', Kept == 'Kept') %>% nrow(),
+              paste(gtex_tissues, collapse = ', ')) %>% as_tibble()
+
+colnames(gtex) <- c('Group','Pre QC Count', 'Count', 'Tissue Types (Count)')
+cat("<div id=\"tbl:tableSampleCounts\">")
+regulartable(table) %>% width(width = c(0.5,1,0.5, 2)) %>% add_footer(., tableSampleCounts_cap) %>% flextable::align(align='center')
+#Output Caption
+cat(paste0("\n",tableSampleCounts_cap,"\n"))
+# Output </div>
+cat("</div>")
+```
+```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap}
+tableSampleCountsGTEx_cap <- tab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
+cat("<div id=\"tbl:tableSampleCountsGTEx\">")
+regulartable(gtex) %>% width(width = c(0.5,1,0.5,5)) %>% flextable::align(align='center')
+#Output Caption
+cat(paste0("\n",tableSampleCountsGTEx_cap,"\n"))
+# Output </div>
+cat("</div>")
+```
+
 ## eyeIntegration web app provides interactive visual portal to all data 
-All data generated by our analyses is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
+The EiaD 2019 dataset is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
 The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
@@
 ## Gene expression in individual retina cells types with scRNA-seq 
-We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a], allowing for examination of gene expression in individual cell types in the retina.
-The count data is taken directly from the publishing authors (see Methods).
-Single cell gene expression data is visualized through a heatmap showing the expression of a single gene across multiple retinal cell types and different developmental time points (E11-P14), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C).
-We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D).
-The plots show all cell types present a given developmental stage, and highlights cells expressing a gene at a user determined level.
+We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
+This allows researchers to quickly examine gene expression across individual cell types in the retina.
+Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C).
+We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D, see Methods).
+The plots show all cell types present a given developmental stage, and highlights cells expressing a gene above a user given level.
+
+## Retina Organoids, Fetal Tissue, and iPSC-derived retina are closely related
+
+There are, currently, three approaches to collecting human fetal retina: 1.
+post-mortem tissue, 2.
+organoids, and 3.
+iPSC-derived cells.
+As an example of how the EiaD 2019 dataset can be used to answer pan-study questions, we looked at whether at how well these three approaches to generating fetal retina compare at a transcriptomic level.
+Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
 

### 2019-01-20 src/data_pull.R
pull script to extract data and make df to save on knitting time

@@ -1,0 +1,14 @@
+# pull data from pools and save as Rdata
+library(pool)
+library(RSQLite)
+library(tidyverse)
+
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
+core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
+core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+load('data/human_tx_studies.Rdata')
+tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
+gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
+tx_anno <- gene_pool_2019 %>% tbl('tx_IDs') %>% as_tibble()
+save(core_tight_2017, core_tight_2019, tsne_50, gene_anno, tx_anno, file = 'data/data_pull.Rdata')

### 2019-01-20 manuscript.Rmd
move pool data pulls into  separate script src/data_pull.R to improve knit time

@@
 library(colorspace)
-library(png)
 library(flextable)
 library(captioner)
+library(dbscan)
 
@@
 # attach data pools
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
-core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
-core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+# source('src/data_pull.R')
+load('data/data_pull.Rdata')
+# vinny sra search
 load('data/human_tx_studies.Rdata')
-tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
-anno <- gene_pool_2019 %>% tbl('gene_IDs')
-
 ```
@@
 
-dbscan_cluster <- dbscan::dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
+dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
 tsne_50$Cluster <- dbscan_cluster$cluster
@@
 Another substantial addition to the 2019 EiaD are non-protein coding genes.
-We now quantify expression across `r gene_pool_2019 %>% tbl('gene_IDs') %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r gene_pool_2019 %>% tbl('tx_IDs') %>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
+We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
 
@@
 cat("<div id=\"tbl:tableSampleCountsGTEx\">")
-regulartable(gtex) %>% width(width = c(0.5,1,0.5,5)) %>% flextable::align(align='center')
+regulartable(gtex) %>% width(width = c(0.5,1,0.5,4)) %>% flextable::align(align='center')
 #Output Caption

### 2019-01-21 manuscript.Rmd
demo fig 4 of fetal retina tsne

@@
     reference_docx: word-styles-reference-01.docx
+    fig_caption: yes
     pandoc_args:
@@
 library(dbscan)
+library(ggforce)
 
@@
 
-After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together.
+After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')`).
 For example, in retina derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina tissue cluster together.
@@
 ```{r tsne_fig, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
-tsne_cap <- fig_cap(name='tSNE', caption = 'Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing')
+tsne_cap <- fig_cap(name='tSNE', caption = 't-SNE tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing.
+Colors match different tissue sets and shapes of points define the origin of the tissues.')
 
@@
 colnames(gtex) <- c('Group','Pre QC Count', 'Count', 'Tissue Types (Count)')
+# Output <div>
 cat("<div id=\"tbl:tableSampleCounts\">")
 regulartable(table) %>% width(width = c(0.5,1,0.5, 2)) %>% add_footer(., tableSampleCounts_cap) %>% flextable::align(align='center')
-#Output Caption
-cat(paste0("\n",tableSampleCounts_cap,"\n"))
+# Output Caption (* * for italics)
+cat(paste0("\n*",tableSampleCounts_cap,"*\n"))
 # Output </div>
@@
 ```
+
 ```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap}
 tableSampleCountsGTEx_cap <- tab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
+# Output <div>
 cat("<div id=\"tbl:tableSampleCountsGTEx\">")
-regulartable(gtex) %>% width(width = c(0.5,1,0.5,4)) %>% flextable::align(align='center')
-#Output Caption
-cat(paste0("\n",tableSampleCountsGTEx_cap,"\n"))
+regulartable(gtex) %>% width(width = c(0.5,1,0.5,3)) %>% flextable::align(align='center')
+#Output Caption (* * for italics)
+cat(paste0("\n*",tableSampleCountsGTEx_cap,"*\n"))
 # Output </div>
@@
 The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.
-
 ## Custom gene and tissue expression boxplots
-The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (Figure \@ref(fig:appScreenshots)A).
+The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
 The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
@@
 
-```{r appScreenshots, echo = F, fig.cap = 'Various screenshots from eyeIntegration web app.
+```{r appScreenshots, echo = F, fig.cap = appScreenshots_cap, fig.width=12}
+appScreenshots_cap <- fig_cap(name='appScreenshots', caption = 'Various screenshots from eyeIntegration web app.
 A.
@@
 Differential gene expression across different tissues and with GO term enrichment.
-', fig.width=12}
+')
+
 knitr::include_graphics('figures_and_tables/Figure 2.png')
@@
 
-Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues (Figure \@ref(fig:appScreenshots)B).
+Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
 Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.
@@
 Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes.
-The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (Figure \@ref(fig:appScreenshots)F).
+The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
 As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons.
@@
 This allows researchers to quickly examine gene expression across individual cell types in the retina.
-Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C).
-We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D, see Methods).
+Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'C')`).
+We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'D, see Methods')`).
 The plots show all cell types present a given developmental stage, and highlights cells expressing a gene above a user given level.
@@
 iPSC-derived cells.
-As an example of how the EiaD 2019 dataset can be used to answer pan-study questions, we looked at whether at how well these three approaches to generating fetal retina compare at a transcriptomic level.
+As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to generating fetal retina compare at a transcriptomic level.
 Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
+
+To evaluate this question, we look at the t-SNE plot of all tissues and focus in on the four types of retina tissue (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+
+```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F}
+tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes')
+
+tsne_zoom <- tsne_50_prep  %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+                                                        TRUE ~ Tissue),
+                                      Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
+                                      Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+  filter(Group == 'Retina') %>%
+  mutate( zoom = TRUE )
+
+tsne_50_prep %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+                                          TRUE ~ Tissue),
+                        Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
+                        Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+  ggplot(aes(x=X1,y=X2)) +
+  scale_shape_manual(values=c(0:20,35:50)) +
+  geom_point(size=6, alpha=0.1, aes(colour=Group)) +
+  geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+  xlab('t-SNE 1') + ylab('t-SNE 2') +
+  facet_zoom(xy = Group == 'Retina', zoom.size = 1, zoom.data=zoom) +
+  geom_label_repel(data=tsne_zoom,size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
+  scale_color_discrete_sequential(palette = 'viridis') + 
+  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  scale_fill_continuous(guide = guide_legend()) +
+  theme(legend.position="bottom")
+```
 

### 2019-01-21 manuscript.Rmd
improve tsne zoom handling

@@
 
-tsne_zoom <- tsne_50_prep  %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+retina_plus_group <- tsne_50 %>% filter(X1 > 34, X1 < 51, X2 > -25, X2 < 21)
+dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2.0)
+
+retina_plus_group$Cluster <- dbscan_cluster$cluster
+col1 = 'X1'
+col2 = 'X2'
+# create label points for each cluster
+cluster_centers <- retina_plus_group %>%
+  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
+  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))
+
+# samples closest to center in each cluster
+center_samples <- retina_plus_group %>% left_join(.,core_tight_2019)  %>%
+  left_join(.,cluster_centers, by=c('Cluster')) %>%
+  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
+  group_by(Cluster) %>%
+  dplyr::slice(which.min(Distance)) %>%
+  dplyr::slice(1) %>% 
+  .[['sample_accession']]
+
+# cluster stats
+cluster_stats <- retina_plus_group %>% left_join(.,core_tight_2019)  %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  group_by(Cluster) %>%
+  summarise(Cluster_Tissues = paste(unique(Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep=''))
+
+# set up for ggplot
+retina_plus_group_prep <- retina_plus_group %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  select(-Cluster_Tissues) %>% 
+  left_join(., cluster_stats, by=c('Cluster')) %>%
+  mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
+  mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
+
+tsne_zoom <- retina_plus_group_prep  %>% 
+  mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                            TRUE ~ Tissue),
@@
          Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
-  filter(Group == 'Retina') %>%
   mutate( zoom = TRUE )
@@
                         Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+  rowwise() %>% 
+  mutate(Age = str_split(sample_attribute, '_')[[1]][2]) %>% 
   ggplot(aes(x=X1,y=X2)) +
   scale_shape_manual(values=c(0:20,35:50)) +
-  geom_point(size=6, alpha=0.1, aes(colour=Group)) +
-  geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+  geom_point(size=9, alpha=0.1, aes(colour=Group)) +
+  geom_point(size=1, alpha=0.4, position='jitter', aes(shape=Origin)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
-  facet_zoom(xy = Group == 'Retina', zoom.size = 1, zoom.data=zoom) +
-  geom_label_repel(data=tsne_zoom,size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
+  facet_zoom(xy = Group == 'Retina', zoom.size = 1, zoom.data=zoom, horizontal = F) +
+  geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
   scale_color_discrete_sequential(palette = 'viridis') + 
-  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
-  scale_fill_continuous(guide = guide_legend()) +
-  theme(legend.position="bottom")
+  guides(colour = guide_legend(override.aes = list(alpha = 1))) 
 ```

### 2019-01-22 manuscript.Rmd
adjust facet zoom tsne plot

--- 2019-01-21 manuscript.Rmd
@@
 
-```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F}
+```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F, warning = F}
 tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes')
 
-tsne_zoom <- tsne_50_prep  %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+retina_plus_group <- tsne_50 %>% filter(X1 > 34, X1 < 51, X2 > -25, X2 < 21,sample_accession!='SRS1572426')
+dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 1)
+
+retina_plus_group$Cluster <- dbscan_cluster$cluster
+col1 = 'X1'
+col2 = 'X2'
+# create label points for each cluster
+cluster_centers <- retina_plus_group %>%
+  left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept")) %>% group_by(Cluster) %>%
+  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Sub_Tissue),collapse=','))
+
+# samples closest to center in each cluster
+center_samples <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
+  left_join(.,cluster_centers, by=c('Cluster')) %>%
+  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
+  group_by(Cluster) %>%
+  dplyr::slice(which.min(Distance)) %>%
+  dplyr::slice(1) %>% 
+  .[['sample_accession']]
+
+# cluster stats
+cluster_stats <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  group_by(Cluster) %>%
+  summarise(Cluster_Tissues = paste(unique(Sub_Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep=''))
+
+# set up for ggplot
+retina_plus_group_prep <- retina_plus_group %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 40 days', 'Organoid > 40 days'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  select(-Cluster_Tissues) %>% 
+  left_join(., cluster_stats, by=c('Cluster')) %>%
+  mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
+  mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
+
+tsne_zoom <- retina_plus_group_prep  %>% 
+  mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                            TRUE ~ Tissue),
@@
          Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
-  filter(Group == 'Retina') %>%
   mutate( zoom = TRUE )
 
-tsne_50_prep %>% mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+tsne_50_prep %>% 
+  rowwise() %>% 
+  mutate(Age = as.integer(str_split(sample_attribute, '_')[[1]][2])) %>% 
+  mutate(Origin = as.character(Origin)) %>% 
+  mutate(Origin = case_when(Age < 40 ~ 'Organoid < 40 days',
+                            Age > 40 ~ 'Organoid > 40 days',
+                            TRUE ~ Origin)) %>% 
+    mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 40 days', 'Organoid > 40 days'))) %>% 
+  mutate(Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
+                            TRUE ~ Tissue)) %>% 
+  mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                                           TRUE ~ Tissue),
@@
   scale_shape_manual(values=c(0:20,35:50)) +
-  geom_point(size=6, alpha=0.1, aes(colour=Group)) +
-  geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+  geom_point(size=8, alpha=0.2, aes(colour=Group)) +
+  geom_point(size=2, alpha=0.4, position='jitter', aes(shape=Origin)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
-  facet_zoom(xy = Group == 'Retina', zoom.size = 1, zoom.data=zoom) +
-  geom_label_repel(data=tsne_zoom,size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
+  facet_zoom(xy = Group == 'Retina', zoom.size = 2, zoom.data=zoom, horizontal = F) +
+  #geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 50) + 
   scale_color_discrete_sequential(palette = 'viridis') + 
   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
-  scale_fill_continuous(guide = guide_legend()) +
-  theme(legend.position="bottom")
+  theme(axis.text.x = element_text(size = 8),
+        axis.text.y = element_text(size = 8),
+        text = element_text(family = 'Helvetica', size = 8))
 ```

### 2019-01-22 manuscript.Rmd
convex hulls used for grouping in tsne zoom fetal retina

@@
 The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov).
-The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
+The app can be deployed on a local computer with 30GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
 
@@
   ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
-  theme(text = element_text(family = 'Helvetica', size = 10))
+  theme(text = element_text(family = 'Helvetica', size = 10)) +
+  scale_fill_discrete_sequential(palette = 'viridis')  
 ```
@@
 retina_plus_group_prep <- retina_plus_group %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 40 days', 'Organoid > 40 days'))) %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
   mutate( zoom = TRUE )
+StatChull <- ggproto("StatChull", Stat,
+  compute_group = function(data, scales) {
+    data[chull(data$x, data$y), , drop = FALSE]
+  },
+
+  required_aes = c("x", "y")
+)
+
+stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
+                       position = "identity", na.rm = FALSE, show.legend = NA, 
+                       inherit.aes = TRUE, ...) {
+  layer(
+    stat = StatChull, data = data, mapping = mapping, geom = geom, 
+    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
+    params = list(na.rm = na.rm, ...)
+  )
+}
 
@@
   rowwise() %>% 
-  mutate(Age = as.integer(str_split(sample_attribute, '_')[[1]][2])) %>% 
+  mutate(Age = suppressWarnings(as.integer(str_split(sample_attribute, '_')[[1]][2]))) %>% 
   mutate(Origin = as.character(Origin)) %>% 
-  mutate(Origin = case_when(Age < 40 ~ 'Organoid < 40 days',
-                            Age > 40 ~ 'Organoid > 40 days',
+  mutate(Origin = case_when(Age < 30 ~ 'Organoid < 30 days',
+                            Age > 30 ~ 'Organoid > 30 days',
                             TRUE ~ Origin)) %>% 
-    mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 40 days', 'Organoid > 40 days'))) %>% 
+    mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>% 
   mutate(Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
                             TRUE ~ Tissue)) %>% 
+  mutate(ZoomTissue = case_when((Tissue == 'Retina' & Sub_Tissue != 'Retina - Adult Tissue') | Tissue == 'ESC' ~ 'Early Retina',
+                                 TRUE ~ Sub_Tissue)) %>% 
   mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
@@
   scale_shape_manual(values=c(0:20,35:50)) +
-  geom_point(size=8, alpha=0.2, aes(colour=Group)) +
-  geom_point(size=2, alpha=0.4, position='jitter', aes(shape=Origin)) +
+  #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
+  geom_point(size=2, alpha=0.8, position='jitter', aes(shape=Origin, colour = Group)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
-  facet_zoom(xy = Group == 'Retina', zoom.size = 2, zoom.data=zoom, horizontal = F) +
+  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
   #geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 50) + 
+  stat_chull(data=tsne_zoom,aes(fill=Cluster_Tissues), alpha = 0.3) +
   scale_color_discrete_sequential(palette = 'viridis') + 
+  scale_fill_discrete_sequential(palette = 'viridis') + 
+  theme_bw() +
   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
@@
         text = element_text(family = 'Helvetica', size = 8))
+
 ```

### 2019-01-22 manuscript.Rmd
font size adjust

@@
   geom_label_repel(size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
-  theme_minimal() +
+  theme_bw() +
   theme(text = element_text(family = 'Helvetica', size = 30)) +
@@
   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
-  theme(axis.text.x = element_text(size = 8),
-        axis.text.y = element_text(size = 8),
-        text = element_text(family = 'Helvetica', size = 8))
+  theme(axis.text.x = element_text(size = 6),
+        axis.text.y = element_text(size = 6),
+        text = element_text(family = 'Helvetica', size = 6))
 

### 2019-01-22 manuscript.Rmd
2 column legend for tsne zoom fig

@@
   theme_bw() +
-  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
+         shape = guide_legend(ncol =2 )) +
   theme(axis.text.x = element_text(size = 6),

### 2019-01-22 manuscript.Rmd
adjust legend naming

--- 2019-01-22 manuscript.Rmd
@@
 
-To evaluate this question, we look at the t-SNE plot of all tissues and focus in on the four types of retina tissue (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+To evaluate this question, we look at the t-SNE plot of all tissues and focus in on the four types of retina tissues (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+We see that there are three distinct groups (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')` [35, 10], [37, 10], and [47, -5]), which correspond to iPSC and Retina Organoid, the three types of developing retina, and adult retina, respectively.
+Despite the technical differences
 
@@
   #geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 50) + 
-  stat_chull(data=tsne_zoom,aes(fill=Cluster_Tissues), alpha = 0.3) +
+  stat_chull(data=tsne_zoom %>% mutate(`Tissue Cluster` = Cluster_Tissues),aes(fill=`Tissue Cluster`), alpha = 0.3) +
   scale_color_discrete_sequential(palette = 'viridis') + 

### 2019-01-22 manuscript.Rmd
change all figure/table to Linux Libertine O font

--- 2019-01-22 manuscript.Rmd
@@
                         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
-                        Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+                        Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>% 
   ggplot(aes(x=X1,y=X2)) +
@@
   theme_bw() +
-  theme(text = element_text(family = 'Helvetica', size = 30)) +
+  theme(text = element_text(family = 'Linux Libertine O', size = 30)) +
   scale_color_discrete_sequential(palette = 'viridis') + 
@@
   ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
-  theme(text = element_text(family = 'Helvetica', size = 10)) +
+  theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
   scale_fill_discrete_sequential(palette = 'viridis')  
@@
 cat("<div id=\"tbl:tableSampleCounts\">")
-regulartable(table) %>% width(width = c(0.5,1,0.5, 2)) %>% add_footer(., tableSampleCounts_cap) %>% flextable::align(align='center')
+regulartable(table) %>% 
+  width(width = c(0.5,1,0.5, 2)) %>% 
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
 # Output Caption (* * for italics)
@@
 cat("<div id=\"tbl:tableSampleCountsGTEx\">")
-regulartable(gtex) %>% width(width = c(0.5,1,0.5,3)) %>% flextable::align(align='center')
+regulartable(gtex) %>% 
+  width(width = c(0.5,1,0.5,3)) %>% 
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
 #Output Caption (* * for italics)
@@
 
-## Retina Organoids, Fetal Tissue, and iPSC-derived retina are closely related
-
+## EiaD 2019 demonstrates that organoids, iPSC-derived, and fetal tissue retina have closely related transcriptomes
 There are, currently, three approaches to collecting human fetal retina: 1.
@@
 iPSC-derived cells.
-As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to generating fetal retina compare at a transcriptomic level.
+As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to generating developing retina compare at a transcriptomic level.
 Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
 
-To evaluate this question, we look at the t-SNE plot of all tissues and focus in on the four types of retina tissue (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+Here we see three distinct groupings: adult retina (2), developing retina (3), and iPSC/undifferentiated (1).
+We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
+All of the iPSC-based retina samples are older than 50 days.
 
 ```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F, warning = F}
-tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes')
+tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes.
+The zoom inset shows the retina samples.
+The "Tissue Cluster" shading shows the cluster membership of the three major groups.
+The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.')
 
 retina_plus_group <- tsne_50 %>% filter(X1 > 34, X1 < 51, X2 > -25, X2 < 21,sample_accession!='SRS1572426')
-dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 1)
+dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)
 
@@
   group_by(Cluster) %>%
-  summarise(Cluster_Tissues = paste(unique(Sub_Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep=''))
+  summarise(Cluster_Tissues = paste(unique(Sub_Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep='')) %>% 
+  mutate(Cluster_Tissues = paste0(Cluster, ': ', Cluster_Tissues))
 
@@
   left_join(., cluster_stats, by=c('Cluster')) %>%
-  mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
+  mutate(Label = Cluster) %>%
   mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
@@
          Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
-         Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>%
+         Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>%
   mutate( zoom = TRUE )
+
+# https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html
 StatChull <- ggproto("StatChull", Stat,
@@
 )
-
 stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
@@
                             TRUE ~ Origin)) %>% 
-    mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>% 
-  mutate(Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
+         Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
                             TRUE ~ Tissue)) %>% 
   mutate(ZoomTissue = case_when((Tissue == 'Retina' & Sub_Tissue != 'Retina - Adult Tissue') | Tissue == 'ESC' ~ 'Early Retina',
-                                 TRUE ~ Sub_Tissue)) %>% 
-  mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
+                                TRUE ~ Sub_Tissue),
+         Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                            TRUE ~ Tissue),
-                        Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
-                        Group = factor(Group, levels = c('GTEx', 'Cornea', 'ESC', 'Lens', 'Retina', 'RPE'))) %>% 
+         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)) %>% 
+  mutate(Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>% 
   ggplot(aes(x=X1,y=X2)) +
@@
   #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
-  geom_point(size=2, alpha=0.8, position='jitter', aes(shape=Origin, colour = Group)) +
+  geom_point(size=4, alpha=1, aes(shape=Origin, colour = Group)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
   facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
-  #geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 50) + 
-  stat_chull(data=tsne_zoom,aes(fill=Cluster_Tissues), alpha = 0.3) +
+  geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
+  stat_chull(data=tsne_zoom %>% 
+               mutate(`Tissue Cluster` = Cluster_Tissues),aes(fill=`Tissue Cluster`), alpha = 0.3) +
   scale_color_discrete_sequential(palette = 'viridis') + 
-  scale_fill_discrete_sequential(palette = 'viridis') + 
+  scale_fill_viridis_d(option = 'plasma') +
   theme_bw() +
-  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
+         shape = guide_legend(ncol =2 )) +
   theme(axis.text.x = element_text(size = 6),
         axis.text.y = element_text(size = 6),
-        text = element_text(family = 'Helvetica', size = 6))
+        text = element_text(family = 'Linux Libertine O', size = 6))
 
@@
 All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’).
-Code used to generate data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+Code for the data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+The code used to create this manuscript's figures and layout can be accessed at https://github.com/davemcg/eyeIntegration_v1_app_manuscript.
 

### 2019-01-22 manuscript.Rmd
new text, supplementary figs and tables added, conclusion started

@@
 
-# Abstract
+# Abstract (I write this damn section last)
 
@@
 
+## RNA-seq is the predominant technology for deciphering transcriptomes
 From anterior to posterior, the human eye is composed of the cornea, lens, retina, and RPE (choroid).
@@
 
+## GTEx is a tremendously useful gene expression resource that, unfortunately, lacks eye tissues 
 The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues and hundreds unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
@@
 
+## eyeIntegration app reactively serves huge GTEx and human eye tissue dataset - EiaD
 The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provided the largest set of hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
@@
 
+## Manuscript as code
+This manuscript's figures (excepting (Figure 2), tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
+The knitr and pandoc program is used to lay out the figures and tables and output a docx file.
+The code that generates this manuscript can be found at https://github.com/davemcg/eyeIntegration_v1_app_manuscript.
+
 # Results
 
-## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes
+## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes that consistently cluster together by tissue type
 ```{bash, echo = F, eval = F}
@@
 ```
-We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data [@kosterSnakemakeScalableBioinformatics2012].
-In order to create a high quality final dataset across the 1,314 initial samples and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
+In order to create a high quality final dataset across the 1,314 initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
 To identify tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
@@
 
-```{r tsne_fig, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
+While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regards.
+We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from the human body tissues.
+Adult retina is most similar to the brain tissue, pituitary is similar to adult RPE, and skin and blood are similar to cornea.
+
+```{r tsne, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
 tsne_cap <- fig_cap(name='tSNE', caption = 't-SNE tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing.
@@
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
-The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r paste0('Tables ', tab_cap(name = 'tableSampleCounts', display = 'num'), ', ', tab_cap(name = 'tableSampleCountsGTEx', display = 'num'))`.
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`.
 The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
@@
 ```{r tableSampleCounts, echo = F, results='asis', warning = F, fig.cap = tableSampleCounts_cap}
-tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD and eyeIntegration contain a huge set of eye tissues')
+tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD contain a huge set of diverse eye tissues')
 table <- left_join(core_tight_2019  %>% 
@@
 
-```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap}
-tableSampleCountsGTEx_cap <- tab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
-# Output <div>
-cat("<div id=\"tbl:tableSampleCountsGTEx\">")
-regulartable(gtex) %>% 
-  width(width = c(0.5,1,0.5,3)) %>% 
-  flextable::align(align='center') %>% 
-  font(fontname = 'Linux Libertine O', part = 'all')
-#Output Caption (* * for italics)
-cat(paste0("\n*",tableSampleCountsGTEx_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
 ## eyeIntegration web app provides interactive visual portal to all data 
@@
 The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.
+
 ## Custom gene and tissue expression boxplots
@@
 
-## Differential expression and gene ontology enrichment tests 
+## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
 Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes.
@@
 
-## Gene expression in individual retina cells types with scRNA-seq 
+## Murine scRNA-seq enables testing of retina cell type specific expression
 We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
@@
 Code for the data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
-The code used to create this manuscript's figures and layout can be accessed at https://github.com/davemcg/eyeIntegration_v1_app_manuscript.
 
-# Conclusions
+# Conclusions (need more fleshing out)
+
+EiaD 2019 contains a large set of carefully curated, robustly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from GTEx.
+It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov.
+The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues.
+Furthermore, scRNA-seq datasets have been integrated, allowing for examination of cell type specific gene expression.
+As more human and non-human primate samples become publicly available, we will be updating the section eyeIntegration.
+
+We have also provided a small example of the very close transcriptomic distance between fetal tissue - organoid - iPSC based retina.
+As human fetal tissue is extremely difficult to obtain, our dataset agnostically demonstrates the viability of using organoid and iPSC approaches to study human retina development.
+
+Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates are substantially easier to add, and because all the processing is in modular pieces, and it is easier to add new analysis steps.
+In the future we plan on regularly adding new samples to EiaD, and also plan on more substantial steps like building *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic data.
+
+# Supplementary Figures and Tables
+
+```{r snakemakeDAG, echo = F, message = F, warning=F, fig.height = 8, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
+snakemakeDAG_cap <- supFig_cap(name='snakemakeDAG', caption = 'Snakemake pipeline to create EiaD 2019 consists of small modular compute sections to ensure sample tracking through the full pipeline')
+
+knitr::include_graphics('figures_and_tables/2019_workflow.png')
+```
+
+```{r allSamplePCA, echo = F, message = F, warning=F, fig.height = 8, fig.width = 10, fig.pos='H', fig.cap = pca_cap}
+# fig.pos = 'H' puts figure here, not where knitr / pandoc think is best
+
+pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples demonstrates non-eye tissue most similar to adult retina is the brain, pituitary most similar to adult RPE, and skin and blood similar to cornea.')
+
+library(cowplot)
+# data processed with src/tsne_and_pca_calcs.R
+
+load('data/dimRed_plot_data.Rdata')
+
+# pca
+make_pca_tissue <- function(tissue = 'Brain'){
+  ggplot(dimRed_plot_prep_pca %>% 
+           mutate(Origin = case_when(study_accession == 'SRP012682' ~ 'Adult Tissue',
+                                     TRUE ~ as.character(Origin))) %>% 
+           filter(!is.na(Origin)) %>% 
+           mutate(Group = case_when(Tissue == !!(tissue) ~ Tissue,
+                                    study_accession == 'SRP012682' ~ 'GTEx',
+                                    TRUE ~ Tissue)) %>% 
+           mutate(Group = factor(Group, levels= c('Cornea', 'ESC', 'Retina', 'RPE', !!(tissue), 'GTEx'))),
+         aes(x=`Dimension 1`,y=`Dimension 2`)) +
+    scale_shape_manual(values=c(0:20,35:50)) +
+    #ggtitle(paste0('All Tissue PCA: Highlight ', tissue)) +
+    geom_point(size=6, alpha=0.1, aes(colour=Group)) +
+    geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+    xlab('PCA 1') + ylab('PCA 2') +
+    theme_bw() +
+    theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
+    scale_color_discrete_sequential(palette = 'viridis') + 
+    guides(colour = guide_legend(override.aes = list(alpha = 1)))
+}
+theme_set(theme_cowplot(font_size=10,font_family = "Helvetica")) 
+plot_grid(make_pca_tissue('Brain'), make_pca_tissue('Pituitary'),
+          make_pca_tissue('Blood'), make_pca_tissue('Skin'),
+          labels = c('A: Brain','B: Pituitary','C: Blood','D: Skin'),
+          align = 'hv')
+```
+
+```{r metadataTable, echo = F,  results='asis', fig.pos = 'H', fig.cap = metadata_cap}
+metadata_cap <- supTab_cap(name='metadataTable', caption = 'Full metadata for 20 random eye samples. Full metadata available as supplementary file.')
+set.seed(1534)
+cat("<div id=\"tbl:metadataTable\">")
+flextable(core_tight_2019 %>% select(-run_accession) %>% unique() %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
+  font(fontname = 'Linux Libertine O', part = 'all') %>% 
+  fontsize(size = 6) %>% 
+   width(width = c(0.5,0.5,0.5,3,1,0.5,0.5,0.5,0.5))
+#Output Caption (* * for italics)
+cat(paste0("\n*",metadata_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap, fig.pos = 'H'}
+tableSampleCountsGTEx_cap <- supTab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
+# Output <div>
+cat("<div id=\"tbl:tableSampleCountsGTEx\">")
+regulartable(gtex) %>% 
+  width(width = c(0.5,1,0.5,3)) %>% 
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",tableSampleCountsGTEx_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
 

### 2019-01-23 manuscript.Rmd
rewrite conclusions

@@
 
-EiaD 2019 contains a large set of carefully curated, robustly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from GTEx.
+EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project.
 It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov.
 The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues.
-Furthermore, scRNA-seq datasets have been integrated, allowing for examination of cell type specific gene expression.
-As more human and non-human primate samples become publicly available, we will be updating the section eyeIntegration.
+Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression.
+Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and once the raw data becomes publicly available, we will be updating the section eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
 
-We have also provided a small example of the very close transcriptomic distance between fetal tissue - organoid - iPSC based retina.
-As human fetal tissue is extremely difficult to obtain, our dataset agnostically demonstrates the viability of using organoid and iPSC approaches to study human retina development.
+As human fetal tissue is extremely difficult to obtain and not amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed.
+Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level.
+This result underscores the value of these approaches in building models for human retina developmental.
 
-Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates are substantially easier to add, and because all the processing is in modular pieces, and it is easier to add new analysis steps.
-In the future we plan on regularly adding new samples to EiaD, and also plan on more substantial steps like building *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic data.
+Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with minimal effort, and because all the processing is in modular pieces it is easier to add new analysis steps.
+In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic datasets.
 

### 2019-01-23 manuscript.Rmd
change pca plot point coloring, move libraries and data pull to top of doc

@@
 library(ggforce)
+library(cowplot)
+library(viridis)
 
@@
 load('data/human_tx_studies.Rdata')
+# source('src/tsne_and_pca_calcs.R')
+load('data/dimRed_plot_data.Rdata')
 ```
@@
 All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’).
-Code for the data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+Code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
 
@@
 
-library(cowplot)
-# data processed with src/tsne_and_pca_calcs.R
-
-load('data/dimRed_plot_data.Rdata')
+colors = viridis(n=6, direction = -1)
+colors[5] <- "red"
 
@@
     theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
-    scale_color_discrete_sequential(palette = 'viridis') + 
+    scale_color_manual(values = colors) + 
     guides(colour = guide_legend(override.aes = list(alpha = 1)))
 }
-theme_set(theme_cowplot(font_size=10,font_family = "Helvetica")) 
+theme_set(theme_cowplot(font_size=10)) 
 plot_grid(make_pca_tissue('Brain'), make_pca_tissue('Pituitary'),

### 2019-01-23 manuscript.Rmd
abstract added, doodle figure added, re-organized results, added visitor map

@@
 ---
-title: 'Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00'
+title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00'
 author:
@@
 csl: investigative-ophthalmology-and-visual-science.csl
+abstract: "PURPOSE: To provide the eye community with a high quality RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
+METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across over 44 tissues in the GTEx project as a non-eye reference set.
+We then queried several public repositories to find all healthy, non perturbed, human eye tissue RNA-seq samples.
+The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote.
+The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and likely mislabeling, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.00, https://eyeIntegration.nei.nih.gov) to display this data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues.
+In a powerful demonstration of the value of this unified pan-eye dataset, we agnostically show that, at a transcriptome level, fetal, stem cell, and organoid retina tissues are highly similar.
+CONCLUSION: The eyeIntegration v1.00 web app serves the pan-eye and body transcriptome dataset, EiaD.
+This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
+keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
 ---
@@
 library(viridis)
+library(rworldmap)
 
@@
 load('data/dimRed_plot_data.Rdata')
+# run 'src/viewing_stats.Rmd'
+load('data/visitor_stats.Rdata')
+# write supplementary file for metadata
+write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
 ```
-
-# Abstract (I write this damn section last)
 
@@
 We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test.
-We extracted significant genes (FDR p <.05) for each eye sub-tissue vs body comparisons and used these to calculate GO enrichment for each eye sub-tissue.
+We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
 The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
@@
 
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sample along with `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
+
+```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
+tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type  from 2017 to 2019 EiaD')
+
+core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
+
+core_both %>% filter(grepl('Retina|Cornea|Lens|RPE|ESC', Tissue)) %>% unique() %>% mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>% group_by(Tissue, Sub_Tissue, Version) %>% summarise(Count=n()) %>%
+  mutate(Dataset = Version) %>% 
+  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
+  theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
+  scale_fill_discrete_sequential(palette = 'viridis')  
+```
+
+Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19, 2017.
+We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`.
+The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
+Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
+Another substantial addition to the 2019 EiaD are non-protein coding genes.
+We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
+
 ## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes that consistently cluster together by tissue type
@@
 ```
-We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
-In order to create a high quality final dataset across the 1,314 initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+
+```{r workflowDoodle, echo = F, fig.width = 10, fig.height = 10, results='asis', fig.cap = workflowDoodle_cap}
+workflowDoodle_cap <- fig_cap(name='workflowDoodle', caption = paste0('Raw RNA-seq data from the SRA is run through our pipeline to create the EiaD, which is used by eyeIntegration app to serve interactive gene expression visualizations across ', core_tight_2019 %>% pull(Sub_Tissue) %>% unique() %>% length(), ' tissues'))
+
+knitr::include_graphics('figures_and_tables/flow_sketch.png')
+```
+
+We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r fig_cap(name='workflowDoodle', display = 'cite')`, `r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
+In order to create a high quality final dataset across the `r core_tight_2019 %>% select(-run_accession) %>% unique() %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
 To identify tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
@@
   guides(colour = guide_legend(override.aes = list(alpha = 1)))
-```
-
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sample along with `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
-Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19th, 2017.
-We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
-The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`.
-The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
-Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
-Another substantial addition to the 2019 EiaD are non-protein coding genes.
-We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
-
-```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
-tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type  from 2017 to 2019 EiaD')
-
-core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
-
-core_both %>% filter(grepl('Retina|Cornea|Lens|RPE|ESC', Tissue)) %>% unique() %>% mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>% group_by(Tissue, Sub_Tissue, Version) %>% summarise(Count=n()) %>%
-  mutate(Dataset = Version) %>% 
-  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
-  theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
-  scale_fill_discrete_sequential(palette = 'viridis')  
 ```
@@
 ## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
-Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes.
+Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
 The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
@@
 The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values.
+Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (`r supTab_cap(name = 'limmaTests', display = 'cite')`).
 The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format.
@@
 
-# Conclusions (need more fleshing out)
+# Discussion
 
@@
 
+```{r visitorMap, echo = F, results='asis', fig.pos = 'H', fig.cap = visitorMap_cap, fig.height=4, fig.width=10}
+visitorMap_cap <- supFig_cap(name='visitorMap', caption = paste('As of January 2019, eyeIntegration has had usage across', cities %>% unique() %>% length(), 'cities and', geo_locate$countries %>% unique() %>% length(), 'countries.'))
+
+world <- map_data("world")
+world <- world[world$region != "Antarctica",]
+
+ggplot(data = geo_locate, aes(x = longitudes, y = latitudes, colour = countries)) + 
+  geom_map(data=world, map=world,
+           aes(x=long, y=lat, map_id=region),
+           color="#090D2A", fill="lightgray", size=0.05, alpha=1/4) + 
+  geom_point() + 
+  theme_void() +
+  scale_color_viridis_d()
+
+```
+
 ```{r metadataTable, echo = F,  results='asis', fig.pos = 'H', fig.cap = metadata_cap}
@@
 
+```{r limmaTests, echo = F, results='asis', warning = F, fig.cap = limmaTests_cap, fig.pos = 'H'}
+limmaTests_cap <- supTab_cap(name='limmaTests', caption = 'Thousands of differentially expressed genes found (q < 0.01) for the 105 comparisons we ran.')
+# Output <div>
+cat("<div id=\"tbl:limmaTests\">")
+regulartable(q01_limma_gene_sig) %>% 
+  width(width = c(4,2)) %>% 
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",limmaTests_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+# Acknowledgements
+
+We would like the thank the dozens of groups who provided the raw data required to create this project.
+We are keeping a running list of the projects and citations at https://github.com/davemcg/eyeIntegration_app/blob/master/inst/citations.md and strongly encourage anyone who uses EiaD and eyeIntegration to cite relevant projects.
+We would also like to thank Kapil Bharti, Robert Hufnagel, and Brian Brooks for their continuous set of critiques and suggestions of the eyeIntegration app over the past two years.
+This work utilized the computational resources and of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
+
 # Bibliography

### 2019-01-23 src/data_pull.R
now outputs limma de info

@@ -11,4 +11,12 @@ load('data/human_tx_studies.Rdata')
 tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
 gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tx_anno <- gene_pool_2019 %>% tbl('tx_IDs') %>% as_tibble()
-save(core_tight_2017, core_tight_2019, tsne_50, gene_anno, tx_anno, file = 'data/data_pull.Rdata')
+limma_DE_tests <- gene_pool_2019 %>% tbl('limma_DE_tests') %>% as_tibble()
+q01_limma_gene_sig <- gene_pool_2019 %>% tbl('limma_DE_gene') %>% filter(adj.P.Val < 0.01) %>% group_by(Comparison) %>% summarise(Count=n()) %>% as_tibble()
+save(core_tight_2017, 
+     core_tight_2019, 
+     tsne_50, gene_anno, 
+     tx_anno, 
+     limma_DE_tests, 
+     q01_limma_gene_sig,
+     file = 'data/data_pull.Rdata')

### 2019-01-23 src/viewing_stats.Rmd
viewing stats code

---
title: "eyeIntegration Viewing Stats"
output: html_notebook
---

Pull in IP addresses, filter to only keep IP addresses over 11 views (I think any 11 or below is a bot, since they are so common)

`eyeIntegration.stats` is generated as follows:

1. Select all `Daily_eyeinegration_frequent_IP_access_list` emails in Outlook
2. Drag all emails to empty folder
3. Run `cat *eml | grep "^[[:digit:]]" > eyeIntegration.stats`

```{r}
library(tidyverse)
library(rworldmap)
library(ipstack)
ipaddresses <- read_delim('~/Desktop/stats/eyeIntegration.stats', col_names = F, delim = '\\s+') %>% 
  separate(col = X1, into = c('IP Address', 'Page Views'), sep='\\s+') %>% 
  mutate(`Page Views` = as.numeric(`Page Views`)) %>%  
  filter(`Page Views` > 11)
```

Total page view from May 2017 to November 2017
```{r}
sum(ipaddresses$`Page Views`)
```

Get latitude and longitude for all unique IPs (takes a few minutes to query)
```{r}
unique_IP <- ipaddresses$`IP Address` %>% unique()
ip_info <- list()
for (i in unique_IP){
  ip_info[[i]] <- ip_lookup(i, ipstack_api_key = '7d06f26c4f0f61bd9edd20df0219f874', hostname = T)
}

```

Process and clean up
```{r}
ips <- ''
latitudes <- ''
longitudes <- ''
cities <- ''
countries <- ''
for (i in unique_IP){
  if (!is.null(ip_info[[i]]$country_name)){
  ips <- c(ips, ip_info[[i]]$ip)
  latitudes <- c(latitudes, ip_info[[i]]$latitude)
  longitudes <- c(longitudes, ip_info[[i]]$longitude)
  cities <- c(cities, ip_info[[i]]$city)
  countries <- c(countries, ip_info[[i]]$country_name)
  }
}

geo_locate <- cbind(latitudes, longitudes, countries) %>% 
  data.frame() %>% 
  mutate(latitudes = as.numeric(as.character(latitudes)), longitudes = as.numeric(as.character(longitudes))) %>% 
  filter(abs(latitudes) > 0) #remove empty coordinates

save(cities, geo_locate, file = '../data/visitor_stats.Rdata')
```

Interactive Map
```{r}
# library(ggiraph)
# world <- map_data("world")
# world <- world[world$region != "Antarctica",]

#world <- c(geom_polygon(aes(long,lat,group=group), size = 0.1, colour= "#090D2A", fill="white", alpha=0.8, data=worldmap))
# code <- ggplot() + geom_map(data=world, map=world,
#                     aes(x=long, y=lat, map_id=region),
#                     color="#090D2A", fill="white", size=0.05, alpha=1/4) + 
#   geom_point_interactive(aes(x = geo_locate$longitudes, y = geo_locate$latitudes, tooltip=geo_locate$cities)) + theme_void() 
# ggiraph(code = print(code) )
```

Regular Map
```{r, fig.width = 6, fig.height=2}
ggplot(data = geo_locate, aes(x = longitudes, y = latitudes, colour = countries)) + 
  geom_map(data=world, map=world,
           aes(x=long, y=lat, map_id=region),
           color="#090D2A", fill="lightgray", size=0.05, alpha=1/4) + 
  geom_point() + 
  theme_void() +
  scale_color_viridis_d()

```

### 2019-01-24 manuscript.Rmd
hand comment out supplement file writing

@@
 # write supplementary file for metadata
-write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
+# write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
 ```

### 2019-01-26 manuscript.Rmd
fix number issue with counting lanes instead of samples, update text with Vinny's suggestions, tweak tsne zoom fig to match zoom coordinates

--- 2019-01-23 manuscript.Rmd
@@
 # write supplementary file for metadata
-write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
+# write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
+
+core_tight_2019 <- core_tight_2019 %>% mutate(Origin = case_when( grepl('Transformed', Sub_Tissue, ignore.case = T) ~ 'Cell Line',
+                                               grepl('Fetal', Sub_Tissue) ~ 'Fetal Tissue',
+                                               grepl('Stem', Sub_Tissue) ~ 'Stem Cell',
+                                               grepl('Organoid', Sub_Tissue) ~ 'Organoid',
+                                               grepl('Cell Line', Sub_Tissue) ~ 'Cell Line',
+                                               TRUE ~ 'Adult Tissue')) %>% 
+  select(-run_accession) %>% 
+                     unique()
+
+# set up maps
+visitorMap_cap <- supFig_cap(name='visitorMap', caption = paste('As of January 2019, eyeIntegration has had usage across', cities %>% unique() %>% length(), 'cities and', geo_locate$countries %>% unique() %>% length(), 'countries.'))
+
+world <- map_data("world")
+world <- world[world$region != "Antarctica",]
+
 ```
@@
 ## RNA-seq is the predominant technology for deciphering transcriptomes
-From anterior to posterior, the human eye is composed of the cornea, lens, retina, and RPE (choroid).
-The development and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, or transcriptome.
-Today, RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome.
-Analysis of gene patterns across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
+From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid.
+The differentiation, development, and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, also known as the transcriptome.
+Today, RNA-sequencing (RNA-seq) is the predominant technology for quantifying the transcriptome.
+Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
 To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids.
@@
 ## GTEx is a tremendously useful gene expression resource that, unfortunately, lacks eye tissues 
-The Gene Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues and hundreds unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
+The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
 Unfortunately, no eye tissues are included.
-Two groups have quantified huge portions of the RNA-seq data, including some human eye tissues, from the Sequence Read Archive (SRA): recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
-However no curation of the sample level metadata was done, so it is very difficult to parse out which eye tissues are exactly present and, more importnatly, to determine whether any samples were chemically or genetically perturbed.
+The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified huge portions of the RNA-seq data, including some human eye tissues, from the (SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
+However no curation of the sample level metadata was done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed.
 More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016].
@@
 
-## eyeIntegration app reactively serves huge GTEx and human eye tissue dataset - EiaD
+## eyeIntegration app reactively serves huge GTEx and human eye tissue datasets (EiaD)
 The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provided the largest set of hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
-This dataset was served in a responsive web app allowing for quick comparisons across many eye tissues and dozens of other body tissues.
-In order to better serve the eye research community, we have re-written the bioinformatic pipeline that creates the Eye in a Disk (EiaD) datasets for eyeIntegration to nearly fully automate the EiaD creation, ensure full reproducbility of the results, allow for external data comparison, provide consistent sample quality control, and improve efficiency for future sample updates.
-Additionally, we are prototyping new tools to display murine single cell RNA-seq (scRNA-seq) data to provide researchers access to information about cell type specific gene expression in retinal development.
+This responsive web app allowed for quick comparisons across many eye tissues and dozens of other body tissues.
+The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples.
+The interactive nature of some of the steps precludes efficient and regular updates for the data.
+
+In order to better serve the eye research community with regular, versioned updates, we have re-written the bioinformatic pipeline that creates the eye and body RNA-seq dataset for eyeIntegration.
+We call this reproducible and versioned transcriptome dataset Eye in a Disk (EiaD).
+The pipeline nearly fully automates the EiaD creation, ensure full reproducbility of the results, allow for external data comparison, providse consistent sample quality control, and improves efficiency for future sample updates.
 The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples.
 The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands.
+Additionally, we are prototyping new tools to display single cell RNA-seq (scRNA-seq) data to provide researchers access to cell type specific information about gene expression across murine retinal development.
 
@@
 ## Identifying potential eye samples with exhaustive search of SRA
-We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|ocular|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
+We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
 We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids, and immortalized cell lines.
@@
 We downloaded the relevant SRA files for each sample directly from the NCBI ftp server using the file transfer software Aspera.
-SRA files were converted to FASTQ format using the tool fastq-dump from the SRAtoolkit software package.
+SRA files were converted to FASTQ format using the tool fastq-dump from the SRAtoolkit software package [@leinonenSequenceReadArchive2011].
 Samples only available in the BAM format were converted to FASTQ format using SAMTools [@liSequenceAlignmentMap2009].
 Sample transcriptomes were quantified using the alignment free quantification software Salmon, using transcriptomic index built from gencode v28 protein coding transcript sequences using the transcriptomic aligner Salmon [@harrowGENCODEReferenceHuman2012; @patroSalmonProvidesFast2017a].
-Using the resulting expression quantification, we identified lowly or unused transcripts within annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene as per Sonneson et al [@sonesonIsoformPrefilteringImproves2016].
+Using the resulting expression quantification, we identified lowly or unused transcripts within the gencode annotation, and removed transcripts that accounted for 5% or less of the total expression for its parent gene as per Sonneson et al [@sonesonIsoformPrefilteringImproves2016].
 Samples were re-quantified against a transcriptomic index built on the filtered transcript sequences.
@@
 We began the quality control procedure by removing samples with a Salmon calculated mapping rate less than 50%.
-We further filtered samples by calculating the median gene expression for each sample and only kept sample that had a median count >2 across all genes.
+We further filtered samples by calculating the median gene expression for each sample and only kept samples that had a median count >2 across all genes.
 We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples.
-To reduce the noise experimental variability between each study, we normalized samples by library using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
+To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
 
-We generated sample level two-dimensional groupings by with the t-Stochastic Neighbor Embedding (t-SNE) algorithm, and then calculated the center of the resulting sub-tissue clusters by finding the average coordinate of all samples of a given sub-tissue type [@maatenAcceleratingTSNEUsing2014].
-We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type.
-We calculated the euclidean distance of each sample from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance.
+We generated sample level two-dimensional groupings by with the t-Stochastic Neighbor Embedding (t-SNE) algorithm [@maatenAcceleratingTSNEUsing2014].
+We then calculated the center of the resulting sub-tissue clusters by finding the average two-dimensional coordinates of all samples of a given sub-tissue type (p1).
+We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type by measuring the euclidean distance of each sample (p2) from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance with the equation $dist = \sqrt\sum(p1-p2)^{2}$.
 
@@
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test.
+We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
@@
 ## eyeIntegration web app and R package
-The data generated in the above steps is consolidated into a SQLite database, with the original EiaD data from 2017 and the new 2019 EiaD dataset each getting a separated database file.
+The data generated in the above steps is consolidated into a SQLite database, with the original dataset for eyeIntegration and the new 2019 EiaD dataset each getting a separate database file.
 The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov).
-The app can be deployed on a local computer with 30GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
+The app can be deployed on a local computer with 35GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
 
@@
 While the sample search and metadata parsing in a semi-curated process, the processing from the raw data to the creation of the SQLite EiaD database underlying eyeIntegration is wrapped in a Snakemake pipeline, which ensures full reproducibility of the results [@kosterSnakemakeScalableBioinformatics2012].
-We make the code for the pipeline available at https://github.com/davemcg/eyeIntegration_data_build.
+We make the code for the pipeline available at https://github.com/davemcg/EiaD_build.
 
 ## scRNA-seq processing
-The eyeIntegration site currently hosts two large scRNA-seq datasets from Macoscko et al.
-and a unpublished set kindly made available to the public from Clark et al [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
-We use the processed gene count data from each group, as well as the cluster assignments that specify what cell type each individual cell is.
+The eyeIntegration site, as of January 2019, hosts two large scRNA-seq datasets from Macosko et al.
+and Clark et al [@macoskoHighlyParallelGenomewide2015a; @clarkComprehensiveAnalysisRetinal2018].
+We use the processed gene count data from each group, as well as their cluster assignments which specify what cell type each individual cell is.
 The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration.
 We also display t-SNE and UMAP-based two-dimensional visualizations of the Macosko and Clark data, respectively, in the web app.
-For detail so the t-SNE processing we did on the Macosko dataset, see the methods of [@Bryan_2018].
+For detail so the t-SNE processing we did on the Macosko dataset, see the methods of Bryan et al [@Bryan_2018].
 
@@
 
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sample along with `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and many sub-tissue types
 
 ```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
-tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type  from 2017 to 2019 EiaD')
+tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type from 2017 to 2019 EiaD.
+We also improved the metadata labelling for the cornea samples, now delineating endothelial and epithelial tissues.')
 
-core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
+core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()
 
-core_both %>% filter(grepl('Retina|Cornea|Lens|RPE|ESC', Tissue)) %>% unique() %>% mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>% group_by(Tissue, Sub_Tissue, Version) %>% summarise(Count=n()) %>%
+core_both %>% filter(grepl('Retina|Cornea|Lens|RPE', Tissue)) %>% unique() %>% 
+  #mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>%
+  group_by(Tissue, Sub_Tissue, Version) %>% summarise(Count=n()) %>% 
   mutate(Dataset = Version) %>% 
@@
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
-The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`.
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison (`r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`).
 The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
+
 Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
+We have also substantially improved the granualarity of the cornea tissue metadata, now delinerating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
 Another substantial addition to the 2019 EiaD are non-protein coding genes.
@@
 We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r fig_cap(name='workflowDoodle', display = 'cite')`, `r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
-In order to create a high quality final dataset across the `r core_tight_2019 %>% select(-run_accession) %>% unique() %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
-To identify tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
+In order to create a high quality final dataset across the `r core_tight_2019 %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+To identify tube mislabelling or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
 
@@
 cluster_stats <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
 # set up for ggplot
-tsne_50_prep <- tsne_50 %>% 
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid'))) %>%
+tsne_50_prep <- tsne_50 %>% select(-Origin) %>% 
+  left_join(., core_tight_2019 %>% select(sample_accession, Origin), by = 'sample_accession') %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
 table <- left_join(core_tight_2019  %>% 
-                     select(-run_accession) %>% 
-                     unique() %>% 
                      filter(study_accession != 'SRP012682',Tissue %in% c('RPE','Lens','Cornea', 'Retina')) %>% 
@@
                    core_tight_2019  %>%
-                     select(-run_accession) %>% 
-                     unique() %>% 
                      filter(study_accession != 'SRP012682', Tissue %in% c('RPE','Lens','Cornea', 'Retina'), Kept == 'Kept') %>% 
@@
 gtex_tissues <- core_tight_2019 %>% 
-  select(-run_accession) %>% 
-  unique() %>%
   filter(study_accession == 'SRP012682', Kept == 'Kept') %>%
@@
 gtex <- cbind('GTEx', core_tight_2019 %>% 
-                select(-run_accession) %>% 
-                unique() %>%
                 filter(study_accession == 'SRP012682') %>% nrow(),
               core_tight_2019 %>% 
-                select(-run_accession) %>% 
-                unique() %>%
                 filter(study_accession == 'SRP012682', Kept == 'Kept') %>% nrow(),
@@
 ## eyeIntegration web app provides interactive visual portal to all data 
-The EiaD 2019 dataset is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
+The EiaD 2019 dataset is used directly by the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
 The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
@@
 The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
-A boxplot is then generated afer hitting the "Re(Draw) Plot" with overlaid individual data points.
+A boxplot is then generated afer hitting the "Re(Draw) Plot" button with overlaid individual data points.
+On mouse-over, the metadata for the individual sample is displayed.
 A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue.
@@
 ## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
-Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in the GTEx data set, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
+Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
 The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
@@
 ## EiaD 2019 demonstrates that organoids, iPSC-derived, and fetal tissue retina have closely related transcriptomes
-There are, currently, three approaches to collecting human fetal retina: 1.
-post-mortem tissue, 2.
+There are, currently, three approaches to studying developing human retina: 1.
+post-mortem fteal tissue, 2.
 organoids, and 3.
 iPSC-derived cells.
-As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to generating developing retina compare at a transcriptomic level.
+As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level.
 Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
@@
 The zoom inset shows the retina samples.
-The "Tissue Cluster" shading shows the cluster membership of the three major groups.
+The "Sub-Tissue Cluster" shading shows the cluster membership of the three major groups.
 The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.')
 
-retina_plus_group <- tsne_50 %>% filter(X1 > 34, X1 < 51, X2 > -25, X2 < 21,sample_accession!='SRS1572426')
+retina_plus_group <- tsne_50 %>% filter(X1 > 30, X1 < 50, X2 > -17, X2 < 16)
 dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)
@@
 retina_plus_group_prep <- retina_plus_group %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
                             TRUE ~ Origin)) %>% 
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell Line', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
          Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
@@
   stat_chull(data=tsne_zoom %>% 
-               mutate(`Tissue Cluster` = Cluster_Tissues),aes(fill=`Tissue Cluster`), alpha = 0.3) +
+               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
   scale_color_discrete_sequential(palette = 'viridis') + 
@@
 Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression.
-Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and once the raw data becomes publicly available, we will be updating the section eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
+Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
 
-As human fetal tissue is extremely difficult to obtain and not amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed.
+As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed.
 Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level.
-This result underscores the value of these approaches in building models for human retina developmental.
+This result underscores the value of these approaches in building models for human retina developmental .
 
@@
     theme_bw() +
-    theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
+   # theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
     scale_color_manual(values = colors) + 
@@
 
-```{r visitorMap, echo = F, results='asis', fig.pos = 'H', fig.cap = visitorMap_cap, fig.height=4, fig.width=10}
-visitorMap_cap <- supFig_cap(name='visitorMap', caption = paste('As of January 2019, eyeIntegration has had usage across', cities %>% unique() %>% length(), 'cities and', geo_locate$countries %>% unique() %>% length(), 'countries.'))
-
-world <- map_data("world")
-world <- world[world$region != "Antarctica",]
+```{r visitorMap, echo = F, results='asis', fig.pos = 'H', fig.cap = visitorMap_cap, fig.height=4, fig.width=10, warning=F}
 
@@
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% select(-run_accession) %>% unique() %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
+flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
   font(fontname = 'Linux Libertine O', part = 'all') %>% 
@@
 We would also like to thank Kapil Bharti, Robert Hufnagel, and Brian Brooks for their continuous set of critiques and suggestions of the eyeIntegration app over the past two years.
-This work utilized the computational resources and of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
+This work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
 

### 2019-01-26 src/data_pull.R
update data with corrected EiaD (fixed tissue mislabels)

@@ -3,12 +3,12 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_01_25.sqlite')
 gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 load('data/human_tx_studies.Rdata')
-tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
+tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 50) %>% as_tibble()
 gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tx_anno <- gene_pool_2019 %>% tbl('tx_IDs') %>% as_tibble()
 limma_DE_tests <- gene_pool_2019 %>% tbl('limma_DE_tests') %>% as_tibble()

### 2019-01-28 manuscript.Rmd
text updates, move sup figs around

@@
 Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
-To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids.
-These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
+To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
+These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014b; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014b; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015a; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
 
@@
 We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
-We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids, and immortalized cell lines.
+We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, stem cell derived tissue, organoids, and immortalized cell lines.
 In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
@@
 
-## EiaD 2019 demonstrates that organoids, iPSC-derived, and fetal tissue retina have closely related transcriptomes
+## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
 There are, currently, three approaches to studying developing human retina: 1.
@@
 organoids, and 3.
-iPSC-derived cells.
+stem cell-derived cells.
 As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level.
-Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
+Should organoid and stem cell-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
 
 To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
-Here we see three distinct groupings: adult retina (2), developing retina (3), and iPSC/undifferentiated (1).
+Here we see three distinct groupings: adult retina (2), developing retina (3), and stem cell/undifferentiated (1).
 We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
-All of the iPSC-based retina samples are older than 50 days.
+All of the stem cell-based retina samples are older than 50 days.
 
 ```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F, warning = F}
-tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes.
+tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, stem cell retina, and fetal retina tissue have highly similar transcriptomes.
 The zoom inset shows the retina samples.
@@
 Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression.
-Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
+Several human and non-human primate studies have been posted in the past year on the bioRxiv pre-print server.
+As the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
 
-As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed.
+As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for stem cell and organoid based developmental methods to be developed.
 Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level.
@@
 # Supplementary Figures and Tables
+
+```{r visitorMap, echo = F, results='asis', fig.pos = 'H', fig.cap = visitorMap_cap, fig.height=4, fig.width=10, warning=F}
+
+ggplot(data = geo_locate, aes(x = longitudes, y = latitudes, colour = countries)) + 
+  geom_map(data=world, map=world,
+           aes(x=long, y=lat, map_id=region),
+           color="#090D2A", fill="lightgray", size=0.05, alpha=1/4) + 
+  geom_point() + 
+  theme_void() +
+  scale_color_viridis_d()
+
+```
 
@@
           align = 'hv')
-```
-
-```{r visitorMap, echo = F, results='asis', fig.pos = 'H', fig.cap = visitorMap_cap, fig.height=4, fig.width=10, warning=F}
-
-ggplot(data = geo_locate, aes(x = longitudes, y = latitudes, colour = countries)) + 
-  geom_map(data=world, map=world,
-           aes(x=long, y=lat, map_id=region),
-           color="#090D2A", fill="lightgray", size=0.05, alpha=1/4) + 
-  geom_point() + 
-  theme_void() +
-  scale_color_viridis_d()
-
 ```

### 2019-01-30 manuscript.Rmd
added funding note
citation update, new figure

--- 2019-01-26 manuscript.Rmd
@@
 Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
-To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids.
+To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and adult, post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
@@
 
-knitr::include_graphics('figures_and_tables/flow_sketch.png')
+knitr::include_graphics('figures_and_tables/simplified_workflow.png')
 ```

### 2019-01-30 manuscript.Rmd

--- 2019-01-28 manuscript.Rmd
@@
 ## RNA-seq is the predominant technology for deciphering transcriptomes
+
 From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid.
@@
 
+# Funding
+
+This research was supported by the Intramural Research Program of the National Eye Institute, National Institutes of Health.
+
 # Bibliography

### 2019-02-14 src/retina_development_heatmaps.R
prototype heatmaps for ret development

@@ -1,0 +1,66 @@
+library(viridis)
+library(ComplexHeatmap)
+library(tidyverse)
+library(circlize)
+rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','SLX2')
+pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
+
+gene <- pr
+query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
+p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as.tibble()) %>% 
+  as.tibble()
+
+ESC <- p %>% 
+  filter(Tissue == 'ESC') %>% 
+  mutate(Days = 0) %>% group_by(ID, Days) %>% summarise(value = mean(value)) %>% mutate(Days = as.integer(Days))
+organoid_swaroop_GFP <- p %>% 
+  filter(sample_accession %in% (meta_ret %>% filter(Sub_Tissue == 'Retina - 3D Organoid Stem Cell') %>% pull(sample_accession))) %>%
+  filter(!grepl('GFP negative', sample_attribute), study_accession != 'SRP159246') %>% 
+  left_join(meta_ret, by = 'sample_accession') %>% group_by(ID, Days) %>% summarise(value = mean(value)) %>% mutate(Days = as.integer(Days))
+organoid_swaroop_GFPneg <- p %>% 
+  filter(sample_accession %in% (meta_ret %>% filter(Sub_Tissue == 'Retina - 3D Organoid Stem Cell') %>% pull(sample_accession))) %>%
+  filter(grepl('GFP negative', sample_attribute), study_accession != 'SRP159246') %>% 
+  left_join(meta_ret, by = 'sample_accession') %>% group_by(ID, Days) %>% summarise(value = mean(value)) %>% mutate(Days = as.integer(Days))
+organoid_johnston <- p %>% 
+  filter(sample_accession %in% (meta_ret %>% filter(Sub_Tissue == 'Retina - 3D Organoid Stem Cell') %>% pull(sample_accession))) %>%
[…]

### 2019-02-14 manuscript.Rmd

@@
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of sub-tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
+We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
@@
 Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
+We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues.
 The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).

### 2019-02-14 src/retina_development_heatmaps.R
change colors

@@ -4,8 +4,9 @@ library(tidyverse)
 library(circlize)
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','SLX2')
 pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
+progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
 
-gene <- pr
+gene <- progenitor
 query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
   left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as.tibble()) %>% 
@@ -45,7 +46,8 @@ colnames(y) <- y['Days',]
 y <- y[-1,]
 
 two <- Heatmap(log2(y), cluster_columns = F, column_title = 'Swaroop 3D Organoid\nGFP Neg',
-               col = colorRamp2(c(-0, 15), viridis(2)))
+               col = colorRamp2(c(-0, 15), viridis(2)),
+               show_heatmap_legend = F)
 
 x <- rbind(organoid_swaroop_GFPneg, ESC)
 y <- x %>% spread(ID, value) %>% t()
@@ -53,7 +55,8 @@ colnames(y) <- y['Days',]
 y <- y[-1,]
 
 three <- Heatmap(log2(y), cluster_columns = F, column_title = 'Swaroop 3D Organoid\nGFP Neg',
-               col = colorRamp2(c(-0, 15), viridis(2)))
+               col = colorRamp2(c(-0, 15), viridis(2)),
+               show_heatmap_legend = F)
 
 x <- rbind(organoid_johnston, ESC)
 y <- x %>% spread(ID, value) %>% t()
@@ -61,6 +64,7 @@ colnames(y) <- y['Days',]
 y <- y[-1,]
 
 four <- Heatmap(log2(y), cluster_columns = F, column_title = 'Johnston 3D Organoid', 
-                 col = colorRamp2(c(0, 15), viridis(2)))
[…]

### 2019-02-15 manuscript.Rmd
updated data source

@@
 
-retina_plus_group <- tsne_50 %>% filter(X1 > 30, X1 < 50, X2 > -17, X2 < 16)
+retina_plus_group <- tsne_50 %>% filter(X1 > -5, X1 < 40, X2 > 30, X2 < 40)
 dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)
@@
   fontsize(size = 6) %>% 
-  width(width = c(0.5,0.5,0.5,3,1,0.5,0.5,0.5,0.5))
+  width(width = c(0.5,0.5,0.5,3,1,0.5,0.5,0.5,0.5,0.5))
 #Output Caption (* * for italics)

### 2019-02-15 src/data_pull.R

@@ -3,8 +3,10 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_02.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
+
+
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 load('data/human_tx_studies.Rdata')

### 2019-02-15 src/retina_development_heatmaps.R
heatmap dev

@@ -2,69 +2,105 @@ library(viridis)
 library(ComplexHeatmap)
 library(tidyverse)
 library(circlize)
+library(pool)
+library(RSQLite)
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','SLX2')
 pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
 
-gene <- progenitor
+gene <- pr
 query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
-  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as.tibble()) %>% 
-  as.tibble()
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+  as_tibble()
 
 ESC <- p %>% 
   filter(Tissue == 'ESC') %>% 
-  mutate(Days = 0) %>% group_by(ID, Days) %>% summarise(value = mean(value)) %>% mutate(Days = as.integer(Days))
+  mutate(Days = 0, Type = 'ESC') %>% 
+  group_by(ID, Days) %>% 
+  summarise(value = mean(value)) %>% 
+  mutate(Days = as.integer(Days))
 organoid_swaroop_GFP <- p %>% 
-  filter(sample_accession %in% (meta_ret %>% filter(Sub_Tissue == 'Retina - 3D Organoid Stem Cell') %>% pull(sample_accession))) %>%
-  filter(!grepl('GFP negative', sample_attribute), study_accession != 'SRP159246') %>% 
[…]

### 2019-02-17 manuscript.Rmd
data update

@@
 
-retina_plus_group <- tsne_50 %>% filter(X1 > -5, X1 < 40, X2 > 30, X2 < 40)
+retina_plus_group <- tsne_50 %>% filter(X1 > 20, X1 < 50, X2 < -15, X2 > -50)
+retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 35))
 dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)

### 2019-02-17 src/data_pull.R

@@ -3,14 +3,14 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2019/EiaD_human_expression_2019_04.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
 
 
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 load('data/human_tx_studies.Rdata')
-tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 50) %>% as_tibble()
+tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
 gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tx_anno <- gene_pool_2019 %>% tbl('tx_IDs') %>% as_tibble()
 limma_DE_tests <- gene_pool_2019 %>% tbl('limma_DE_tests') %>% as_tibble()

### 2019-02-17 src/euc_dist.R
quick workup to generate euclidean distance from a certain gene

@@ -1,0 +1,29 @@
+library(viridis)
+library(ComplexHeatmap)
+library(tidyverse)
+library(circlize)
+library(pool)
+library(RSQLite)
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2019/EiaD_human_expression_2019_04.sqlite')
+
+core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+
+gene <- all_genes
+fetal_ret <- core_tight_2019 %>% filter(Sub_Tissue == 'Retina - Fetal Tissue') %>% pull(sample_accession)
+query = paste0('select * from lsTPM_gene where sample_accession in ("',paste(fetal_ret, collapse='","'),'")')
+p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+  as_tibble()
+
+x <- p %>% 
+  group_by(ID, Age_Days) %>% 
+  summarise(value = mean(value)) %>% 
+  mutate(Days = as.integer(Age_Days)) %>% 
+  select(-Age_Days) %>% 
+  spread(ID, value) %>% t()
+colnames(x) <- x[1,]
+x <- x[-1,]
+
+euc_dist <- function(x1, x2) sqrt(sum((x1 - x2) ^ 2))
+e<-apply(x, 1, function(y) euc_dist(x['OTX2',],y))
+e %>% sort() %>% head()

### 2019-02-18 src/retina_development_heatmaps.R
fixed gene names, trying some modeling

@@ -4,12 +4,20 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
-rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','SLX2')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2019/EiaD_human_expression_2019_04.sqlite')
+rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
 pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
+cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
+cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
+rod <- c('NR2E3','NRL','MEF2C','ESRRB','CNGB1','GNAT1','GNGT1','GRK1','PDE6G','PDE6A','CNGA1','RHO','SAG','GNB1','PDE6B')
 
-gene <- pr
+
+all_markers <- c(rgc, pr, progenitor, cone_rod, cone, rod) %>% unique()
+
+core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+
+gene <- cone
 query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
[…]

### 2019-02-19 src/retina_development_heatmaps.R
thinking that merging dates into groups makes sense

@@ -144,7 +144,7 @@ Heatmap(log2(y), cluster_columns = T,
 
 
 # summarise by type -----
-sum_type <- function(gene_set, type){
+sum_type <- function(gene_set){
   query = paste0('select * from lsTPM_gene where ID in ("',paste(gene_set, collapse='","'),'")')
   p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
     left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
@@ -176,21 +176,66 @@ sum_type <- function(gene_set, type){
   out
 }
 
-z <- bind_rows(
-  sum_type(rgc),
-  sum_type(pr),
-  sum_type(progenitor),
-  sum_type(cone_rod),
-  sum_type(cone),
-  sum_type(rod)
-)
+z <- rbind(
+  sum_type(rgc)$data,
+  sum_type(pr)$data,
+  sum_type(progenitor)$data,
+  sum_type(cone_rod)$data,
+  sum_type(cone)$data,
+  sum_type(rod)$data
+) %>% as.matrix()
+type <- sum_type(rgc)$type
 row.names(z) <- c('RGC','PR','Progenitor','ConeRod','Cone','Rod')
-Heatmap(log2(z %>% as.matrix()) , cluster_columns = T, 
+
+ha_column = HeatmapAnnotation(df = data.frame(Type = type), 
+                              col = list(Type = c("ESC" = magma(10)[1],
+                                                  "Tissue" = magma(10)[3],
+                                                  "Eldred" = magma(10)[5],
+                                                  "Kaewkhaw GFP-" = magma(10)[7],
+                                                  "Kaewkhaw GFP+" = magma(10)[9])))
+Heatmap(log2(z %>% as.matrix()) , 
+        cluster_columns = F, 
         col = viridis(10),
+        top_annotation = ha_column,
         clustering_distance_rows = "pearson", 
         clustering_distance_columns = "euclidean", 
         show_heatmap_legend = F)
 
+
+  # let's try merging ages into groups -----
+ages <- colnames(z)
+zz <- z %>% 
+  t() %>% 
+  as_tibble() 
+zz$Age <- as.numeric(ages)
+zz <- zz %>% mutate(Range = case_when(Age == 0 ~ 0,
+                                Age < 40 ~ 35,
+                                Age < 55 ~ 50,
+                                Age < 60 ~ 60,
+                                Age < 73 ~ 70,
[…]

### 2019-02-19 src/retina_development_heatmaps.R
creating heatmap with merged time poitns

@@ -17,7 +17,7 @@ all_markers <- c(rgc, pr, progenitor, cone_rod, cone, rod) %>% unique()
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-gene <- cone
+gene <- all_genes
 query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
   left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
@@ -113,19 +113,39 @@ four <- Heatmap(log2(y), cluster_columns = F, column_title = 'Eldred 3D Organoid
 
 one + two + three + four
 
-################ 
-# all in one
-################
+# all in one ------
 tissue <- bind_rows(fetal_tissue %>% mutate(Type = 'Tissue'), 
                     adult_tissue %>% mutate(Type = 'Tissue'),
                     ESC %>% mutate(Type = 'ESC'),
                     organoid_johnston %>% mutate(Type = 'Eldred'),
                     organoid_swaroop_GFP %>% mutate(Type = 'Kaewkhaw GFP+'),
                     organoid_swaroop_GFPneg %>% mutate(Type = 'Kaewkhaw GFP-'))
-x <- tissue
-y <- x %>% spread(ID, value) %>% select(-Days, -Type) %>% t()
-type <- (x %>% spread(ID, value) %>% t())[2,]
+x <- tissue %>% mutate(Range = case_when(Days == 0 ~ 0,
+                                         Days <= 10 ~ 10,
+                                         Days <= 20 ~ 20,
+                                         Days < 40 ~ 35,
+                                         #Days < 55 ~ 50,
+                                         Days < 65 ~ 55,
+                                         #Days < 75 ~ 70,
+                                         Days < 85 ~ 75,
+                                         #Days < 95 ~ 90,
+                                         Days < 105 ~ 100,
+                                         #Days < 115 ~ 110,
+                                         Days < 125 ~ 120,
+                                         #Days < 135 ~ 130,
+                                         Days < 145 ~ 140,
+                                         #Days < 165 ~ 160,
+                                         Days < 185 ~ 180,
+                                         #Days < 205 ~ 200,
+                                         #Days < 225 ~ 220,
[…]

### 2019-02-19 src/euc_dist.R
finalize heatmap

@@ -8,7 +8,7 @@ gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-gene <- all_genes
+#gene <- all_genes
 fetal_ret <- core_tight_2019 %>% filter(Sub_Tissue == 'Retina - Fetal Tissue') %>% pull(sample_accession)
 query = paste0('select * from lsTPM_gene where sample_accession in ("',paste(fetal_ret, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
@@ -27,3 +27,4 @@ x <- x[-1,]
 euc_dist <- function(x1, x2) sqrt(sum((x1 - x2) ^ 2))
 e<-apply(x, 1, function(y) euc_dist(x['OTX2',],y))
 e %>% sort() %>% head()
+distances <- dist(x)

### 2019-02-19 src/retina_development_heatmaps.R

@@ -8,7 +8,7 @@ gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
 pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
-cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
+# cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
 cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
 rod <- c('NR2E3','NRL','MEF2C','ESRRB','CNGB1','GNAT1','GNGT1','GRK1','PDE6G','PDE6A','CNGA1','RHO','SAG','GNB1','PDE6B')
 
@@ -17,7 +17,7 @@ all_markers <- c(rgc, pr, progenitor, cone_rod, cone, rod) %>% unique()
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-gene <- all_genes
+gene <- all_markers
 query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
 p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
   left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
@@ -148,21 +148,34 @@ days[0] <- 'ESC'
 days[length(days)] <- 'Adult'
 colnames(y) <- days
 
[…]

### 2019-02-19 src/euc_dist.R
pare down euc_dist to just search for set list, makes funcs in retina_dev_heatmap

@@ -8,23 +8,48 @@ gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-#gene <- all_genes
+# gene list ------
+rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
+progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
+# cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
+cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
+rod <- c('NR2E3','NRL','MEF2C','ESRRB','CNGB1','GNAT1','GNGT1','GRK1','PDE6G','PDE6A','CNGA1','RHO','SAG','GNB1','PDE6B')
+pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW', cone, rod) %>% unique()
+all_markers <- c(rgc, pr, progenitor, cone, rod) %>% unique()
+
+
+# fetal retina
 fetal_ret <- core_tight_2019 %>% filter(Sub_Tissue == 'Retina - Fetal Tissue') %>% pull(sample_accession)
 query = paste0('select * from lsTPM_gene where sample_accession in ("',paste(fetal_ret, collapse='","'),'")')
-p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
-  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
-  as_tibble()
-
-x <- p %>% 
-  group_by(ID, Age_Days) %>% 
-  summarise(value = mean(value)) %>% 
[…]

### 2019-02-19 src/retina_development_heatmaps.R

@@ -4,20 +4,273 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2019/EiaD_human_expression_2019_04.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
-pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
 # cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
 cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
 rod <- c('NR2E3','NRL','MEF2C','ESRRB','CNGB1','GNAT1','GNGT1','GRK1','PDE6G','PDE6A','CNGA1','RHO','SAG','GNB1','PDE6B')
-
-
-all_markers <- c(rgc, pr, progenitor, cone_rod, cone, rod) %>% unique()
+pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW', cone, rod) %>% unique()
+all_markers <- c(rgc, pr, progenitor, cone, rod) %>% unique()
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-gene <- all_markers
+gene <- aman
+
[…]

### 2019-02-21 src/retina_development_heatmaps.R
fix plotting logic, new example for diff expression

@@ -4,7 +4,7 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
 
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
@@ -19,7 +19,7 @@ core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 gene <- aman
 
 # function to split by experiment / study / type -------
-plotter_split <- function(gene_vector, annotation = F) {
+plotter_split <- function(gene_vector, annotation = F, breaks = c(0,5,10,15)) {
   gene <- gene_vector
   query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
   p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
@@ -70,7 +70,7 @@ plotter_split <- function(gene_vector, annotation = F) {
   y <- y[-1,]
   
   one <- Heatmap(log2(y+1), cluster_columns = F,   column_title = 'Retina Tissue',
-                 col = viridis(10),
+                 col = colorRamp2(breaks = breaks, colors = viridis(length(breaks))),
                  show_row_names = FALSE,
                  name = 'log2(TPM+1)',
                  clustering_distance_rows = "pearson", 
@@ -83,7 +83,7 @@ plotter_split <- function(gene_vector, annotation = F) {
   y <- y[-1,]
   
   two <- Heatmap(log2(y), cluster_columns = F, column_title = 'Kaewkhaw\nGFP+ 3D\nRetina',
-                 col = viridis(10),
+                 col = colorRamp2(breaks = breaks, colors = viridis(length(breaks))),
                  clustering_distance_rows = "pearson", 
                  clustering_distance_columns = "euclidean", 
                  show_row_names = FALSE,
[…]

### 2019-02-21 src/euc_dist.R
adding big euc dist search for hpc

@@ -21,6 +21,10 @@ all_markers <- c(rgc, pr, progenitor, cone, rod) %>% unique()
 # fetal retina
 fetal_ret <- core_tight_2019 %>% filter(Sub_Tissue == 'Retina - Fetal Tissue') %>% pull(sample_accession)
 query = paste0('select * from lsTPM_gene where sample_accession in ("',paste(fetal_ret, collapse='","'),'")')
+p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+  as_tibble()
+
 x <- p %>%
   select(ID, sample_accession, value) %>% 
   unique() %>% 
@@ -53,3 +57,34 @@ for (i in all_markers){
 link <- cbind(all_markers, top[2:length(top)])
 colnames(link) <- c('Marker', 'TopRelated')
 save(link, file = 'data/markers_related_retina.Rdata')
+
+
+
+
+
+
+
+
+
+
+# ALL (for HPC)
+library(parallelDist)
+library(reshape)
+query = paste0('select * from lsTPM_gene')
+p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+  as_tibble()
+x <- p %>%
+  select(ID, sample_accession, value) %>% 
+  unique() %>% 
+  spread(ID, value) %>% 
+  select(-sample_accession) %>% 
+  t()
+sample_accession <- p %>%
+  select(ID, sample_accession, value) %>% 
+  unique() %>% 
+  spread(ID, value) %>% 
+  pull(sample_accession)
+colnames(x) <- sample_accession
+y <- parDist(x, method = 'euclidean', threads = 16) %>% as.matrix()
+euc_dist_all_by_all <- melt(y)

### 2019-02-21 manuscript.Rmd
improve coords

--- 2019-02-15 manuscript.Rmd
@@
 
-retina_plus_group <- tsne_50 %>% filter(X1 > -5, X1 < 40, X2 > 30, X2 < 40)
+retina_plus_group <- tsne_50 %>% filter(X1 > 17, X1 < 50, X2 < -15, X2 > -50)
+retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 35))
 dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)

### 2019-02-21 src/euc_dist.R
grouping

@@ -88,3 +88,4 @@ sample_accession <- p %>%
 colnames(x) <- sample_accession
 y <- parDist(x, method = 'euclidean', threads = 16) %>% as.matrix()
 euc_dist_all_by_all <- melt(y)
+euc_dist_all_by_all_top100 <- euc_dist_all_by_all %>% group_by(Var2) %>% arrange(value) %>% top_n(-100)

### 2019-02-21 manuscript.Rmd
text updates

@@
 ## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
-There are, currently, three approaches to studying developing human retina: 1.
-post-mortem fteal tissue, 2.
-organoids, and 3.
-stem cell-derived cells.
-As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level.
-Should organoid and stem cell-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
+There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids.
+We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.
 
-To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
-Here we see three distinct groupings: adult retina (2), developing retina (3), and stem cell/undifferentiated (1).
+To evaluate how the tissues and organoids compare at a transcriptome level, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the three types of retina tissue (adult, fetal, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
+Here we see three distinct groupings: adult retina (2), developing fetal retina and stem cell organoid (3), and undifferentiated and early differentiating stem cells (1).
 We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
-All of the stem cell-based retina samples are older than 50 days.
+All of the stem cell-based retina samples in cluster 2 are older than 50 days.
 
@@
 
-tsne_50_prep %>% 
+zoom_plot <- tsne_50_prep %>% 
   rowwise() %>% 
@@
 ```
+
+Furthermore we also
 

### 2019-02-21 manuscript.Rmd
move figure creation to separate script

@@
 
-retina_plus_group <- tsne_50 %>% filter(X1 > 17, X1 < 50, X2 < -15, X2 > -50)
-retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 35))
-dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)
-
-retina_plus_group$Cluster <- dbscan_cluster$cluster
-col1 = 'X1'
-col2 = 'X2'
-# create label points for each cluster
-cluster_centers <- retina_plus_group %>%
-  left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept")) %>% group_by(Cluster) %>%
-  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Sub_Tissue),collapse=','))
-
-# samples closest to center in each cluster
-center_samples <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
-  left_join(.,cluster_centers, by=c('Cluster')) %>%
-  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
-  group_by(Cluster) %>%
-  dplyr::slice(which.min(Distance)) %>%
-  dplyr::slice(1) %>%
-  .[['sample_accession']]
-
-# cluster stats
-cluster_stats <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
-  mutate(Cluster = as.factor(Cluster)) %>%
-  group_by(Cluster) %>%
-  summarise(Cluster_Tissues = paste(unique(Sub_Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep='')) %>% 
-  mutate(Cluster_Tissues = paste0(Cluster, ': ', Cluster_Tissues))
-
-# set up for ggplot
-retina_plus_group_prep <- retina_plus_group %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>%
-  mutate(Cluster = as.factor(Cluster)) %>%
-  select(-Cluster_Tissues) %>%
-  left_join(., cluster_stats, by=c('Cluster')) %>%
-  mutate(Label = Cluster) %>%
-  mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
-
-tsne_zoom <- retina_plus_group_prep  %>%
-  mutate(Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
-                           TRUE ~ Tissue),
-         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
-         Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>%
-  mutate( zoom = TRUE )
-
-# https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html
-StatChull <- ggproto("StatChull", Stat,
-                     compute_group = function(data, scales) {
-                       data[chull(data$x, data$y), , drop = FALSE]
-                     },
-
-                     required_aes = c("x", "y")
-)
-stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
-                       position = "identity", na.rm = FALSE, show.legend = NA, 
-                       inherit.aes = TRUE, ...) {
-  layer(
-    stat = StatChull, data = data, mapping = mapping, geom = geom, 
-    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
-    params = list(na.rm = na.rm, ...)
-  )
-}
-
-zoom_plot <- tsne_50_prep %>% 
-  rowwise() %>% 
-  mutate(Age = suppressWarnings(as.integer(str_split(sample_attribute, '_')[[1]][2]))) %>% 
-  mutate(Origin = as.character(Origin)) %>% 
-  mutate(Origin = case_when(Age < 30 ~ 'Organoid < 30 days',
-                            Age > 30 ~ 'Organoid > 30 days',
-                            TRUE ~ Origin)) %>% 
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
-         Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
-                            TRUE ~ Tissue)) %>% 
-  mutate(ZoomTissue = case_when((Tissue == 'Retina' & Sub_Tissue != 'Retina - Adult Tissue') | Tissue == 'ESC' ~ 'Early Retina',
-                                TRUE ~ Sub_Tissue),
-         Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
-                           TRUE ~ Tissue),
-         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)) %>% 
-  mutate(Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>% 
-  ggplot(aes(x=X1,y=X2)) +
-  scale_shape_manual(values=c(0:20,35:50)) +
-  #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
-  geom_point(size=4, alpha=1, aes(shape=Origin, colour = Group)) +
-  xlab('t-SNE 1') + ylab('t-SNE 2') +
-  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
-  geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
-  stat_chull(data=tsne_zoom %>% 
-               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
-  scale_color_discrete_sequential(palette = 'viridis') + 
-  scale_fill_viridis_d(option = 'plasma') +
-  theme_bw() +
-  guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
-         shape = guide_legend(ncol =2 )) +
-  theme(axis.text.x = element_text(size = 6),
-        axis.text.y = element_text(size = 6),
-        text = element_text(family = 'Linux Libertine O', size = 6))
-
+knitr::include_graphics('figures_and_tables/zoom_heatmap_retina.png')
 ```

### 2019-02-21 figures_and_tables/zoom_heatmap_figure.R

@@ -1,0 +1,110 @@
+load('data/data_pull.Rdata')
+library(tidyverse)
+library(ggforce)
+library(dbscan)
+library(cowplot)
+
+retina_plus_group <- tsne_50 %>% filter(X1 > 17, X1 < 50, X2 < -15, X2 > -50)
+retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 35))
+dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 2)
+
+retina_plus_group$Cluster <- dbscan_cluster$cluster
+col1 = 'X1'
+col2 = 'X2'
+# create label points for each cluster
+cluster_centers <- retina_plus_group %>%
+  left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept")) %>% group_by(Cluster) %>%
+  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Sub_Tissue),collapse=','))
+
+# samples closest to center in each cluster
+center_samples <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
+  left_join(.,cluster_centers, by=c('Cluster')) %>%
+  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
+  group_by(Cluster) %>%
+  dplyr::slice(which.min(Distance)) %>%
+  dplyr::slice(1) %>%
+  .[['sample_accession']]
+
+# cluster stats
+cluster_stats <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("sample_accession", "study_accession", "study_title", "study_abstract", "sample_attribute", "Tissue", "Sub_Tissue", "Origin", "Kept"))  %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  group_by(Cluster) %>%
+  summarise(Cluster_Tissues = paste(unique(Sub_Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep='')) %>% 
+  mutate(Cluster_Tissues = paste0(Cluster, ': ', Cluster_Tissues))
+
+# set up for ggplot
[…]

### 2019-02-21 src/retina_development_heatmaps.R

@@ -251,6 +251,11 @@ plotter_merge <- function(gene_vector, annotation = F, link = NA, breaks = c(0,5
 # all makers 
 marker_split_plot <- plotter_split(all_markers, annotation = T)
 draw(marker_split_plot, padding = unit(c(15,15,15,15),"mm"))
+pdf("figures_and_tables/heatmap_retina_time_series.pdf", width = 12, height = 8)
+draw(marker_split_plot, padding = unit(c(15,15,15,15),"mm"))
+dev.off()
+
+
 marker_merge_plot <- plotter_merge(all_markers, annotation = T)
 draw(marker_split_plot +marker_merge_plot, padding = unit(c(15,15,15,15),"mm"))
 

### 2019-02-21 manuscript.Rmd
figure updates, some text updates

@@
 tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type from 2017 to 2019 EiaD.
-We also improved the metadata labelling for the cornea samples, now delineating endothelial and epithelial tissues.')
+We also improved the metadata labelling, the cornea samples now delineats endothelial and epithelial tissues and the retina samples distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.')
 
@@
   ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
-  theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
+  theme(text = element_text(size = 10)) +
   scale_fill_discrete_sequential(palette = 'viridis')  
@@
   theme_bw() +
-  theme(text = element_text(family = 'Linux Libertine O', size = 30)) +
+  theme(text = element_text(size = 30)) +
   scale_color_discrete_sequential(palette = 'viridis') + 
@@
 
-To evaluate how the tissues and organoids compare at a transcriptome level, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the three types of retina tissue (adult, fetal, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`).
-Here we see three distinct groupings: adult retina (2), developing fetal retina and stem cell organoid (3), and undifferentiated and early differentiating stem cells (1).
-We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
-All of the stem cell-based retina samples in cluster 2 are older than 50 days.
+To evaluate how the tissues and organoids compare at a transcriptome level, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the three types of retina tissue (adult, fetal, and organoid) (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'A')`).
+Here we see three distinct groupings: adult retina (1), developing fetal retina and stem cell organoid (2), and undifferentiated and early differentiating stem cells (3).
+We do see several organoid samples in cluster 3, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
+All of the organoid retina samples in cluster 2 are older than 50 days.
+
+To assess how similarly the fetal and organoid retina develop through time, we plot expression of retinal progenitors, photoreceptors, and retinan ganglion markers by time in days (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`).
+Each row is a gene marker of either progenitor, photoreceptor, or retinal ganglion cells.
+The rows are hierarchially clustered to put more similar expression patterns closer together, as denoted by the height of the dendrogram.
+We split the organoid tissues into three groups: Kaewkhaw et al.
+GFP+ and GFP- samples, and Eldred et al.
+samples [@eldredThyroidHormoneSignaling2018; @kaewkhawTranscriptomeDynamicsDeveloping2015].
+The Kaewkhaw samples are flow sorted for a GFP marker under the control of a *CRX* promoter, an important of photoreceptor development.
+GFP+ cells thus are believed to be enriched in photoreceptor populations.
+We see that the progenitor, photoreceptor, and retinal ganglion groups are largely clustered together, with patterns of expression consistent across the retinal and organoid groups.
 
@@
 tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, stem cell retina, and fetal retina tissue have highly similar transcriptomes.
-The zoom inset shows the retina samples.
+The zoom inset (A) shows the retina samples.
 The "Sub-Tissue Cluster" shading shows the cluster membership of the three major groups.
-The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.')
+The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.
+Major markers of retina progenitor, photoreceptors (cone and rod), and retinal ganglia cells (RGC) have simliar gene expression patterns across development in retina fetal tissue and organoids.')
 

### 2019-02-21 figures_and_tables/zoom_heatmap_figure.R

@@ -3,6 +3,18 @@ library(tidyverse)
 library(ggforce)
 library(dbscan)
 library(cowplot)
+library(ggrepel)
+library(colorspace)
+
+# set up for ggplot
+tsne_50_prep <- tsne_50 %>% select(-Origin) %>% 
+  left_join(., core_tight_2019 %>% select(sample_accession, Origin), by = 'sample_accession') %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  select(-Cluster_Tissues) %>% 
+  left_join(., cluster_stats, by=c('Cluster')) %>%
+  mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
+  mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
 
 retina_plus_group <- tsne_50 %>% filter(X1 > 17, X1 < 50, X2 < -15, X2 > -50)
 retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 35))
@@ -90,7 +102,7 @@ zoom_plot <- tsne_50_prep %>%
   #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
   geom_point(size=2, alpha=1, aes(shape=Origin, colour = Group)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
-  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = T, zoom.data=zoom) + 
+  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
   geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
   stat_chull(data=tsne_zoom %>% 
                mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
@@ -101,10 +113,14 @@ zoom_plot <- tsne_50_prep %>%
          shape = guide_legend(ncol =2 )) +
   theme(axis.text.x = element_text(size = 6),
         axis.text.y = element_text(size = 6),
-        text = element_text(family = 'Linux Libertine O', size = 6))
-
[…]

### 2019-02-21 src/retina_development_heatmaps.R

@@ -4,7 +4,7 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
 
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
@@ -68,11 +68,20 @@ plotter_split <- function(gene_vector, annotation = F, breaks = c(0,5,10,15)) {
   colnames(y) <- y['Days',]
   colnames(y)[ncol(y)] <- 'Adult'
   y <- y[-1,]
-  
-  one <- Heatmap(log2(y+1), cluster_columns = F,   column_title = 'Retina Tissue',
+  # ht_global_opt(heatmap_legend_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               annotation_legend_labels_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_row_names_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_column_names_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_row_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_column_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_legend_labels_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
+  #               heatmap_legend_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'))
+                
+  one <- Heatmap(log2(y+1), cluster_columns = F,   
+                 column_title =  'Retina Tissue', 
                  col = colorRamp2(breaks = breaks, colors = viridis(length(breaks))),
                  show_row_names = FALSE,
[…]

### 2019-02-22 manuscript.Rmd
text updates

@@
 
-Furthermore we also
+# Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
+
+While organoid and retina tissues are highly similar at a pan-transcriptome level, differential expression analysis between organoid and fetal retina reveals differences in related groups of genes.
 

### 2019-02-23 manuscript.Rmd
terxt updates

@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To provide the eye community with a high quality RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
+abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
 METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across over 44 tissues in the GTEx project as a non-eye reference set.
@@
 The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote.
-The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and likely mislabeling, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
-RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.00, https://eyeIntegration.nei.nih.gov) to display this data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues.
-In a powerful demonstration of the value of this unified pan-eye dataset, we agnostically show that, at a transcriptome level, fetal, stem cell, and organoid retina tissues are highly similar.
-CONCLUSION: The eyeIntegration v1.00 web app serves the pan-eye and body transcriptome dataset, EiaD.
+The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.02 as of Februay 24 2019, https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues.
+As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest targets for organoid process improvements.
+CONCLUSION: The eyeIntegration v1.02 web app serves the pan-eye and body transcriptome dataset, EiaD.
 This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
@@
 From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid.
-The differentiation, development, and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, also known as the transcriptome.
+The differentiation, maturation, and function of these tissues is mediated through spatial and temporal specific transcript and gene expression patterns, also known as the transcriptome.
 Today, RNA-sequencing (RNA-seq) is the predominant technology for quantifying the transcriptome.
-Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
-To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
+Analysis of the transcript expression across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
+To this end, researchers have used a wide variety of human tissue sources to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014b; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014b; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015a; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
 
-## GTEx is a tremendously useful gene expression resource that, unfortunately, lacks eye tissues 
-The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presented the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
-Unfortunately, no eye tissues are included.
-The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified huge portions of the RNA-seq data, including some human eye tissues, from the (SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
-However no curation of the sample level metadata was done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed.
+## The GTEx gene expression resource lacks eye-specific tissues
+The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
+However, no eye tissues are included in their resource.
+The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
+To date, no curation of the sample level metadata has been done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed.
 More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016].
@@
 
-## eyeIntegration app reactively serves huge GTEx and human eye tissue datasets (EiaD)
-The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provided the largest set of hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
-This responsive web app allowed for quick comparisons across many eye tissues and dozens of other body tissues.
+**ADD MENTION OF NEW SWAROOP DATA "GREx"**
+
+## eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
+The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
+This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues.
 The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples.
-The interactive nature of some of the steps precludes efficient and regular updates for the data.
+The interactive nature of some of the steps precludes efficient and regular data updates for the data.
 
-In order to better serve the eye research community with regular, versioned updates, we have re-written the bioinformatic pipeline that creates the eye and body RNA-seq dataset for eyeIntegration.
-We call this reproducible and versioned transcriptome dataset Eye in a Disk (EiaD).
-The pipeline nearly fully automates the EiaD creation, ensure full reproducbility of the results, allow for external data comparison, providse consistent sample quality control, and improves efficiency for future sample updates.
-The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples.
+To better meet the needs of the eye research community we have re-written the bioinformatic pipeline that creates the eye and body RNA-seq dataset to allow for regular, versioned updates for eyeIntegration.
+We call this reproducible and versioned transcriptome dataset "Eye in a Disk" (EiaD).
+The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provide consistent sample quality control, and improves efficiency for future sample updates.
+The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling.
 The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands.
@@
 
+## EiaD dataset and eyeIntegration used to identify potential avenues to improve retina organoid maturation
+
+Retina organoids
+
 # Methods
-## Identifying potential eye samples with exhaustive search of SRA
-We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
-We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, stem cell derived tissue, organoids, and immortalized cell lines.
+## Identification of potential eye samples 
+We exhaustively searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
+We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, induced pluripotent stem cell (iPSC)-derived tissue, stem cell derived organoids, and immortalized cell lines.
 In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
@@
 Any of the preceding information missing from the SRA metadata was added by hand, when available.
-Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.
+Stem cell-derived tissues and cell lines are marked as sub-tissues of the tissue they model.
 
@@
 Samples were re-quantified against a transcriptomic index built on the filtered transcript sequences.
-The Salmon count values were were quantified as (transcript) length scaled Transcripts Per Million (TPM) to the transcript and gene level using tximport [@sonesonDifferentialAnalysesRNAseq2016a].
+The Salmon count values were quantified as (transcript) length scaled Transcripts Per Million (TPM) to the transcript and gene level using tximport [@sonesonDifferentialAnalysesRNAseq2016a].
 
 ## Quality control 
-We began the quality control procedure by removing samples with a Salmon calculated mapping rate less than 50%.
+We first removed  samples with a Salmon calculated mapping rate less than 40%.
 We further filtered samples by calculating the median gene expression for each sample and only kept samples that had a median count >2 across all genes.
@@
 We then calculated the center of the resulting sub-tissue clusters by finding the average two-dimensional coordinates of all samples of a given sub-tissue type (p1).
-We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type by measuring the euclidean distance of each sample (p2) from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance with the equation $dist = \sqrt\sum(p1-p2)^{2}$.
+We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type, by measuring the euclidean distance of each sample (p2) from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance with the equation $dist = \sqrt\sum(p1-p2)^{2}$.
 
 ## Differential Gene Expression Analysis and GO term enrichment
-We used the cleaned data to determine differential gene expression between different sub-tissue types.
+We used the processed data to determine differential gene expression between different sub-tissue types.
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the Limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
+We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchielimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
@@
 ## Manuscript as code
-This manuscript's figures (excepting (Figure 2), tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
+This manuscript's figures (excepting (Figure 2)), tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
 The knitr and pandoc program is used to lay out the figures and tables and output a docx file.

### 2019-02-25 manuscript.Rmd
text updates

@@
 
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and many sub-tissue types
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(Sub_Tissue)) %>% unique() %>% length()) - 1` eye sub-tissue types
 
@@
   mutate(Dataset = Version) %>% 
-  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') + scale_alpha_manual(values = c(0.5,1)) +
+  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') +
+  ylab('Count of Samples') + scale_alpha_manual(values = c(0.5,1)) +
   theme(text = element_text(size = 10)) +
@@
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
-The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison (`r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`).
-The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
+The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissue samples and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison (`r tab_cap(name = 'tableSampleCounts', display = 'cite')`, `r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`).
+The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'ESC') %>% nrow()` undifferentiated iPSC, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
 
-Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
-We have also substantially improved the granualarity of the cornea tissue metadata, now delinerating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
+Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delinerating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
+Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC).
 Another substantial addition to the 2019 EiaD are non-protein coding genes.
-We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.
+We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).
 
@@
 We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r fig_cap(name='workflowDoodle', display = 'cite')`, `r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
-In order to create a high quality final dataset across the `r core_tight_2019 %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
-To identify tube mislabelling or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
+To create a high quality final dataset across the `r core_tight_2019 %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`)  and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+To identify tube mislabeling or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
 
 After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')`).
-For example, in retina derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina tissue cluster together.
+For example, in retina-derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina tissue cluster together.
 Our ability to uniformly cluster data by known biological function independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.
@@
 ```{r tsne, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
-tsne_cap <- fig_cap(name='tSNE', caption = 't-SNE tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing.
-Colors match different tissue sets and shapes of points define the origin of the tissues.')
+tsne_cap <- fig_cap(name='tSNE', caption = 't-SNE two-dimensional transcriptome profiles by sample demonstrate effective quality control and transcriptome processing.
+Colors match different tissue types and shapes of points define the origin of the tissues.')
 
@@
 The EiaD 2019 dataset is used directly by the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
-The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information.
+The web app was designed to provide a simple interface that has the same general concept – select specific genes and tissue and view relevant information.
 The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.
@@
 ## Custom gene and tissue expression boxplots
-The expression tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
+The 'Expression' tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
 The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
@@
 
+```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap, fig.pos = 'H'}
+tableSampleCountsGTEx_cap <- supTab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
+# Output <div>
+cat("<div id=\"tbl:tableSampleCountsGTEx\">")
+regulartable(gtex) %>% 
+  width(width = c(0.5,1,0.5,3)) %>% 
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",tableSampleCountsGTEx_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r tx_and_gene_types, echo = F, results='asis', warning = F, fig.cap =  tx_and_gene_types_cap, fig.pos = 'H'}
+tx_and_gene_types_cap <- supTab_cap(name='tx_and_gene_types', caption = 'Dozens of different types of gene and transcript types quantified')
+# Output <div>
+cat("<div id=\"tbl:tx_and_gene_types\">")
+table <- bind_rows(
+  tx_anno %>% group_by(transcript_type) %>% summarise(Count=n()) %>% mutate(Type = 'Transcript') %>% rename(Class = transcript_type), 
+  gene_anno %>% group_by(gene_type) %>% summarise(Count=n()) %>% mutate(Type = 'Gene') %>% rename(Class = gene_type))
+regulartable(table %>% group_by(Type) %>% arrange(-Count)) %>% 
+  width(width = c(2,1,2)) %>% 
+  flextable::align(align='right') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",tx_and_gene_types_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
 ```{r metadataTable, echo = F,  results='asis', fig.pos = 'H', fig.cap = metadata_cap}
@@
 cat(paste0("\n*",metadata_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap, fig.pos = 'H'}
-tableSampleCountsGTEx_cap <- supTab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
-# Output <div>
-cat("<div id=\"tbl:tableSampleCountsGTEx\">")
-regulartable(gtex) %>% 
-  width(width = c(0.5,1,0.5,3)) %>% 
-  flextable::align(align='center') %>% 
-  font(fontname = 'Linux Libertine O', part = 'all')
-#Output Caption (* * for italics)
-cat(paste0("\n*",tableSampleCountsGTEx_cap,"*\n"))
 # Output </div>
@@
 ```
+
 # Acknowledgements

### 2019-02-25 manuscript.Rmd
finished incorporating TC edits

@@
 ## Custom gene and tissue expression boxplots
-The 'Expression' tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
+The 'Expression' tab of the webpage provides a wealth of information about both gene- and transcript-level expression for eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
 The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
-A boxplot is then generated afer hitting the "Re(Draw) Plot" button with overlaid individual data points.
+A boxplot is then generated after hitting the "Re(Draw) Plot" button with overlaid individual data points.
 On mouse-over, the metadata for the individual sample is displayed.
@@
 ```{r appScreenshots, echo = F, fig.cap = appScreenshots_cap, fig.width=12}
-appScreenshots_cap <- fig_cap(name='appScreenshots', caption = 'Various screenshots from eyeIntegration web app.
+appScreenshots_cap <- fig_cap(name='appScreenshots', caption = 'Screenshots from eyeIntegration web app.
 A.
@@
 B.
-Alternative heatmap visualization.
+Heatmap visualization.
 C.
@@
 D.
-t-SNE visualization of individual cell gene expression profiles from Macosko et al.
+t-SNE visualization of gene expression profiles of individual cells from Macosko et al.
 E.
@@
 F.
-Differential gene expression across different tissues and with GO term enrichment.
-')
+Differential gene expression across different tissues and with GO term enrichment.')
 
@@
 ## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
-Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
+We performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing quick identification of eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
 We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues.
-The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
-As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons.
-The user additionally has the option to select different gene classes to examine, protein coding, lincRNA etc.
+The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
+As with 'Expression', the user can select which version of the web app to draw data from as well as select for gene- or transcript-level comparisons.
+The user additionally has the option to select different gene classes to examine, e.g.
+protein coding, lincRNA.
 
@@
 The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format.
-Differential expression can be visualized through fold change bar graphs in the ‘pan-tissue plots’ section under expression.
+Differential expression can be visualized through fold change bar graphs with the ‘Pan-tissue plots’ selection under ‘Expression’.
 Additionally, we performed GO enrichment for all differential comparisons.
@@
 ## Murine scRNA-seq enables testing of retina cell type specific expression
-We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
+We incorporated scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
 This allows researchers to quickly examine gene expression across individual cell types in the retina.
-Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'C')`).
+Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points, from embryonic day (E)11 to postnatal day (P)14 (when available), and a table of expression values is generated containing the expression data used to draw the heatmap (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'C')`).
 We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'D, see Methods')`).
-The plots show all cell types present a given developmental stage, and highlights cells expressing a gene above a user given level.
+The plots show all cell types present at a given developmental stage, and highlights cells expressing a gene above a user-selected given level.
 
-## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
+## EiaD 2019 demonstrates that iPSC-derived organoids and fetal retina have closely related transcriptomes
 There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids.
@@
 All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’).
-Code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+The code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
 
@@
 
-EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project.
-It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov.
+EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a human body tissue comparison set from the GTEx project.
+It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and it is served via a web app, eyeIntegration at https://eyeIntegration.nei.nih.gov.
 The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues.
-Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression.
-Several human and non-human primate studies have been posted in the past year on the bioRxiv pre-print server.
-As the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
+Furthermore, murine scRNA-seq datasets have been incorporated, allowing for examination of retina cell type-specific gene expression.
+Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
 
-As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for stem cell and organoid based developmental methods to be developed.
-Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level.
-This result underscores the value of these approaches in building models for human retina developmental .
+As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed.
+Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level.
+This result underscores the value of these approaches in building models for human retina development.
 
@@
 
-pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples demonstrates non-eye tissue most similar to adult retina is the brain, pituitary most similar to adult RPE, and skin and blood similar to cornea.')
+pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples suggests that the non-eye tissue most similar to adult retina is the brain, pituitary is the most similar to adult RPE, and skin and blood are similar to cornea.')
 

### 2019-02-25 src/age_prediction.R
fetal organoid retina prediction

@@ -1,0 +1,79 @@
+
+library(tidyverse)
+
+library(pool)
+library(RSQLite)
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+
+
+###################
+
+# build model to predict fetal tissue age
+# apply to organoid 
+# works OK on fetal tissue, but doesn't 
+# work well with organoid
+
+###################
+tissue <- bind_rows(ESC, fetal_tissue, adult_tissue)
+
+query = 'select * from lsTPM_gene'
+p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+  as_tibble()
+fetalTissue_allGene <- p %>% 
+  filter(gene_type == 'protein_coding') %>% 
+  filter(Sub_Tissue == 'Retina - Fetal Tissue' | study_accession == 'SRP159246') %>% 
+  select(ID, sample_accession, Age_Days, value) %>% 
+  mutate(age_accession = paste0(Age_Days, sep='_', sample_accession)) %>% 
+  select(-sample_accession, -Age_Days) %>% 
+  unique()
+
+# set up train test
+age_access <- fetalTissue_allGene$age_accession %>% unique() %>% sort()
+train_samp <- age_access[seq(1, length(age_access), 2)]
+train <- fetalTissue_allGene %>% filter(age_accession %in% train_samp)
+test <- fetalTissue_allGene %>% filter(!age_accession %in% train_samp)
+
+# mean summarise multiple overlappping time points
+train$age <- as.numeric(gsub('_.*','', train$age_accession))
+train <- train %>% group_by(ID, age) %>% summarise(value = mean(value))
+
+trainer <- train %>% spread(ID, value) 
+tester <- test %>% spread(ID, value)
+
+# caret
+library(caret)
+library(xgboost)
+set.seed(96)
+
+# rf_fit <- train(x = trainer %>% select(-age) %>% as.matrix(), 
+#                 y = trainer$age,
+#                 method = "rf")
+xgb <- xgboost(data = trainer %>% select(-age) %>% as.matrix(), 
+               label = trainer$age,
+               nrounds = 50)
+
+# predict on test
+## rf
[…]

### 2019-02-25 manuscript.Rmd
starting to write last sections

@@
 
-# Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
+## Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
 
 While organoid and retina tissues are highly similar at a pan-transcriptome level, differential expression analysis between organoid and fetal retina reveals differences in related groups of genes.
+Analysis of broad trends in GO term enrichment between differentially expressed genes shows that the HOXB family and protocadherins are differentially expressed between fetal retina and organoids.
+We noticed that several BRN3/POU4F genes were also differentially regulated
 
@@
 
+```{r GOterm_fetal_retina_organoid, echo = F, results='asis', warning = F, fig.cap = GOterm_cap, fig.pos = 'H'}
+GOterm_cap <- supTab_cap(name='GOterm_fetal_retina_organoid', caption = 'Top GO terms enriched between fetal retina and organoid retina, after paring of redundant terms with REVIGO')
+# Output <div>
+cat("<div id=\"tbl:GOterm_fetal_retina_organoid\">")
+gene_pool_2019 %>% tbl('all_vs_all_GO') %>% filter(Set == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell', ONTOLOGY == 'BP') %>% as_tibble() %>% mutate(qvalue = as.numeric(qvalue)) %>% arrange(qvalue) %>% filter(ID %in% c('GO:0001503','GO:0031589','GO:0050878','GO:0061448','GO:0050678','GO:0030335','GO:0030198', 'GO:0007156', 'GO:0050769', 'GO:0030099')) %>% mutate(geneID = gsub('<br>', ' ', geneID)) %>% as_tibble() %>% regulartable() %>%
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",GOterm_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
 # Acknowledgements

### 2019-02-25 src/retina_development_heatmaps.R

@@ -14,6 +14,7 @@ rod <- c('NR2E3','NRL','MEF2C','ESRRB','CNGB1','GNAT1','GNGT1','GRK1','PDE6G','P
 pr <- c('OTX2','RCVRN','AIPL1','NRL','CRX','PDE6B','NR2E3','ROM1','GNGT2','GNAT1','PDE6H','CNGB1','OPN1SW','GUCA1A','GNAT2','CNGA1','RHO','OPN1MW', cone, rod) %>% unique()
 all_markers <- c(rgc, pr, progenitor, cone, rod) %>% unique()
 
+
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
 gene <- aman
@@ -68,15 +69,8 @@ plotter_split <- function(gene_vector, annotation = F, breaks = c(0,5,10,15)) {
   colnames(y) <- y['Days',]
   colnames(y)[ncol(y)] <- 'Adult'
   y <- y[-1,]
-  # ht_global_opt(heatmap_legend_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               annotation_legend_labels_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_row_names_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_column_names_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_row_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_column_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_legend_labels_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'), 
-  #               heatmap_legend_title_gp = gpar(fontsize = 16, fontfamily = 'Times New Roman'))
-                
+
+  
   one <- Heatmap(log2(y+1), cluster_columns = F,   
                  column_title =  'Retina Tissue', 
                  col = colorRamp2(breaks = breaks, colors = viridis(length(breaks))),
@@ -284,83 +278,18 @@ related_split_plot + related_merge_plot
[…]

### 2019-03-01 manuscript.Rmd
figure and text updates

--- 2019-02-25 manuscript.Rmd
@@
 cluster_stats <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
+  mutate(Origin = gsub('Organoid','Stem Cell', Origin)) %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
   left_join(., core_tight_2019 %>% select(sample_accession, Origin), by = 'sample_accession') %>% 
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
@@
 
-# Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
+## Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
 
 While organoid and retina tissues are highly similar at a pan-transcriptome level, differential expression analysis between organoid and fetal retina reveals differences in related groups of genes.
+We  Analysis of broad trends in GO term enrichment between differentially expressed genes shows that the HOXB family and protocadherins are differentially expressed between fetal retina and organoids.
+We noticed that several BRN3/POU4F genes were also differentially regulated
 
@@
 
+```{r GOterm_fetal_retina_organoid, echo = F, results='asis', warning = F, fig.cap = GOterm_cap, fig.pos = 'H'}
+GOterm_cap <- supTab_cap(name='GOterm_fetal_retina_organoid', caption = 'Top GO terms enriched between fetal retina and organoid retina, after paring of redundant terms with REVIGO')
+# Output <div>
+cat("<div id=\"tbl:GOterm_fetal_retina_organoid\">")
+organoid_fetal_GO %>% filter(ID %in% c('GO:0001503','GO:0031589','GO:0050878','GO:0061448','GO:0050678','GO:0030335','GO:0030198', 'GO:0007156', 'GO:0050769', 'GO:0030099')) %>% mutate(geneID = gsub('<br>', ' ', geneID)) %>% as_tibble() %>% regulartable() %>%
+  flextable::align(align='center') %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",GOterm_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
 # Acknowledgements

### 2019-03-01 figures_and_tables/zoom_heatmap_figure.R

@@ -9,6 +9,8 @@ library(colorspace)
 # set up for ggplot
 tsne_50_prep <- tsne_50 %>% select(-Origin) %>% 
   left_join(., core_tight_2019 %>% select(sample_accession, Origin), by = 'sample_accession') %>% 
+  mutate(Origin = case_when(Sub_Tissue == 'Retina - 3D Organoid Stem Cell' ~ 'Organoid',
+         TRUE ~ Origin)) %>% 
   mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
   select(-Cluster_Tissues) %>% 
@@ -46,7 +48,7 @@ cluster_stats <- retina_plus_group %>% left_join(.,core_tight_2019, by = c("samp
 
 # set up for ggplot
 retina_plus_group_prep <- retina_plus_group %>%
-  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days'))) %>%
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid'))) %>%
   mutate(Cluster = as.factor(Cluster)) %>%
   select(-Cluster_Tissues) %>%
   left_join(., cluster_stats, by=c('Cluster')) %>%
@@ -83,7 +85,7 @@ stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
 
 zoom_plot <- tsne_50_prep %>% 
   rowwise() %>% 
-  mutate(Age = suppressWarnings(as.integer(str_split(sample_attribute, '_')[[1]][2]))) %>% 
+  mutate(Age = as.numeric(Age_Days)) %>% 
   mutate(Origin = as.character(Origin)) %>% 
   mutate(Origin = case_when(Age < 30 ~ 'Organoid < 30 days',
                             Age > 30 ~ 'Organoid > 30 days',
@@ -118,7 +120,7 @@ zoom_plot <- tsne_50_prep %>%
 svg('figures_and_tables/zoom_plot.svg', width = 4, height = 4)
 zoom_plot
 dev.off()
[…]

### 2019-03-01 src/retina_development_heatmaps.R

@@ -6,7 +6,7 @@ library(pool)
 library(RSQLite)
 gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
 
-rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2')
+rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2', 'POU4F2', 'POU4F3', 'NEFL', 'GAP43', 'SNCG')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
 # cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
 cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
@@ -43,13 +43,13 @@ plotter_split <- function(gene_vector, annotation = F, breaks = c(0,5,10,15)) {
     filter(Sub_Tissue == 'Retina - 3D Organoid Stem Cell', grepl('GFP negative', sample_attribute), study_accession != 'SRP159246') %>% 
     group_by(ID, Age_Days) %>% 
     summarise(value = mean(value)) %>% 
-    mutate(Days = as.integer(Age_Days), Type = 'Kaewkhaw GFP- 3D Retina')%>% 
+    mutate(Days = as.integer(Age_Days), Type = 'GFP- 3D Retina (Kaewkhaw)')%>% 
     select(-Age_Days)
   organoid_johnston <-  p %>% 
     filter(study_accession == 'SRP159246') %>% 
     group_by(ID, Age_Days) %>% 
     summarise(value = mean(value)) %>% 
-    mutate(Days = as.integer(Age_Days), Type = 'Kaewkhaw GFP+ 3D Retina') %>% 
[…]

### 2019-03-01 manuscript.Rmd
updating last section

@@
 
-While organoid and retina tissues are highly similar at a pan-transcriptome level, differential expression analysis between organoid and fetal retina reveals differences in related groups of genes.
-We  Analysis of broad trends in GO term enrichment between differentially expressed genes shows that the HOXB family and protocadherins are differentially expressed between fetal retina and organoids.
+To identify specific changes between organoid and retina tissue, we did differential gene expression and GO term enrichment analyses.
+GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets.
+We also noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`.
+
+We plotted these sets of genes 
+Analysis of broad trends in GO term enrichment between differentially expressed genes shows that the HOXB family and protocadherins are differentially expressed between fetal retina and organoids.
 We noticed that several BRN3/POU4F genes were also differentially regulated

### 2019-03-01 src/retina_development_heatmaps.R

@@ -275,15 +275,14 @@ related_split_plot + related_merge_plot
 
 
 # diff gene set -----
-plotter_split(gene_pool_2019 %>% tbl('limma_DE_gene') %>% filter(Comparison == 'Retina_3D.Organoid.Stem.Cell-Retina_Fetal.Tissue') %>% left_join(gene_pool_2019 %>% tbl('gene_IDs')) %>% filter(gene_type == 'protein_coding') %>% head(20) %>% pull(ID))
+plotter_split(gene_pool_2019 %>% tbl('limma_DE_gene') %>% filter(Comparison == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell') %>% left_join(gene_pool_2019 %>% tbl('gene_IDs')) %>% filter(gene_type == 'protein_coding') %>% head(20) %>% pull(ID))
 
 
 # GO Terms
 cadherin <- (gene_pool_2019 %>% tbl('all_vs_all_GO') %>% filter(Set == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell', ONTOLOGY == 'BP') %>% as_tibble() %>% filter(!grepl('HIST1', geneID)) %>% head(1) %>% pull(geneID) %>% str_split(.,"<br>"))[[1]]
-histones <- (gene_pool_2019 %>% tbl('all_vs_all_GO') %>% filter(Set == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell', ONTOLOGY == 'BP') %>% as_tibble() %>% filter(grepl('HIST1', geneID)) %>% head(1) %>% pull(geneID) %>% str_split(.,"<br>"))[[1]]
+
 
 plotter_split(cadherin)
-plotter_split(histones)
 plotter_split(gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble() %>% filter(grepl('HOXB', ID)) %>% filter(gene_type == 'protein_coding') %>% pull(ID))
 
 

### 2019-03-01 manuscript.Rmd
text changes
sup figs added

@@
 
-We plotted these sets of genes 
-Analysis of broad trends in GO term enrichment between differentially expressed genes shows that the HOXB family and protocadherins are differentially expressed between fetal retina and organoids.
-We noticed that several BRN3/POU4F genes were also differentially regulated
+We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina and rows are genes.
+We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
+We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.
 
@@
 
+```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
+rgc_cadherin_hoxb_cap <- supFig_cap(name='rgc_cadherin_hoxb', caption = 'Heatmap of three sets of genes across fetal retina and organoids divided by age in days (A).
+Bar plot of differential expression from retina to organoid, where positive values are genes that are higher expressed in fetal tissue than organoid.
+All logFC expression values have FDR corrected p value  < 0.01.')
+
+knitr::include_graphics('figures_and_tables/rgc_cadheren_box_fig.png')
+```
+
 ```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap, fig.pos = 'H'}
@@
 cat("<div id=\"tbl:GOterm_fetal_retina_organoid\">")
-organoid_fetal_GO %>% filter(ID %in% c('GO:0001503','GO:0031589','GO:0050878','GO:0061448','GO:0050678','GO:0030335','GO:0030198', 'GO:0007156', 'GO:0050769', 'GO:0030099')) %>% mutate(geneID = gsub('<br>', ' ', geneID)) %>% as_tibble() %>% regulartable() %>%
+organoid_fetal_GO %>% 
+  dplyr::select(-ONTOLOGY, -BgRatio, -p.adjust, -Count, -Test, -Set) %>% 
+  filter(ID %in% c('GO:0001503','GO:0031589','GO:0050878','GO:0061448','GO:0050678','GO:0030335','GO:0030198', 'GO:0007156', 'GO:0050769', 'GO:0030099')) %>% 
+  mutate(geneID = gsub('<br>', ' ', geneID)) %>% 
+  as_tibble() %>% 
+  regulartable() %>% 
+  flextable::fontsize(size = 8) %>% 
+  width(width = c(1,1,1,1,1,2)) %>% 
   flextable::align(align='center') %>% 
   font(fontname = 'Linux Libertine O', part = 'all') 
+#
 #Output Caption (* * for italics)

### 2019-03-01 src/data_pull.R

@@ -3,8 +3,8 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2019/EiaD_human_expression_2019_04.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/inst/app/www/2019/EiaD_human_expression_2019_03.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/inst/app/www/2017/eyeIntegration_human_2017_01.sqlite')
 
 
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
@@ -15,10 +15,12 @@ gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tx_anno <- gene_pool_2019 %>% tbl('tx_IDs') %>% as_tibble()
 limma_DE_tests <- gene_pool_2019 %>% tbl('limma_DE_tests') %>% as_tibble()
 q01_limma_gene_sig <- gene_pool_2019 %>% tbl('limma_DE_gene') %>% filter(adj.P.Val < 0.01) %>% group_by(Comparison) %>% summarise(Count=n()) %>% as_tibble()
+organoid_fetal_GO <- gene_pool_2019 %>% tbl('all_vs_all_GO') %>% filter(Set == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell', ONTOLOGY == 'BP') %>% as_tibble() %>% mutate(qvalue = as.numeric(qvalue)) %>% arrange(qvalue)
 save(core_tight_2017, 
      core_tight_2019, 
      tsne_50, gene_anno, 
      tx_anno, 
      limma_DE_tests, 
      q01_limma_gene_sig,
+     organoid_fetal_GO,
      file = 'data/data_pull.Rdata')

### 2019-03-01 src/retina_development_heatmaps.R

@@ -6,7 +6,7 @@ library(pool)
 library(RSQLite)
 gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
 
-rgc <- c('GAP43', 'POU4F1', 'ISL1', 'POU4F2','ATOH7','DLX2','SHH','DLX2', 'POU4F2', 'POU4F3', 'NEFL', 'GAP43', 'SNCG')
+rgc <- c('GAP43', 'POU4F1', 'ISL1', 'ATOH7','SHH','DLX2', 'POU4F2', 'POU4F3', 'NEFL', 'SNCG')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
 # cone_rod <- c('NEUROD1','CRX','RORB','GUCA1B','GUCA1A','GUCY2D','PRPH2','RP1','RBP3','TULP1','AIPL1','RCVRN','GUCY2F','SLC24A1')
 cone <- c('RXRB','THRB','RORA','GNAT2','ARR3','GNGT2','PDE6C','CNGA3','PDE6H','GNB3','GUCA1C','OPN1MW','OPN1SW','OPN1LW','GRK7')
@@ -274,21 +274,157 @@ related_split_plot + related_merge_plot
 
 
 
-# diff gene set -----
-plotter_split(gene_pool_2019 %>% tbl('limma_DE_gene') %>% filter(Comparison == 'Retina_Fetal.Tissue-Retina_3D.Organoid.Stem.Cell') %>% left_join(gene_pool_2019 %>% tbl('gene_IDs')) %>% filter(gene_type == 'protein_coding') %>% head(20) %>% pull(ID))
 
 
-# GO Terms
+
+# function to split by experiment / study / type -------
+plotter_split2 <- function(gene_vector, annotation = F, breaks = c(0,5,10,15), split) {
+  gene <- gene_vector
+  query = paste0('select * from lsTPM_gene where ID in ("',paste(gene, collapse='","'),'")')
+  p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
+    left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
+    as_tibble()
[…]

### 2019-03-01 manuscript.Rmd
discussion updates

@@
 
-We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina and rows are genes.
+We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina.
+As above, rows are genes, split by the three groups of genes and clustered by how related the expression patterns are.
 We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
@@
 
-As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed.
+As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for organoid-based models to be developed.
 Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level.
-This result underscores the value of these approaches in building models for human retina development.
+However as organoids do yet develop to full function, it is important to look at how gene expression systems differ between organoid and fetal tissue.
+https://www.ncbi.nlm.nih.gov/pubmed/12514132
+
+We use our large dataset to narrow in on some processes which to identify three core processes which differ.
+First we show that the HOXB family is overexpressed in the organoids.
+The homeobox family is well known to initiate polarity of the embryo during early development.
+As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the morphogen.
+While higher HOXB expression may be important to spur differentiation of the multiple cell types of the eye, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.
+
+Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids.
+As protocadherins mediate cell to cell connections and, in the develpoing mouse are shown to be important for spinal internneurons and RGC survival.
+We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result we see.
+Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues.
+This result suggests that genetically modifying stem cells to increase protocadherin expression may result in higher RGC activity.
+
+https://www.ncbi.nlm.nih.gov/pubmed/30266927
+
+http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha
+
+http://jcs.biologists.org/content/128/8/1455
+
+We built the Eye in a Disk dataset and the accompanying web app, eyeIntegration in the hopes that easily accessible gene expression across tissue space and time will be a
 

### 2019-03-02 manuscript.Rmd
text updates

--- 2019-03-01 manuscript.Rmd
@@
 abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
-METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across over 44 tissues in the GTEx project as a non-eye reference set.
-We then queried several public repositories to find all healthy, non perturbed, human eye tissue RNA-seq samples.
+METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set.
+We then queried several public repositories to find all healthy, non-perturbed, human eye tissue RNA-seq samples.
 The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote.
@@
 
-library(pool)
-library(RSQLite)
 library(tidyverse)
@@
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
-The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
+The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set f tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
 
@@
 
-Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delinerating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
+Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
 Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC).
@@
 
-While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regards.
+While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regard.
 We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from the human body tissues.
@@
 A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue.
-The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.
+The tables can be arranged or sorted to the user’s preference and can be easily downloaded for local use.
 
@@
 
-Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
+Heatmaps based on expression can be drawn for selected genes and tissues and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
 Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.
@@
 GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets.
-We also noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`.
+As there is some evidence suggesting that protocadherins influence RGC viability CITATION and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression.
 
-We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina and rows are genes.
+We plotted these three sets of genes with a heatmap visualization, with columns as age in days of fetal or organoid retina.
+Rows are genes, split by the three different groups of genes and are internally clustered by how similar the expression patterns are.
 We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
@@
 
-As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed.
-Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level.
-This result underscores the value of these approaches in building models for human retina development.
+As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for organoid-based models to be developed.
+Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues, to a first approximation, at the whole transcriptome level.
+However as organoids do yet develop to full function, it is important to look at how gene expression systems differ between organoid and fetal tissue so as to suggest areas for improvement.
+https://www.ncbi.nlm.nih.gov/pubmed/12514132
 
-Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with minimal effort, and because all the processing is in modular pieces it is easier to add new analysis steps.
-In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic datasets.
+We use our large dataset to narrow in on some processes which to identify three core processes which differ.
+First we show that the homebox (HOXB) family is overexpressed in the organoids.
+The homeobox family is well known to initiate polarity of the embryo during early development.
+As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the application of retinoic acid.
+While higher HOXB expression may be important to spur differentiation of the multiple cell types of the retina, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.
+
+Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids.
+Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival.
+We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result.
+Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues.
+This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.
+
+https://www.ncbi.nlm.nih.gov/pubmed/30266927
+
+http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha
+
+http://jcs.biologists.org/content/128/8/1455
+
+We built the Eye in a Disk dataset and the accompanying web app, eyeIntegration in the hopes that easily accessible gene expression across tissue space and time will be a useful tool for hypothesis generation and refinement in eye research.
+Wrapping all of the data processing steps in a Snakemake pipeline has several important features for the community: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with less effort, and because all the processing is in modular pieces it is easier to add new analysis steps.
+In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, expanding the single cell RNA-seq expression tooling, adding non-human eye samples, and epigenetic datasets.
 
@@
 rgc_cadherin_hoxb_cap <- supFig_cap(name='rgc_cadherin_hoxb', caption = 'Heatmap of three sets of genes across fetal retina and organoids divided by age in days (A).
-Bar plot of differential expression from retina to organoid, where positive values are genes that are higher expressed in fetal tissue than organoid.
+Bar plot of differential expression from retina to organoid, where positive values are genes that are higher expressed in fetal tissue than organoid (B).
 All logFC expression values have FDR corrected p value < 0.01.')

### 2019-03-02 manuscript.Rmd
acknowldgements edit

@@
 We would like the thank the dozens of groups who provided the raw data required to create this project.
-We are keeping a running list of the projects and citations at https://github.com/davemcg/eyeIntegration_app/blob/master/inst/citations.md and strongly encourage anyone who uses EiaD and eyeIntegration to cite relevant projects.
-We would also like to thank Kapil Bharti, Robert Hufnagel, and Brian Brooks for their continuous set of critiques and suggestions of the eyeIntegration app over the past two years.
-This work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
+We keep a running list of the projects and associated citations at https://github.com/davemcg/eyeIntegration_app/blob/master/inst/citations.md and strongly encourage anyone who uses EiaD and eyeIntegration to cite relevant projects.
+We would also like to thank Kapil Bharti, Robert Hufnagel, and Brian Brooks for their continuous set of critiques and suggestions in the development of eyeIntegration app over the past two years.
+Tiziana Cogliati was especially helpful in the editing of this manuscript.
+Finally, this work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
 

### 2019-03-02 cover_letter.Rmd
cover letter, text updates

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database
  version 1.0'
author: "David McGaughey"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: word_document
---

To the editor,

Dozens of research groups have deposited hundreds of human eye tissue RNA-seq transcriptomes in publicly available databases. To maximize the value of these datasets, we have built the "Eye in a Disk" (EiaD) dataset which contains all available non-perturbed eye tissues and the Gene Tissue Expression (GTEx) project human body tissues processed in a consistent and reproducible manner. To best serve this huge dataset, we have re-built the eyeIntegration web app (https://eyeIntegration.nei.nih.gov) to efficiently access the data and provide power interactive visualizations of gene expression across several tissue types within the human eye and across the entire human body.

In a demonstration of the value of this large, uniformly processed dataset, we show both that 3D retina organoids and fetal tissue retina have highly similar transcriptomes. Since organoids do not yet fully mature, we also carefully studied gene expression differences between these two types and pinpoint how homeobox and protocadherin developmental processes are differentially expressed between the organoids and tissue. These processses may have relevance in retinal ganglion activity and the maturation of the organoids.

The EiaD dataset and the eyeIntegration web app is freely available to download and deploy on local computers. As of January 2019, this resource has been used by researchers across 285 cities and 40 countries. The improvements we have made to this project in creating the reproducible databases and the matched web app will allow for continuous improvements in sample and feature additions over time.

Sincerely,
David McGaughey Ph.D, Bioinformatics Group, OGVFB, NEI

### 2019-03-02 manuscript.Rmd
cover letter, text updates

@@
 ---
-title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00'
+title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
 author:
@@
 The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
-RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.02 as of Februay 24 2019, https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues.
-As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest targets for organoid process improvements.
-CONCLUSION: The eyeIntegration v1.02 web app serves the pan-eye and body transcriptome dataset, EiaD.
+RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing quick visualizations by database version, gene/transcript, and 60 tissues.
+As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest organoid process improvements.
+CONCLUSION: The eyeIntegration v1.0 web app serves the pan-eye and body transcriptome dataset, EiaD.
 This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."

### 2019-03-02 manuscript.Rmd
updates

@@
 RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing quick visualizations by database version, gene/transcript, and 60 tissues.
-As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest organoid process improvements.
+As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but display distinct differences in protocadherin and HOXB family members.
 CONCLUSION: The eyeIntegration v1.0 web app serves the pan-eye and body transcriptome dataset, EiaD.
@@
 The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
-However, no eye tissues are included in their resource.
+Recently Ratnapriya et al.
+have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link.
+This data, as of March 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
 The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
@@
 
-**ADD MENTION OF NEW SWAROOP DATA "GREx"**
-
 ## eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
@@
 The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples.
-The interactive nature of some of the steps precludes efficient and regular data updates for the data.
+The interactive nature of some of the steps precluded efficient and regular data updates for the data.
 
@@
 We call this reproducible and versioned transcriptome dataset "Eye in a Disk" (EiaD).
-The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provide consistent sample quality control, and improves efficiency for future sample updates.
+The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provides consistent sample quality control, and improves efficiency for future sample updates.
 The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling.
@@
 
-## EiaD dataset and eyeIntegration used to identify potential avenues to improve retina organoid maturation
+## EiaD dataset used to identify potential avenues to improve retina organoid maturation
 
-Retina organoids
+Retina organoids are an increasingly popular means to model human retina development.
+We use our pan-study EiaD dataset to demonstrate that, at a pan-transcriptome level, organoids are highly similar to fetal retina tissue.
+We also demonstrate that important temporal gene expression patterns in the fetal retina tissue is recapitulated in the organoids.
+As the organoid differentiation methods do not yet produce fully mature retina, we focused on identifying differentially expressed processes, identifying protocadherin and HOXB family gene expression differences which suggest means to improve and benchmark organoid differentiation methods.
 
@@
 
-Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
+Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD.
+We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
 Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC).
@@
 ## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
+
 We performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing quick identification of eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons.
@@
 First we show that the homebox (HOXB) family is overexpressed in the organoids.
-The homeobox family is well known to initiate polarity of the embryo during early development.
-As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the application of retinoic acid.
-While higher HOXB expression may be important to spur differentiation of the multiple cell types of the retina, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.
+The homeobox family is well known to initiate polarity of the embryo during early development [@philippidouHoxGenesChoreographers2013].
+Retinoic acid is applied at about day 20 to help differentiate stem cell to organoids and is also known to activate the HOXB family.
+The lack of HOXB expression at any age in our fetal tissue retina dataset and the broad chromatin and gene expression changes HOXB family members can mediate suggests that HOXB activity may be unwanted for organoid maturation.
 
 Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids.
-Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival.
-We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result.
-Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues.
+Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival [@hayashiEmergingRolesProtocadherins2015; @lefebvreGProtocadherinsRegulateNeuronal2008].
+We would predict that decreased protocadherin expression would reduce RGC activity.
+Indeed we see that many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues.
 This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.
-
-https://www.ncbi.nlm.nih.gov/pubmed/30266927
-
-http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha
-
-http://jcs.biologists.org/content/128/8/1455
 

### 2019-03-15 manuscript.Rmd
adding tc edits round 2

@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
+abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq based transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
 METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set.
-We then queried several public repositories to find all healthy, non-perturbed, human eye tissue RNA-seq samples.
-The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote.
-The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
-RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing quick visualizations by database version, gene/transcript, and 60 tissues.
-As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but display distinct differences in protocadherin and HOXB family members.
-CONCLUSION: The eyeIntegration v1.0 web app serves the pan-eye and body transcriptome dataset, EiaD.
-This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
+We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples.
+The 1311 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD.
+RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, and 60 tissues.
+As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retinae are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families such as, for example, protocadherin and HOXB family members.
+CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD.
+This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 44 body and 16 eye tissues."
 keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
@@
 
-From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid.
+From anterior to posterior along the light trajectory, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid.
 The differentiation, maturation, and function of these tissues is mediated through spatial and temporal specific transcript and gene expression patterns, also known as the transcriptome.
 Today, RNA-sequencing (RNA-seq) is the predominant technology for quantifying the transcriptome.
-Analysis of the transcript expression across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
-To this end, researchers have used a wide variety of human tissue sources to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
+Analysis of the transcripts' expression across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function.
+To this end, a wide variety of human tissue sources have been used to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
 These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014b; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014b; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015a; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
@@
 ## The GTEx gene expression resource lacks eye-specific tissues
-The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
+The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a comprehensive and user-friendly web app (https://gtexportal.org/); however eye tissues have not been included [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
 Recently Ratnapriya et al.
@@
 The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
-To date, no curation of the sample level metadata has been done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed.
+To date, no curation of the sample level metadata has been done, therefore it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed.
 More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016].
-iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.
+However iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.
+We thus aimed our efforts at developing an easily accessible and reliable RNA-seq based transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
 
-## eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
-The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
+## The eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
+The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017 at version 0.6, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
 This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues.
-The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples.
+The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to manually assess quality control for the samples.
 The interactive nature of some of the steps precluded efficient and regular data updates for the data.
@@
 
-## EiaD dataset used to identify potential avenues to improve retina organoid maturation
+## The EiaD dataset can be used to identify potential avenues to improve retina organoid maturation
 
 Retina organoids are an increasingly popular means to model human retina development.
-We use our pan-study EiaD dataset to demonstrate that, at a pan-transcriptome level, organoids are highly similar to fetal retina tissue.
-We also demonstrate that important temporal gene expression patterns in the fetal retina tissue is recapitulated in the organoids.
-As the organoid differentiation methods do not yet produce fully mature retina, we focused on identifying differentially expressed processes, identifying protocadherin and HOXB family gene expression differences which suggest means to improve and benchmark organoid differentiation methods.
+We used our pan-study EiaD dataset to show that, at a pan-transcriptome level, organoids are highly similar to early fetal retina tissue.
+We also show that important temporal gene expression patterns in the fetal retina tissue are recapitulated in the organoids.
+As the organoid differentiation methods do not yet produce fully mature retina, we focused on identifying differentially expressed processes between organoid retina and embryonic retina and detected, for example, identifying protocadherin and HOXB family gene expression differences which suggest targetable pathways to improve and benchmark organoid differentiation methods.
 
@@
 ## Identification of potential eye samples 
-We exhaustively searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
+We exhaustively searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'SRA' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
 We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, induced pluripotent stem cell (iPSC)-derived tissue, stem cell derived organoids, and immortalized cell lines.
@@
 ```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
-tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type from 2017 to 2019 EiaD.
-We also improved the metadata labelling, the cornea samples now delineats endothelial and epithelial tissues and the retina samples distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.')
+tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = paste0('Substantial increase in eye tissue count and type from 2017 (', `r core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682'`, ' lighter color) to 2019 (', `r core_tight_2017 %>% filter(Kept=='Kept', study_accession != 'SRP012682'`, ' darker color) EiaD.
+We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (teal) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.'))
 
@@
 
-Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19, 2017.
+Our query on January 7th 2019 to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies.
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.
@@
 The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'ESC') %>% nrow()` undifferentiated iPSC, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`).
+We refer to native-tissue extracted RPE as RPE (choroid) because it is not possible to remove the choroid from the RPE without culturing.
 
@@
 We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue.
-Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC).
-Another substantial addition to the 2019 EiaD are non-protein coding genes.
-We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).
+Another substantial addition to the 2019 EiaD are non-protein coding genes; while protein-coding is the most common gene and transcript typse, there are dozens of different non-coding classes.
+The 2017 version of eyeIntegration only quantified protein coding genes and transcripts.
+We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodogenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).
 
@@
 After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')`).
-For example, in retina-derived tissue, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina tissue cluster together.
-Our ability to uniformly cluster data by known biological function independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.
+For example, in the retina group, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina samples cluster together.
+Our ability to uniformly cluster data by known biological source independent of study origin demonstrates that our workflow can effectively account for technical variation between studies.
 
 While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regard.
-We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from the human body tissues.
+We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from all other human body tissues.
 Adult retina is most similar to the brain tissue, pituitary is similar to adult RPE, and skin and blood are similar to cornea.
@@
 
-## eyeIntegration web app provides interactive visual portal to all data 
+## The eyeIntegration web app provides interactive visual portal to all data 
 The EiaD 2019 dataset is used directly by the eyeIntegration web app (https://eyeIntegration.nei.nih.gov).
 The web app was designed to provide a simple interface that has the same general concept – select specific genes and tissue and view relevant information.
-The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.
+The web-app is divided into four general categories: expression, two-dimension sample relationships, gene networks, and data tables.
 
@@
 
-Heatmaps based on expression can be drawn for selected genes and tissues and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
+Heatmaps based on expression can be drawn for selected genes and tissues and gene expression can be compared across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
 Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.
@@
 
-The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values.
+The results of differential expression are presented in a tabular format showing log~2~ fold change, average expression, and p-values.
 Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (`r supTab_cap(name = 'limmaTests', display = 'cite')`).
@@
 
-## EiaD 2019 demonstrates that iPSC-derived organoids and fetal retina have closely related transcriptomes
-There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids.
+## EiaD 2019 suggests that iPSC-derived organoids and fetal retina have closely related transcriptomes
+There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell derived organoids.
 We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.
 
-To evaluate how the tissues and organoids compare at a transcriptome level, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the three types of retina tissue (adult, fetal, and organoid) (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'A')`).
-Here we see three distinct groupings: adult retina (1), developing fetal retina and stem cell organoid (2), and undifferentiated and early differentiating stem cells (3).
-We do see several organoid samples in cluster 3, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
+To evaluate how the tissues and organoids compare at a transcriptome level, we looked at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focused in on the three types of retina tissue (adult, fetal, and organoid) (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'A')`).
+Here we saw three distinct groupings: adult retina (1), developing fetal retina and stem cell-derived organoid (2), and undifferentiated and early differentiating stem cells (3).
+We identified several organoid samples in cluster 3, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X').
 All of the organoid retina samples in cluster 2 are older than 50 days.
 
-To assess how similarly the fetal and organoid retina develop through time, we plot expression of retinal progenitors, photoreceptors, and retinan ganglion markers by time in days (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`).
-Each row is a gene marker of either progenitor, photoreceptor, or retinal ganglion cells.
+To assess how similarly the fetal and organoid retina develop through time, we plotted expression of retinal progenitors, photoreceptors, and retinal ganglion markers by time in days (`r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`).
+Each row is a gene marker of either retinal progenitor, photoreceptor, or RGC.
 The rows are hierarchially clustered to put more similar expression patterns closer together, as denoted by the height of the dendrogram.
@@
 samples [@eldredThyroidHormoneSignaling2018; @kaewkhawTranscriptomeDynamicsDeveloping2015].
-The Kaewkhaw samples are flow sorted for a GFP marker under the control of a *CRX* promoter, an important of photoreceptor development.
-GFP+ cells thus are believed to be enriched in photoreceptor populations.
-We see that the progenitor, photoreceptor, and retinal ganglion groups are largely clustered together, with patterns of expression consistent across the retinal and organoid groups.
+The Kaewkhaw samples are flow sorted for a GFP marker (GFP+) under the control of the *CRX* promoter, an important regulator of photoreceptor development.
+GFP+ cells wouldbe enriched in photoreceptor populations.
+We saw that the retinal progenitor, photoreceptor, and RGC groups are largely clustered together, with patterns of expression consistent across the fetal retinal and organoid groups.
 
@@
 
-## Organoid retina differential gene expression versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and retinal ganglion function (BRN3/POU4F, NEFL, GAP43, SNCG)
+## Differential gene expression of organoid retina versus fetal tissue identifies sets of genes relating to patterning (HOXB family), cell adhesion (protocadherin family), and RGC identity (BRN3/POU4F, NEFL, GAP43, SNCG)
 
-To identify specific changes between organoid and retina tissue, we did differential gene expression and GO term enrichment analyses.
-GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets.
-As there is some evidence suggesting that protocadherins influence RGC viability CITATION and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression.
+To identify specific changes between retinal organoid and fetal retina tissue, we performed differential gene expression and GO term enrichment analyses.
+The GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets in retinal organoids As there is some evidence suggesting that protocadherins influence RGC viability CITATION and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression [@lefebvreGProtocadherinsRegulateNeuronal2008].
 
-We plotted these three sets of genes with a heatmap visualization, with columns as age in days of fetal or organoid retina.
+We plotted HOXB family, protocadherin family and RGC genes in a heatmap visualization, with columns as age in days of fetal or organoid retina.
 Rows are genes, split by the three different groups of genes and are internally clustered by how similar the expression patterns are.
-We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
-We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.
+We observed that there are strong, consistent gene expression differences in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
+We also plotted the differential expression values between all organoids and all fetal retina samples; all genes across all three groups are significantly differentially expressed with an FDR corrected p value < 0.01.
 
@@
 As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for organoid-based models to be developed.
-Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues, to a first approximation, at the whole transcriptome level.
-However as organoids do yet develop to full function, it is important to look at how gene expression systems differ between organoid and fetal tissue so as to suggest areas for improvement.
-https://www.ncbi.nlm.nih.gov/pubmed/12514132
+Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) indicates that these two approaches successfully recapitulate fetal retina tissue, to a first approximation, at the whole transcriptome level.
+However as organoids do not develop to full function, it is important to look at how gene expression differs between retinal organoid and fetal tissue so as to suggest areas for improvement.
 
-We use our large dataset to narrow in on some processes which to identify three core processes which differ.
-First we show that the homebox (HOXB) family is overexpressed in the organoids.
-The homeobox family is well known to initiate polarity of the embryo during early development [@philippidouHoxGenesChoreographers2013].
-Retinoic acid is applied at about day 20 to help differentiate stem cell to organoids and is also known to activate the HOXB family.
-The lack of HOXB expression at any age in our fetal tissue retina dataset and the broad chromatin and gene expression changes HOXB family members can mediate suggests that HOXB activity may be unwanted for organoid maturation.
+We used our large dataset to narrow in on three core processes which differ significantly and substantially between retinal organoids and fetal retina.
+First we showed that the HOXB family is overexpressed in the organoids.
+The homeobox family is well known to initiate polarity of the embryo during early development63 [@philippidouHoxGenesChoreographers2013].
+Retinoic acid is applied at about day 20 in culture to help differentiate stem cell to organoids and is also known to activate genes members of the HOXB family.
+The lack of HOXB expression at any age in fetal retina and the broad chromatin and gene expression changes HOXB family members can mediate suggests that HOXB activity may be unwanted for organoid maturation.
 
-Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids.
+Next we detected many protocadherins that are more highly expressed in the fetal tissue, relative to the organoids.
 Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival [@hayashiEmergingRolesProtocadherins2015; @lefebvreGProtocadherinsRegulateNeuronal2008].
-We would predict that decreased protocadherin expression would reduce RGC activity.
-Indeed we see that many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues.
-This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.
+We would predict that decreased protocadherin expression reduces the number and maturation of RGC.
+Indeed we observed that many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed relative to fetal tissue.
+This result suggests that modifying culture conditions to promote protocadherin expression may result in higher RGC yield and survival.
 
 We built the Eye in a Disk dataset and the accompanying web app, eyeIntegration in the hopes that easily accessible gene expression across tissue space and time will be a useful tool for hypothesis generation and refinement in eye research.
-Wrapping all of the data processing steps in a Snakemake pipeline has several important features for the community: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with less effort, and because all the processing is in modular pieces it is easier to add new analysis steps.
+Wrapping all of the data processing steps in a Snakemake pipeline has several important advantages for the community: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with less effort, and because all the processing is in modular pieces it is easier to add new analysis steps.
 In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, expanding the single cell RNA-seq expression tooling, adding non-human eye samples, and epigenetic datasets.

### 2019-03-15 cover_letter.Rmd
final edits before submission to IOVS and bioRxiv!

@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output: word_document
+output:
+  word_document:
+    reference_docx: word-styles-reference-01.docx
+    fig_caption: yes
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
 ---

### 2019-03-15 manuscript.Rmd
final edits before submission to IOVS and bioRxiv!

@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq based transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
-METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set.
+abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
+METHODS: We downloaded the raw sequnce data for 892 RNA-seq samples across 44 tissues in the GTEx project as a non-eye reference set.
 We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples.
-The 1311 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+The 1311 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 450 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
 Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD.
-RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, and 60 tissues.
-As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retinae are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families such as, for example, protocadherin and HOXB family members.
+RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, and 58 tissues.
+As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members.
 CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD.
-This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 44 body and 16 eye tissues."
-keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
+This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 44 body and 14 eye tissues."
+keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
 ---
@@
 ## The EiaD dataset can be used to identify potential avenues to improve retina organoid maturation
-
 Retina organoids are an increasingly popular means to model human retina development.
@@
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchielimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
+We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
@@
 ## Manuscript as code
-This manuscript's figures (excepting (Figure 2)), tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
+This manuscript's figures, tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
 The knitr and pandoc program is used to lay out the figures and tables and output a docx file.
@@
 ```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
-tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = paste0('Substantial increase in eye tissue count and type from 2017 (', `r core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682'`, ' lighter color) to 2019 (', `r core_tight_2017 %>% filter(Kept=='Kept', study_accession != 'SRP012682'`, ' darker color) EiaD.
+tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption =  paste0('Substantial increase in eye tissue count and type from 2017 (', core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow(), ', lighter color) to 2019 (', core_tight_2019 %>% filter(Kept=="Kept", study_accession != 'SRP012682') %>% nrow(),', darker color) EiaD.
 We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (teal) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.'))
@@
 regulartable(table) %>% 
-  width(width = c(0.5,1,0.5, 2)) %>% 
+  width(width = c(1,1,1,2)) %>% 
+  flextable::align_text_col(align = "center", header = TRUE) %>% 
   flextable::align(align='center') %>% 
+  fontsize(size = 10) %>% 
   font(fontname = 'Linux Libertine O', part = 'all')
@@
 C.
-CRX gene expression heatmap and table information from Macosko et al.
-P14 scRNA-seq.
+CRX mouse retina gene expression heatmap and table information from Clark et al.
+E11 to P14 scRNA-seq.
 D.
@@
 The results of differential expression are presented in a tabular format showing log~2~ fold change, average expression, and p-values.
-Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (`r supTab_cap(name = 'limmaTests', display = 'cite')`).
+Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (Supplementary Files).
 The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format.
@@
 Rows are genes, split by the three different groups of genes and are internally clustered by how similar the expression patterns are.
-We observed that there are strong, consistent gene expression differences in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
+We observed that there are strong, consistent gene expression differences in these three groups of genes between fetal retina and the organoid samples (`r paste0(supFig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`).
 We also plotted the differential expression values between all organoids and all fetal retina samples; all genes across all three groups are significantly differentially expressed with an FDR corrected p value < 0.01.
@@
 
-Next we detected many protocadherins that are more highly expressed in the fetal tissue, relative to the organoids.
+Next, we detected several protocadherins more highly expressed in the fetal tissue, relative to the organoids.
 Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival [@hayashiEmergingRolesProtocadherins2015; @lefebvreGProtocadherinsRegulateNeuronal2008].
@@
 
-```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
+```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = rgc_cadherin_hoxb_cap}
 rgc_cadherin_hoxb_cap <- supFig_cap(name='rgc_cadherin_hoxb', caption = 'Heatmap of three sets of genes across fetal retina and organoids divided by age in days (A).
@@
   width(width = c(0.5,1,0.5,3)) %>% 
+  flextable::align_text_col(align = "center", header = TRUE) %>% 
   flextable::align(align='center') %>% 
@@
   width(width = c(2,1,2)) %>% 
+  flextable::align_text_col(align = "center", header = TRUE) %>% 
   flextable::align(align='right') %>% 
@@
 
-```{r limmaTests, echo = F, results='asis', warning = F, fig.cap = limmaTests_cap, fig.pos = 'H'}
-limmaTests_cap <- supTab_cap(name='limmaTests', caption = 'Thousands of differentially expressed genes found (q < 0.01) for the 105 comparisons we ran.')
-# Output <div>
-cat("<div id=\"tbl:limmaTests\">")
-regulartable(q01_limma_gene_sig) %>% 
-  width(width = c(4,2)) %>% 
-  flextable::align(align='center') %>% 
-  font(fontname = 'Linux Libertine O', part = 'all')
-#Output Caption (* * for italics)
-cat(paste0("\n*",limmaTests_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
 ```{r GOterm_fetal_retina_organoid, echo = F, results='asis', warning = F, fig.cap = GOterm_cap, fig.pos = 'H'}
@@
   width(width = c(1,1,1,1,1,2)) %>% 
+  flextable::align_text_col(align = "center", header = TRUE) %>% 
   flextable::align(align='center') %>% 

### 2019-03-15 src/limmaTest_sup.R

@@ -1,0 +1,2 @@
+load('data/data_pull.Rdata')
+WriteXLS::WriteXLS(q01_limma_gene_sig, ExcelFileName  ='supplementary_materials/supplementary_files__limmaTests.xls')

### 2019-03-15 src/limmaTest_sup.R
added both sup materials

@@ -1,2 +1,3 @@
 load('data/data_pull.Rdata')
 WriteXLS::WriteXLS(q01_limma_gene_sig, ExcelFileName  ='supplementary_materials/supplementary_files__limmaTests.xls')
+WriteXLS::WriteXLS(core_tight_2019, ExcelFileName  ='supplementary_materials/supplementary_files__metadata.xls')

### 2019-04-22 2019_04_22_Reviewer_Comments.Rmd
rewviewr comments

---
title: 'Reviewer Comments for Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '2019-04-22'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
---

```
AE Comments to Author:
The manuscript "Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0" is an interesting compilation of information from many datasets. The reviewers have a number of comments and queries for consideration.

Editorial Board Member (Comments for the Author (Required)):

This manuscript describes a database on genes from tissues of the eye into an accessible system for other researchers to access. They use one particular example, fetal eye tissue and organoid retina, as an example of the power of the database as constructed. While there is no hypothesis here, the ability of others to mine this curated data is likely to be of great interest to many researchers. Both reviewers have requested clarification and elaboration.

1. Figure 4 is not readable. If this is going to be useful illustration of potential, it would be better to do each of the 5 subparts as a separate figure. Similarly, Figure 5 would be better separated into 2 figures with increased readability.
2. Minor typos in the manuscript, but otherwise clearly written.
```

```
Reviewer #1 (Comments for the Author (Required)):

Building on the eyeIntegration web resource released in 2017, the authors re-wrote the bioinformatics pipeline and created one of the most comprehensive "reproducible and versioned transcriptome dataset" - eye in a Disk (EiaD) and updated the eyeIntegration web app, now integrating deep sequencing data from 371 human eye tissue samples and 848 GTEx samples across 44 tissues to allow quick search and visualization of human eye-related transcriptomes and make comparisons and identify differentially expressed genes among 14 various eye tissues and 44 other body tissues. It also included newly-published single cell seq data of developing and adult mouse retina. It is a powerful addition to the bioinformatics tools for the visual science community. In general, the manuscript is straightforward description of the development of the dataset and a guide to the web app. There are some minor clarifications should be done:
1) The authors quantified expression "across 42 gene and 46 transcript types" (supplemental table 2). Please describe the difference or relationship between of the "gene" and "transcript" types. Are they pretty much overlapping to each other?
2) "Figure 2: Raw RNA seq data ....... Visualizations across 63 tissues": are these 63 tissues the same as described in the abstract "44 body and 14 eye tissues", what are the extra 5 [63-(44+14)]? 
3) Line 179: "To create a high quality final dataset across the 1311 initial samples (suppl fig and table 2)". Are these "1311 initial samples" the same as "371 human eye tissue samples and 848 GTEx" described on line 158-159. If so, 371+848 ≠ 1311? 
4) In Figure 3, the colors of ESC and lens, retina and RPE are too close to each other, and hard to be distinguished. More distinctive, different colors would help. What is the total number of clusters? A table listing all the clusters and the composition of each cluster should be included as a supplemental table. 
5) Line 208, ".... Then genes (or transcripts), then tissues." The "genes" should be "gene IDs by official gene name". The tables derived from "expression" search (e.g. Fig.4A) should be further annotated, e.g. define what are "decile", "log2(PMT+1)", etc. 
6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
7) Many typos and some unfished edit should be corrected, e.g. Line 285 "citation" should be added!

Reviewer #2 (Comments for the Author (Required)):

In this manuscript the authors describe aggregation of RNA-seq data from 371 ocular tissue samples into an "Eye in a Disk (EiaD)" database. Samples from adult and fetal retina, retinal pigment epithelium (RPE), cornea, lens, induced pluripotent stem cell (iPSC) derived retinal organoids are included. Data from two single cell RNA-seq (scRNA-seq) datasets generated from developing mouse retina are also included. The aggregated data are presented for general use via the eyeIntegration web interface. Data from 844 Genotype-Tissue Expression (GTEx) project samples derived from 44 different tissues are included for reference. To generate the EiaD database, data from RNA-seq studies that were available in the NIH Short Read Archive (SRA) were re-analyzed using the alignment free quantification tool Salmon, using the Gencode transcriptome index as a reference. Transcripts with low levels of expression (<2 TPM) were removed from the resulting dataset, as were samples with low calculated mapping rates. The resulting data were stored in an SQLite database. By using a workflow management system called SnakeMake, the authors ensure that reproducibility is maintained with each iteration of the data, facilitating sample tracking and updates of the database and portal, as new transcriptome data for the eye and other tissues become available. The whole analysis pipeline is available via Github, as is the R package used to generate the eyeIntegration web application.

In addition to generating the EiaD, the authors created an interactive website with various useful features for visualizing and exploring expression data, such as differential gene expression between tissues, gene and tissue clustering based on expression profiles, weighted gene correlation networks around genes specified by the user, and t-SNE plotting, a data visualization tool for comparing expression data between different sample sets. To demonstrate these tools, the authors used t-SNE and PCA analyses to compare the gene expression profiles of the ocular and non-ocular tissues in the EiaD. They found that 
in general data from eye tissues cluster together, and that these eye tissues clusters are mostly separate from non-ocular GTEx tissue clusters. Further, analyses of RNA-seq data from iPSC derived retinal organoids and fetal retinal samples showed that the retinal organoids are similar but not identifical to fetal retina with regard to gene expression profiles. For example, HOXB genes are expressed in the retinal organoids, not fetal retina.

The EiaD and eyeIntegration web application are a unique transcriptome database and data visualization portal centered on the eye, which enables quick exploration of gene and transcript expression between various healthy eye tissue types and a range of non-ocular human tissues. While the database and web tools are a valuable resource for the vision research community, there are a number of issues that need to be addressed/clarified.

Comments/Issues:

1. A common practice following identification of differential gene expression between two tissue types is experimental validation of a subset of the identified differences. Given the "dry lab" nature of this manuscript, this does not seem appropriate to request. But, it would be helpful to know if the differences detected in gene expression between iPSC derived retinal organoids and fetal retinal samples have been reported by other investigators. Similarly, it would be helpful to compare the output of the EiaD pipeline with similar data from the GTEx pipeline. For example, since the differences in gene expression between tissue types has been pre-calculated, it should be relatively easy to compare differential expression results for comparisons between GTEx tissues produced from GTEx portal and from the EiaD pipeline.

2. It would be helpful to provide some information about the accuracy of the Salmon transcript quantification tool for the reported analyses. For example, a search for Rhodopsin expression in the single cell RNA seq data shows relatively high expression in cone cells, in which rhodopsin is not usually expressed. Is this an artefact of the scRNA-seq method, or a possible read assignment error from the Salmon tool?

This question is also being asked because Salmon is an alignment-free method that performs light weight mapping of reads to transcript models and estimates expression levels of transcripts correcting for potential biases such as GC content. Given the limitations of transcript reconstruction from short RNA-seq reads, a commonly used approach to unify data from different studies, including GTEx, ENCODE, TopMed, and TCGA has been uniform realignment, gene expression quantification, and batch effect removal (e.g. Sci Data 2018 - DOI: 10.1038/sdata.2018.61). While Salmon has been reported to be among the top performing methods for transcript quantification, it also has limitations. For example, Salmon does not seem to perform as well for lowly expressed transcripts, and its efficacy can be affected by incompleteness of the gene/transcript annotation model or total number of isoforms per gene (BMC Genomics 18:583, 2017; Briefings in Bioinformatics 18: 260, 2017). It would be helpful to include information about the choice and use of this method in the manuscript.

3. It would be helfpul to learn more about how the authors handled non-biological batch effects between studies in their differential expression analyses. For example, in the pan tissue t-SNE plot shown in Figure 1 the majority of eye tissues cluster next to each other, but there are two separate clusters of retina and RPE samples that are distant from the other retina and RPE sample sets. Are these differences real, or due to differences in the RNA-seq library preparations and sequencing protocols between the different samples downloaded from SRA? One potential approach to this issue is to remove non-specific variation from the expression values in the differential expression model, as reported by Wang et al in their unification of GTEx and TCGA RNA-seq sample data in order to compare expression levels between normal and tumor samples from two different studies (Sci Data 2018, DOI: 10.1038/sdata.2018.61). One method that can be used to infer hidden covariates that capture much of the non-specific variation in expression datasets is ComBat in the R package SVAseq that enables indicating known variables that should be adjusted for, such as different studies (Nucleic Acids Res. 42:e161, 2014; Biostatistics 8:118-127, 2007). It is recommended that this be applied to each tissue type separately, combining samples from the same tissue type, such as retina, across different studies, and adding the study as an indicator in ComBat.

4. Further information about the quality control and filtering applied to the RNA-seq data would be helpful to include in the manuscript. For example, how were the parameters for filtered by a mapping rate of 40% per sample and median read count >2 across all genes per sample selected? Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included.

5. To identify outlier samples the authors measured the distances of each sample from the centroid coordinates per cluster using a t-SNE plot. t-SNE is good for visualization but not for quantification of relative distance between samples and sample sets. The t-SNE retains probabilities not distances and hence the Euclidean distances between points on the 2-dimensional t-SNE plot are not informative. t-SNE documentation suggests that Kullback-Leibler divergences reported by t-SNE can be used for this purpose. This reviewer recommends use of coordinates from other dimensionality reduction methods such as Principal Coordinates Analysis (PCoA) or PCA, or a correlation-based statistic that compares the expression profile of each sample to the profiles of all samples in a given sample set (D statistic) to identify outlier samples, such as described in Nature Genetics 46: 430, 2014.

6. It would be helpful to learn why the authors chose 892 samples from 44 tissues in GTEx, when there are over 11,000 RNA-seq sample available for 50 tissues in the current GTEx release v7 that is available on dbGaP. Also, a recent publication from other investigators at NEI described a large RNA-seq dataset from human retinal samples (Nat Genet. 51:606-610, 2019). Will these data be incorporated into the eyeIntegration database? If so, it would be helpful if this were mentioned in the text. If not, this additional study and the resources derived from it could be referred to.
```

### 2019-04-24 2019_04_22_Reviewer_Comments.Rmd
reviewer comment response

@@
 ```
+Move Figure 4 to supplemental figure?
+
+Split Figure 5 into two figures?
 
@@
 There are some minor clarifications should be done:
+```
+
+```
 1) The authors quantified expression "across 42 gene and 46 transcript types" (supplemental table 2).
@@
 Are they pretty much overlapping to each other?
+```
+
+All quantification of expression happens at the transcript level.
+As most people think in terms of gene expression, we collapse all of the transcript level information information into a gene.
+Very briefly, the expression from each transcript of a gene is averaged, with scaling for the length of the transcript.
+If there are transcripts (of equal length) a1 (5 counts), a2 (6 counts), and a3 (7 counts) for gene a, then gene a would have an expression of (5 + 6 + 7) / 3.
+
+```
 2) "Figure 2: Raw RNA seq data ....... Visualizations across 63 tissues": are these 63 tissues the same as described in the abstract "44 body and 14 eye tissues", what are the extra 5 [63-(44+14)]? 
+```
+
+```
 3) Line 179: "To create a high quality final dataset across the 1311 initial samples (suppl fig and table 2)".
@@
 If so, 371+848 ≠ 1311?
+```
+
+```
 4) In Figure 3, the colors of ESC and lens, retina and RPE are too close to each other, and hard to be distinguished.
@@
 A table listing all the clusters and the composition of each cluster should be included as a supplemental table.
+```
+
+```
 5) Line 208, "....
@@
 define what are "decile", "log2(PMT+1)", etc.
+```
+
+```
 6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.

### 2019-05-04 2019_04_22_Reviewer_Comments.Rmd
adding comments and code bits

@@
 All quantification of expression happens at the transcript level.
-As most people think in terms of gene expression, we collapse all of the transcript level information information into a gene.
-Very briefly, the expression from each transcript of a gene is averaged, with scaling for the length of the transcript.
-If there are transcripts (of equal length) a1 (5 counts), a2 (6 counts), and a3 (7 counts) for gene a, then gene a would have an expression of (5 + 6 + 7) / 3.
+As most researchers think in terms of gene expression, we collapse all of the transcript level information information into a gene.
+Very briefly, the tximport tool (https://f1000research.com/articles/4-1521/v2) is used to collapse expression from each transcript of a gene is averaged, with scaling for the length of the transcript.
+
+We have
 
@@
 ```
+
+Number check!
 
@@
 ```
-
+Number check!
 ```
@@
 ```
+HGNC gene names
+```
+6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
+```
+
+Add section to suggest ways researchers can faciliate addition of their data to eyeIntegration.
+
+1. Contact us directly (mcgaugheyd@mail.nih.gov) 
+2. Deposit data on SRA
+3. Clear, descriptive, consistent, metadata for each sample (in the sample_attribute section)
 
 ```
-6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
 7) Many typos and some unfished edit should be corrected, e.g. Line 285 "citation" should be added!
@@
 For example, since the differences in gene expression between tissue types has been pre-calculated, it should be relatively easy to compare differential expression results for comparisons between GTEx tissues produced from GTEx portal and from the EiaD pipeline.
+```
 
+Compare to GTEx (their counts against our counts)
+
+library(tidyverse)
+gtex <- read_tsv('https://storage.googleapis.com/gtex_analysis_v7/rna_seq_data/GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz', skip = 2)
+tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
+metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
+metadata <- metadata %>% mutate(si = str_extract(sample_attribute, "GTEX\\S+"))
+samples <- metadata$si[metadata$si %in% colnames(gtex)]
+genes <- tpm$ID[tpm$ID %in% gtex$Description]
+
+long_tpm <- tpm %>% gather(Sample, TPM, -ID)
+long_gtex <- gtex %>% select(-Name) %>% gather(Sample, Count, -Description)
+
+long_tpm2 <- left_join(long_tpm, metadata %>% select(sample_accession, si), by = c("Sample" = "sample_accession")) %>% filter(si %in% samples)
+
+both <- left_join(long_tpm2 %>% select(-Sample) %>% rename(Sample = si), long_gtex %>% filter(Sample %in% samples) %>% rename(ID=Description), by = c('ID','Sample'))
+
+cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'spearman')
+# spearman 0.91
+cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'pearson')
+# pearson 0.72
+save(both, file = 'gtex_eyeIntegration_compare.Rdata')
+
+# reports pearson cor of TPM of 0.68 to 0.72 between alignment free (salmon/kallisto) methods compared to alignment-based methods (HISAT2)
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6042521/
+
+Compare retina / RPE / cornea markers?
+```
 2.
@@
 Is this an artefact of the scRNA-seq method, or a possible read assignment error from the Salmon tool?
+```
 
+The single cell expression data is in active development.
+Right now we are using the counts quantified by the original research groups.
+As more scRNA-seq datasets come available we will be re-processing the data ourself.
+
+** UPDATE manuscript to make this more explicit. **
+
+```
 This question is also being asked because Salmon is an alignment-free method that performs light weight mapping of reads to transcript models and estimates expression levels of transcripts correcting for potential biases such as GC content.
@@
 It would be helpful to include information about the choice and use of this method in the manuscript.
+```
 
+** UPDATE manuscript to list these limitations **
+
+```
 3.
@@
 It is recommended that this be applied to each tissue type separately, combining samples from the same tissue type, such as retina, across different studies, and adding the study as an indicator in ComBat.
+```
 
+The major way we handle non-biological batch effects is through the use of https://github.com/stephaniehicks/qsmooth, which reshapes the distribution of gene expressions values to be consistent among each tissue type (e.g.
+Retina/RPE/cornea).
+We demonstrate the effect of this tool in **LAST PAPER**
+
+While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some empirical evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).
+
+In these situations, I prefer to do less and let the data "speak" for themselves.
+This is why the primary output of eyeIntegration is a boxplot with the individual data points drawn; this allows the user to get a quick sense of the variance and skew of the data across many datasets.
+
+```
 4.
@@
 Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included.
+```
 
+Parameters selected by checking distributions and picking the parameter that removes the left tail.
+
+Good idea - add those statistics
+
+```
 5.
@@
 This reviewer recommends use of coordinates from other dimensionality reduction methods such as Principal Coordinates Analysis (PCoA) or PCA, or a correlation-based statistic that compares the expression profile of each sample to the profiles of all samples in a given sample set (D statistic) to identify outlier samples, such as described in Nature Genetics 46: 430, 2014.
+```
+
+In principle we agree.
+We were not expecting use the t-SNE parameters so heavily.
+We were surprised in our first paper that the t-SNE worked so well in clustering samples by tissue (though this may be overfitting ** ADD LIMITATION to manuscript**).
+
+We are using the euclidean distance in a *qualitative* manner.
+This "cut-off" distance we select to remove samples is intended to only remove individual samples that are very far away like tissues.
+Nearly all of these outliers are not forming an independent group, but rather are near **other** tissues and thus are likely to be tube mislabels / sample extraction errors.
+Another possibility are samples which had technical issues with sequencing that still passed the basic QC metrics cutoffs.
+
+That said ** TRY THE COOK'S DISTANCE SUGGESTION **
+```
 
@@
 ```
+
+** DO POWER CALCULATION **
+
+library(ssizeRNA)
+library(edgeR)
+library(tidyverse)
+tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
+metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
+
+group <- metadata %>% 
+  filter(sample_accession %in% names(tpm)) %>% 
+  select(sample_accession, Tissue) %>% 
+  unique() %>% 
+  pull(Tissue)
+d <- DGEList(tpm[,2:ncol(tpm)], group = groups)
+d <- calcNormFactors(d)
+d <- estimateCommonDisp(d)
+power <- c()
+for (i in seq(5,100,5)){
+  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 2)$pow_qvalue_ave)}
+
+data %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))
+
+The value of the full GTEx dataset comes in the power to identify expression quantitative trait loci (eQTL).
+For differential gene expression by tissue, far fewer samples are needed.
+A power calculation we ran gives us 80+% power to detect >= 1 log2(FC) gene expression differences with a 20 samples x 20 sample comparision.
+As we have many more eye samples, this is a underestimate of our power when doing eye vs (body) comparisons.
+We have added the power calculation as supplemental figure.

### 2019-05-06 2019_04_22_Reviewer_Comments.Rmd
starting to update manuscript to address reviewer comments

@@
 
-That said ** TRY THE COOK'S DISTANCE SUGGESTION **
+That said ** TRY THE COOK'S DISTANCE SUGGESTIOeyeN **
 ```
@@
   pull(Tissue)
-d <- DGEList(tpm[,2:ncol(tpm)], group = groups)
+d <- DGEList(tpm[,2:ncol(tpm)], group = group)
 d <- calcNormFactors(d)
 d <- estimateCommonDisp(d)
+d <- estimateTagwiseDisp(d)
+
 power <- c()
 for (i in seq(5,100,5)){
-  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 2)$pow_qvalue_ave)}
+  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 5)$pow_qvalue_ave)}
 
-data %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))
+as_tibble(cbind(seq(5,100,5), power)) %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))
 

### 2019-05-06 manuscript.Rmd
starting to update manuscript to address reviewer comments

@@
 have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link.
-This data, as of March 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
+This data, as of May 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
 The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
@@
 ```{r workflowDoodle, echo = F, fig.width = 10, fig.height = 10, results='asis', fig.cap = workflowDoodle_cap}
-workflowDoodle_cap <- fig_cap(name='workflowDoodle', caption = paste0('Raw RNA-seq data from the SRA is run through our pipeline to create the EiaD, which is used by eyeIntegration app to serve interactive gene expression visualizations across ', core_tight_2019 %>% pull(Sub_Tissue) %>% unique() %>% length(), ' tissues'))
+workflowDoodle_cap <- fig_cap(name='workflowDoodle', caption = paste0('Raw RNA-seq data from the SRA is run through our pipeline to create the EiaD, which is used by eyeIntegration app to serve interactive gene expression visualizations across ', core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(Sub_Tissue) %>% unique() %>% length(), ' tissues'))
 
@@
 The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.
-Major markers of retina progenitor, photoreceptors (cone and rod), and retinal ganglia cells (RGC) have simliar gene expression patterns across development in retina fetal tissue and organoids.')
+Major markers of retina progenitor, photoreceptors (cone and rod), and retinal ganglia cells (RGC) have similar gene expression patterns across development in retina fetal tissue and organoids.')
 
@@
 First we showed that the HOXB family is overexpressed in the organoids.
-The homeobox family is well known to initiate polarity of the embryo during early development63 [@philippidouHoxGenesChoreographers2013].
+The homeobox family is well known to initiate polarity of the embryo during early development [@philippidouHoxGenesChoreographers2013].
 Retinoic acid is applied at about day 20 in culture to help differentiate stem cell to organoids and is also known to activate genes members of the HOXB family.

### 2019-05-06 src/power_calc.R

@@ -1,0 +1,26 @@
+# power calculation
+
+library(ssizeRNA)
+library(edgeR)
+library(tidyverse)
+tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
+metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
+
+group <- metadata %>% 
+  filter(sample_accession %in% names(tpm)) %>% 
+  select(sample_accession, Tissue) %>% 
+  unique() %>% 
+  pull(Tissue)
+d <- DGEList(tpm[,2:ncol(tpm)], group = group)
+d <- calcNormFactors(d)
+d <- estimateCommonDisp(d)
+d <- estimateTagwiseDisp(d)
+
+power <- c()
+for (i in seq(5,100,5)){
+  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 5)$pow_qvalue_ave)}
+
+power_data <- as_tibble(cbind(seq(5,100,5), power))
+save(power_data, file = 'data/power_data.Rdata')
+
+# power_data %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))

### 2019-05-07 2019_04_22_Reviewer_Comments.Rmd
adding sup figures, editing tables, small typo fixes

--- 2019-05-04 2019_04_22_Reviewer_Comments.Rmd
@@
 ```
-Move Figure 4 to supplemental figure?
+We have reduce the number of panels in Figure 4 to two and moved the remaining panels to a supplementary figure.
 
-Split Figure 5 into two figures?
+Figure 5 has been split into two separate figures.
 
@@
 
-All quantification of expression happens at the transcript level.
-As most researchers think in terms of gene expression, we collapse all of the transcript level information information into a gene.
-Very briefly, the tximport tool (https://f1000research.com/articles/4-1521/v2) is used to collapse expression from each transcript of a gene is averaged, with scaling for the length of the transcript.
-
-We have
+Supplmental table 2 is unnecessarily confusing.
+There are no differences between the "gene type" whether at the gene or transcript level.
+We have added the Gencode description of the biotypes to the table and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.
 
@@
 
-Number check!
+We originally had 63 tissues - several were removed as they had an insufficient number of samples after quality control. I've updated the text for this figure legend to reflect the final number (58).
 
@@
 ```
-Number check!
+1311 went into the pipeline.
+30 GTEx and 62 eye samples were removed for a variety of QC reasons (see sup table).
+371 + 848 + 30 + 62 = 1311.
+This was made more clear in the text.
+** PUT IN NEW QUOTE AFTER KNITTING WORD DOCUMENTS**
 ```
@@
 ```
-Number check!
+Updated colors.
+
+There are 33 total clustesr. We have added a supplementary table summarizing the Sub Tissue composition of each one.
+
+tsne_50 %>% group_by(Cluster, Sub_Tissue) %>% summarise(Count = n()) %>%  mutate(Sub_Tissue = trimws(Sub_Tissue))  %>% rename(`Sub Tissue` = Sub_Tissue)  %>% flextable::regulartable()
 ```
@@
 
-While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some empirical evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).
+While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).
 
@@
 ```
-
-Parameters selected by checking distributions and picking the parameter that removes the left tail.
-
-Good idea - add those statistics
+Parameters selected by checking distributions and picking the parameter that removes the left tail. We have added mapping rate as a supplementary figure and edited the text in the methods accordingly.
 
@@
 
-In principle we agree.
-We were not expecting use the t-SNE parameters so heavily.
-We were surprised in our first paper that the t-SNE worked so well in clustering samples by tissue (though this may be overfitting ** ADD LIMITATION to manuscript**).
+I wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position.
+While this t-SNE / euclidean distance cutoff approach works on a empirical level, it is a poor idea.
 
-We are using the euclidean distance in a *qualitative* manner.
-This "cut-off" distance we select to remove samples is intended to only remove individual samples that are very far away like tissues.
-Nearly all of these outliers are not forming an independent group, but rather are near **other** tissues and thus are likely to be tube mislabels / sample extraction errors.
-Another possibility are samples which had technical issues with sequencing that still passed the basic QC metrics cutoffs.
-
-That said ** TRY THE COOK'S DISTANCE SUGGESTION **
+We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014).
 ```
@@
 
-** DO POWER CALCULATION **
-
-library(ssizeRNA)
-library(edgeR)
-library(tidyverse)
-tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
-metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
-
-group <- metadata %>% 
-  filter(sample_accession %in% names(tpm)) %>% 
-  select(sample_accession, Tissue) %>% 
-  unique() %>% 
-  pull(Tissue)
-d <- DGEList(tpm[,2:ncol(tpm)], group = groups)
-d <- calcNormFactors(d)
-d <- estimateCommonDisp(d)
-power <- c()
-for (i in seq(5,100,5)){
-  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 2)$pow_qvalue_ave)}
-
-data %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))
-
 The value of the full GTEx dataset comes in the power to identify expression quantitative trait loci (eQTL).
 For differential gene expression by tissue, far fewer samples are needed.
-A power calculation we ran gives us 80+% power to detect >= 1 log2(FC) gene expression differences with a 20 samples x 20 sample comparision.
-As we have many more eye samples, this is a underestimate of our power when doing eye vs (body) comparisons.
+A power calculation we ran gives us 80+% power to detect >= 1 log2(FC) gene expression differences with a 20 sample x 20 sample comparision.
+We would reach a power of 90% with 30 samples, so we have added 10 more samples to each GTEx tissue group.
+As we have many more eye samples, this is likely an underestimate of our power when doing eye vs (body) comparisons.
+
 We have added the power calculation as supplemental figure.
+
+While it would be nice in *principal* to just use the entire GTEx dataset, there are very practical limiations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on.
+Also this amount of storage space brushes against the limits of our compute cluster.
+
+As for the Ratnapriya/Swaroop data, we mention it briefly in our introduction, but we hadn't added the data to eyeIntegration as it became available shortly after our last data pull.
+
+We have surveyed the SRA / GEO again and have added new studies (including the Ratnapriya/Swaroop) project data to eyeIntegration.
+
+As the AMD data they collected are likely to be of interest, we have made an exception to our "normal tissue" policy and have added the different AMD stages as selectable tissue options.

### 2019-05-07 manuscript.Rmd
adding sup figures, editing tables, small typo fixes

@@
 The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling.
-The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands.
+The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data table searching, easy download of core datasets, and local install of the entire app with three commands.
 Additionally, we are prototyping new tools to display single cell RNA-seq (scRNA-seq) data to provide researchers access to cell type specific information about gene expression across murine retinal development.
@@
 We first removed samples with a Salmon calculated mapping rate less than 40%.
-We further filtered samples by calculating the median gene expression for each sample and only kept samples that had a median count >2 across all genes.
+This value was selected as being the far left tail of the distribution of mapping rates across samples (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`).
+We further filtered samples by calculating the median gene expression for each sample and remove outlier samples (data not shown) that had a median count <2 across all genes.
 We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples.
-To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR r-package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
+To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR R package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
 
@@
 
+```{r, echo = F}
+# set color levels
+color <- c('#7fc97f','#beaed4','#ffff99','#fdc086','#386cb0','#f0027f', '#000000', '#000000', '#000000', '#000000')
+names(color) <- c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE', 'GTEx', 'Brain', 'Pituitary', 'Blood', 'Skin')
+custom_fill <- scale_fill_manual(values = color)
+custom_col <- scale_colour_manual(values = color)
+```
+
 ```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
@@
   mutate(Dataset = Version) %>% 
-  ggplot(aes(x=Sub_Tissue, y=Count, fill=Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3)) + scale_colour_manual(values = c('grey','black')) + xlab('') +
+  ggplot(aes(x=Sub_Tissue, y=Count, fill = Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3))  + xlab('') +
   ylab('Count of Samples') + scale_alpha_manual(values = c(0.5,1)) +
   theme(text = element_text(size = 10)) +
-  scale_fill_discrete_sequential(palette = 'viridis')  
+  custom_fill
 ```
@@
 ```{r workflowDoodle, echo = F, fig.width = 10, fig.height = 10, results='asis', fig.cap = workflowDoodle_cap}
-workflowDoodle_cap <- fig_cap(name='workflowDoodle', caption = paste0('Raw RNA-seq data from the SRA is run through our pipeline to create the EiaD, which is used by eyeIntegration app to serve interactive gene expression visualizations across ', core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(Sub_Tissue) %>% unique() %>% length(), ' tissues'))
+workflowDoodle_cap <- fig_cap(name='workflowDoodle', caption = paste0('Raw RNA-seq data from the SRA is run through our pipeline to create the EiaD, which is used by eyeIntegration app to serve interactive gene expression visualizations across ', core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(Sub_Tissue) %>% unique() %>% length() - 1, ' tissues'))
 
@@
 To identify tube mislabeling or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
+After these steps `r core_tight_2019 %>% filter(Kept != 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` eye samples and `r core_tight_2019 %>% filter(Kept != 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples were removed.
 
@@
   theme(text = element_text(size = 30)) +
-  scale_color_discrete_sequential(palette = 'viridis') + 
+  custom_col + 
   guides(colour = guide_legend(override.aes = list(alpha = 1)))
@@
   width(width = c(1,1,1,2)) %>% 
-  flextable::align_text_col(align = "center", header = TRUE) %>% 
+  flextable::align(align = "center",part = "header") %>% 
   flextable::align(align='center') %>% 
@@
 The 'Expression' tab of the webpage provides a wealth of information about both gene- and transcript-level expression for eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`).
-The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues.
+The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then Hugo Gene Nomeclature Committee (HGNC) genes names (or ENSEMBL transcripts), then tissues.
 A boxplot is then generated after hitting the "Re(Draw) Plot" button with overlaid individual data points.
@@
 
+```{r mapping_rate, echo = F, message = F,  warning=F, fig.height = 4, fig.width = 10, fig.pos='H', fig.cap = mapping_rate_cap}
+mapping_rate <- read_delim('data/mapping_rate.txt', delim = ' ', col_names = F) %>% 
+  mutate(X2 = str_sub(X2, 1, 4) %>% as.numeric()) %>%  
+  rename(sample_accession = X1, `Mapping Rate` = X2)
+mapping_stats <- core_tight_2019 %>% left_join(mapping_rate) %>% pull(`Mapping Rate`) %>% summary() %>% broom::tidy()
+mapping_rate_cap <- supFig_cap(name='mapping_rate_cap', caption = paste0('Salmon mapping rate for each sample, grouped by tissue type.
+1st quartile mapping rate is ', mapping_stats$q1, '%, median is ', mapping_stats$median, '%, mean is ', mapping_stats$mean, '%, and third quartile is ', mapping_stats$q3, '%.'))
+core_tight_2019 %>% 
+  left_join(mapping_rate) %>% 
+  ggplot(aes(x=Tissue, y = `Mapping Rate`)) + 
+  ggbeeswarm::geom_quasirandom() + 
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + coord_cartesian(ylim=c(0,100))
+
+```
+
 ```{r snakemakeDAG, echo = F, message = F, warning=F, fig.height = 8, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
@@
 
-colors = viridis(n=6, direction = -1)
-colors[5] <- "red"
+# colors = viridis(n=6, direction = -1)
+# colors[5] <- "red"
 
@@
     geom_point(size=6, alpha=0.1, aes(colour=Group)) +
-    geom_point(size=1, alpha=0.6, position='jitter', aes(shape=Origin)) +
+    geom_point(size=1, alpha=0.8, position='jitter', aes(shape=Origin)) +
     xlab('PCA 1') + ylab('PCA 2') +
@@
     # theme(text = element_text(family = 'Linux Libertine O', size = 10)) +
-    scale_color_manual(values = colors) + 
+    theme_minimal() +
+    custom_col + 
     guides(colour = guide_legend(override.aes = list(alpha = 1)))
@@
   width(width = c(0.5,1,0.5,3)) %>% 
-  flextable::align_text_col(align = "center", header = TRUE) %>% 
+  flextable::align(align = "center", part = 'header') %>% 
   flextable::align(align='center') %>% 
@@
 # Output <div>
+info <- readxl::read_xlsx('data/gencode_type_description.xlsx') %>% fill('Definition') %>% 
+  mutate(Biotype = case_when(Biotype == 'antisense/antisense_RNA' ~ 'antisense',
+                             TRUE ~ Biotype))
 cat("<div id=\"tbl:tx_and_gene_types\">")
@@
   gene_anno %>% group_by(gene_type) %>% summarise(Count=n()) %>% mutate(Type = 'Gene') %>% rename(Class = gene_type))
-regulartable(table %>% group_by(Type) %>% arrange(-Count)) %>% 
-  width(width = c(2,1,2)) %>% 
-  flextable::align_text_col(align = "center", header = TRUE) %>% 
+table <- table %>% 
+  filter(!is.na(Class)) %>% 
+  rename(Biotype = Class) %>% 
+  spread(Type, Count) %>% 
+  rename(`Gene Count` = Gene, `Transcript Count` = Transcript) %>% 
+  left_join(info)
+regulartable(table%>% arrange(Definition)) %>% 
+  width(width = c(2,1,1,4)) %>% 
+ flextable::align(align = "center", part = 'header') %>% 
   flextable::align(align='right') %>% 
@@
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
+flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20) %>% left_join(mapping_rate) %>% rename(Mapping_Rate = `Mapping Rate`)) %>% 
   font(fontname = 'Linux Libertine O', part = 'all') %>% 
   fontsize(size = 6) %>% 
-  width(width = c(0.5,0.5,0.5,3,1,0.5,0.5,0.5,0.5,0.5))
+  width(width = c(0.4,0.4,0.4,2.8,1,0.4,0.4,0.4,0.4,0.4,0.4))
 #Output Caption (* * for italics)
@@
   width(width = c(1,1,1,1,1,2)) %>% 
-  flextable::align_text_col(align = "center", header = TRUE) %>% 
+  flextable::align(align = "center", part = 'header') %>% 
   flextable::align(align='center') %>% 

### 2019-05-07 figures_and_tables/zoom_heatmap_figure.R

@@ -83,6 +83,12 @@ stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
   )
 }
 
+# set color levels
+color <- c('#7fc97f','#beaed4','#ffff99','#fdc086','#386cb0','#f0027f', '#000000', '#000000', '#000000', '#000000')
+names(color) <- c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE', 'GTEx', 'Brain', 'Pituitary', 'Blood', 'Skin')
+custom_fill <- scale_fill_manual(values = color)
+custom_col <- scale_colour_manual(values = color)
+
 zoom_plot <- tsne_50_prep %>% 
   rowwise() %>% 
   mutate(Age = as.numeric(Age_Days)) %>% 
@@ -102,14 +108,15 @@ zoom_plot <- tsne_50_prep %>%
   ggplot(aes(x=X1,y=X2)) +
   scale_shape_manual(values=c(0:20,35:50)) +
   #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
+  stat_chull(data=tsne_zoom %>% 
+               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.6) +
   geom_point(size=2, alpha=1, aes(shape=Origin, colour = Group)) +
   xlab('t-SNE 1') + ylab('t-SNE 2') +
   facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
   geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
-  stat_chull(data=tsne_zoom %>% 
-               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
-  scale_color_discrete_sequential(palette = 'viridis') + 
-  scale_fill_viridis_d(option = 'plasma') +
+  
+  custom_col + 
+  scale_fill_manual(values = c('#3b96e5', '#8ec4f2', '#7be276')) +
   theme_bw() +
   guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
          shape = guide_legend(ncol =2 )) +

### 2019-05-08 manuscript.Rmd
rearrange figures, tweak tables

@@
 B.
-Heatmap visualization.
-C.
-CRX mouse retina gene expression heatmap and table information from Clark et al.
-E11 to P14 scRNA-seq.
-D.
-t-SNE visualization of gene expression profiles of individual cells from Macosko et al.
-E.
-Data table export view.
-F.
-Differential gene expression across different tissues and with GO term enrichment.')
+Heatmap visualization.')
 
@@
 
-Heatmaps based on expression can be drawn for selected genes and tissues and gene expression can be compared across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`).
+Heatmap built by the R package ComplexHeatmaps based on expression can be drawn for selected genes and tissues and gene expression can be compared across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`) [@guComplexHeatmapsReveal2016].
 Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.
@@
 We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues.
-The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).
+The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(supFig_cap(name = 'appScreenshots_sup', display = 'cite'),'D')`).
 As with 'Expression', the user can select which version of the web app to draw data from as well as select for gene- or transcript-level comparisons.
@@
 core_tight_2019 %>% 
-  left_join(mapping_rate) %>% 
+  left_join(mapping_rate, by = 'sample_accession') %>% 
   ggplot(aes(x=Tissue, y = `Mapping Rate`)) + 
@@
 
+```{r appScreenshots_sup, echo = F, message = F, warning=F, fig.height = 8, fig.width = 10, fig.pos='H', fig.cap = appScreenshots_sup}
+appScreenshots_sup <- supFig_cap(name='appScreenshots_sup', caption = 'A.
+CRX mouse retina gene expression heatmap and table information from Clark et al.
+E11 to P14 scRNA-seq.
+B.
+t-SNE visualization of gene expression profiles of individual cells from Macosko et al.
+C.
+Data table export view.
+D.
+Differential gene expression across different tissues and with GO term enrichment.')
+
+knitr::include_graphics('figures_and_tables/eyeIntegration_web_view_01.png')
+```
+
 ```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = rgc_cadherin_hoxb_cap}
@@
   rename(`Gene Count` = Gene, `Transcript Count` = Transcript) %>% 
-  left_join(info)
-regulartable(table%>% arrange(Definition)) %>% 
+  left_join(info, by = 'Biotype')
+regulartable(table %>% arrange(Definition)) %>% 
   width(width = c(2,1,1,4)) %>% 
@@
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20) %>% left_join(mapping_rate) %>% rename(Mapping_Rate = `Mapping Rate`)) %>% 
+flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20) %>% left_join(mapping_rate, by = 'sample_accession') %>% 
+            rename(Mapping_Rate = `Mapping Rate`)) %>% 
   font(fontname = 'Linux Libertine O', part = 'all') %>% 
-  fontsize(size = 6) %>% 
-  width(width = c(0.4,0.4,0.4,2.8,1,0.4,0.4,0.4,0.4,0.4,0.4))
+  fontsize(size = 5) %>% 
+  width(width = c(0.5,0.5,0.5,2,1,0.5,0.5,0.5,0.5,0.5,0.5))
 #Output Caption (* * for italics)
@@
 Tiziana Cogliati was especially helpful in the editing of this manuscript.
+We also would like to thank the two anonymous reviewers for their constructive criticisms.
 Finally, this work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).

### 2019-05-08 figures_and_tables/zoom_heatmap_figure.R

@@ -6,6 +6,32 @@ library(cowplot)
 library(ggrepel)
 library(colorspace)
 
+dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
+tsne_50$Cluster <- dbscan_cluster$cluster
+col1 = 'X1'
+col2 = 'X2'
+# create label points for each cluster
+cluster_centers <- tsne_50 %>%
+  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
+  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))
+
+# samples closest to center in each cluster
+center_samples <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
+  left_join(.,cluster_centers, by=c('Cluster')) %>%
+  mutate(Distance = (!!sym(col1)-C1)+(!!sym(col2)-C2)) %>%
+  group_by(Cluster) %>%
+  dplyr::slice(which.min(Distance)) %>%
+  dplyr::slice(1) %>% 
+  .[['sample_accession']]
+
+# cluster stats
+cluster_stats <- tsne_50 %>% left_join(.,core_tight_2019)  %>%
+  mutate(Origin = gsub('Organoid','Stem Cell', Origin)) %>% 
+  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line'))) %>%
+  mutate(Cluster = as.factor(Cluster)) %>%
+  group_by(Cluster) %>%
+  summarise(Cluster_Tissues = paste(unique(Tissue), collapse=',\n'), Cluster_Counts = paste(n(), ' samples', sep=''))
+
 # set up for ggplot
 tsne_50_prep <- tsne_50 %>% select(-Origin) %>% 
   left_join(., core_tight_2019 %>% select(sample_accession, Origin), by = 'sample_accession') %>% 

### 2019-05-31 manuscript.Rmd

@@
 
-## The GTEx gene expression resource lacks eye-specific tissues
+## The GTEx gene expression web app lacks eye-specific tissues
 The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a comprehensive and user-friendly web app (https://gtexportal.org/); however eye tissues have not been included [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017].
@@
 
-## The eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
+## The eyeIntegration app interactively serves huge GTEx and human eye tissue datasets (EiaD)
 The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017 at version 0.6, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018].
@@
 This value was selected as being the far left tail of the distribution of mapping rates across samples (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`).
-We further filtered samples by calculating the median gene expression for each sample and remove outlier samples (data not shown) that had a median count <2 across all genes.
 We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples.
@@
 
+tsne_50 <- tsne_50 %>% filter(Tissue != 'Choroid Plexus', !(Tissue == 'Retina' & X2 > 20))
 dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)

### 2019-06-01 manuscript.Rmd
update for new metadata

@@
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20) %>% left_join(mapping_rate, by = 'sample_accession') %>% 
-            rename(Mapping_Rate = `Mapping Rate`)) %>% 
+flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
+            #left_join(mapping_rate, by = 'sample_accession') %>% 
+    #        rename(mapping_rate = `Mapping Rate`)) %>% 
   font(fontname = 'Linux Libertine O', part = 'all') %>% 

### 2019-06-01 src/data_pull.R
update for new 04.08 database

@@ -3,8 +3,8 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/inst/app/www/2019/EiaD_human_expression_2019_07.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '~/git/eyeIntegration_app/inst/app/www/2017/eyeIntegration_human_2017_01.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_08.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
 
 
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()

### 2019-06-01 src/retina_development_heatmaps.R

@@ -4,7 +4,7 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_08.sqlite')
 
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'ATOH7','SHH','DLX2', 'POU4F2', 'POU4F3', 'NEFL', 'SNCG')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')

### 2019-06-01 src/tsne_and_pca_calcs.R

@@ -3,7 +3,7 @@ library(pool)
 library(RSQLite)
 library(dbscan)
 library(ggrepel)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/ARC168/eyeIntegration_app/www/2019/EiaD_human_expression_2019_05.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app//www/2019/EiaD_human_expression_2019_05.sqlite')
 core_tight <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble() %>% select(-run_accession) %>% unique()
 anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tpm_gene <- gene_pool_2019 %>% tbl('lsTPM_gene') %>% as_tibble() %>%  spread(sample_accession, value)

### 2019-06-03 manuscript.Rmd
remove mgs from tsne plot

@@
 
-tsne_50 <- tsne_50 %>% filter(Tissue != 'Choroid Plexus', !(Tissue == 'Retina' & X2 > 20))
+tsne_50 <- tsne_50 %>% filter(Tissue != 'Choroid Plexus', !grepl('MGS', Sub_Tissue))
 dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
@@
   left_join(mapping_rate, by = 'sample_accession') %>% 
+  mutate(Tissue = case_when(grepl('MGS', Sub_Tissue) ~ 'Retina MGS',
+                            TRUE ~ Tissue)) %>% 
   ggplot(aes(x=Tissue, y = `Mapping Rate`)) + 

### 2019-06-03 manuscript.Rmd
updated figures with new data, starting to update manuscript code to reflect new gtex and amd samples

@@
 # write supplementary file for metadata
-# write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
+write_csv(core_tight_2019, path = 'supplementary_materials/supplementary_files__metadata.csv')
+# power calculations
+load('data/power_data.Rdata')
 
@@
 
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(Sub_Tissue)) %>% unique() %>% length()) - 1` eye sub-tissue types
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(Sub_Tissue)) %>% unique() %>% length()) - 1` total eye sub-tissue types
 
@@
 
-```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
+```{r tissueCountsBarPlot, fig.height = 5, fig.width = 6, fig.cap=tissueCountsBarPlot_cap, echo=F}
 tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption =  paste0('Substantial increase in eye tissue count and type from 2017 (', core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow(), ', lighter color) to 2019 (', core_tight_2019 %>% filter(Kept=="Kept", study_accession != 'SRP012682') %>% nrow(),', darker color) EiaD.
-We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (teal) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.'))
+We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (orange) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.'))
 
@@
 
-core_both %>% filter(grepl('Retina|Cornea|Lens|RPE', Tissue)) %>% unique() %>% 
+core_both  %>% mutate(Tissue = case_when(study_accession == 'SRP012682' ~ 'GTEx', TRUE ~ Tissue),
+                      Sub_Tissue = case_when(study_accession == 'SRP012682' ~ 'GTEx', TRUE ~ Sub_Tissue),
+                      Sub_Tissue = case_when(grepl('MGS', Sub_Tissue) ~ 'Retina - Adult AMD Tissue', TRUE ~ Sub_Tissue)) %>% 
+                         filter(grepl('GTEx|Retina|Cornea|Lens|RPE', Tissue)) %>% unique() %>% 
     #mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>%
@@
   mutate(Dataset = Version) %>% 
-  ggplot(aes(x=Sub_Tissue, y=Count, fill = Tissue, alpha = Dataset)) + geom_bar(stat = 'identity', position = 'dodge') + theme_minimal() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3))  + xlab('') +
-  ylab('Count of Samples') + scale_alpha_manual(values = c(0.5,1)) +
+  ggplot(aes(x=Sub_Tissue, y= log2(Count), fill = Tissue, alpha = Dataset, label = Count, group = Dataset)) + 
+  geom_bar(stat = 'identity', position = position_dodge(width = NULL)) + 
+  theme_minimal() + 
+  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.3))  + 
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
+  xlab('') + 
+  geom_label_repel(stat = 'identity', position = position_dodge(width = 0.9), size = 4, ylim =c(3, NA)) + 
+  ylab('log2(Count of Samples)') + scale_alpha_manual(values = c(0.5,1)) +
+  #coord_cartesian(ylim = c(0,11)) +
   theme(text = element_text(size = 10)) +
@@
   flextable::align(align='center') %>% 
-  fontsize(size = 10) %>% 
+  flextable::fontsize(size = 10) %>% 
   font(fontname = 'Linux Libertine O', part = 'all')
@@
   font(fontname = 'Linux Libertine O', part = 'all') %>% 
-  fontsize(size = 5) %>% 
+  flextable::fontsize(size = 5) %>% 
   width(width = c(0.5,0.5,0.5,2,1,0.5,0.5,0.5,0.5,0.5,0.5))

### 2019-06-03 figures_and_tables/zoom_heatmap_figure.R

@@ -7,8 +7,7 @@ library(colorspace)
 
 load('data/data_pull.Rdata')
 tsne_50 <- tsne_50 %>% filter(!grepl('MGS', Sub_Tissue)) %>% # remove AMD
-  filter(!(Tissue == 'Retina' & X2 > 20), Tissue != "Choroid Plexus") # remove one retina outlier
-
+  filter(!grepl('MGS', Sub_Tissue), Tissue != "Choroid Plexus", Sub_Tissue != 'Retina - Fetal Eye')
 dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
 tsne_50$Cluster <- dbscan_cluster$cluster
 col1 = 'X1'
@@ -47,9 +46,9 @@ tsne_50_prep <- tsne_50 %>% select(-Origin) %>%
   mutate(Label = paste(Cluster, Cluster_Tissues,sep=': ')) %>%
   mutate(Label = ifelse(sample_accession %in% center_samples, Label, ""))
 
-retina_plus_group <- tsne_50 %>% filter(X1 > -1, X1 < 50, X2 < 10, X2 > -40, Tissue != 'RPE')
+retina_plus_group <- tsne_50 %>% filter(X1 > -30, X1 < -1, X2 < 50, X2 > 10, Tissue != 'RPE')
 #retina_plus_group <- bind_rows(retina_plus_group, tsne_50 %>% filter(X1 > 12))
-dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 5)
+dbscan_cluster <- dbscan(retina_plus_group %>%  select(X1, X2) ,  minPts = 3, eps = 3)
 
 retina_plus_group$Cluster <- dbscan_cluster$cluster
 col1 = 'X1'
@@ -92,8 +91,170 @@ tsne_zoom <- retina_plus_group_prep  %>%
                            TRUE ~ Tissue),
          Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label),
          Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>%
-  filter()
   mutate( zoom = TRUE )
+
+# https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html
+StatChull <- ggproto("StatChull", Stat,
+                     compute_group = function(data, scales) {
[…]

### 2019-06-03 src/data_pull.R

@@ -3,12 +3,12 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_08.sqlite')
-gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/ARC168/eyeIntegration_app/www/2019/EiaD_human_expression_2019_09.sqlite')
+gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/ARC168/eyeIntegration_app/www/2017/eyeIntegration_human_2017_01.sqlite')
 
 
 core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
-core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
+core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% filter(Tissue != 'Choroid Plexus') %>% as_tibble()
 load('data/human_tx_studies.Rdata')
 tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 50) %>% as_tibble()
 gene_anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()

### 2019-06-03 src/retina_development_heatmaps.R

@@ -4,7 +4,7 @@ library(tidyverse)
 library(circlize)
 library(pool)
 library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app/www/2019/EiaD_human_expression_2019_08.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/ARC168/eyeIntegration_app/www/2019/EiaD_human_expression_2019_09.sqlite')
 
 rgc <- c('GAP43', 'POU4F1', 'ISL1', 'ATOH7','SHH','DLX2', 'POU4F2', 'POU4F3', 'NEFL', 'SNCG')
 progenitor <- c('VSX2','SOX2','SOX9','ASCL1','SFRP2','HES1','LHX2','PRTG','LGR5','ZIC1','DLL3','GLI1','FGF19','LIN28B')
@@ -17,7 +17,7 @@ all_markers <- c(rgc, pr, progenitor, cone, rod) %>% unique()
 
 core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
 
-gene <- aman
+#gene <- aman
 
 # function to split by experiment / study / type -------
 plotter_split <- function(gene_vector, annotation = F, breaks = c(0,5,10,15)) {
@@ -242,7 +242,7 @@ plotter_merge <- function(gene_vector, annotation = F, link = NA, breaks = c(0,5
   ht <- Heatmap(log2(y+1), cluster_columns = F, 
                 col = colorRamp2(breaks = breaks, colors = viridis(length(breaks))),
                 clustering_distance_rows = "euclidean", 
-                name = gpar('log(TPM + 1)', fontfamily = 'Linux Libertine O'),
+                name = ('log(TPM + 1)'),
                 #clustering_distance_columns = "euclidean", 
                 #top_annotation = ha_column, 
                 
@@ -263,7 +263,8 @@ dev.off()
 
 
 marker_merge_plot <- plotter_merge(all_markers, annotation = T)
-draw(marker_split_plot +marker_merge_plot, padding = unit(c(15,15,15,15),"mm"))
+
+draw(marker_split_plot + marker_merge_plot, padding = unit(c(15,15,15,15),"mm"))
 
 # most related to gene to each of all_makers by euclidean distance
 # see euc_dist.R

### 2019-06-03 src/tsne_and_pca_calcs.R

@@ -3,7 +3,7 @@ library(pool)
 library(RSQLite)
 library(dbscan)
 library(ggrepel)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/McGaughey_S/eyeIntegration_app//www/2019/EiaD_human_expression_2019_05.sqlite')
+gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/ARC168/eyeIntegration_app//www/2019/EiaD_human_expression_2019_09.sqlite')
 core_tight <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble() %>% select(-run_accession) %>% unique()
 anno <- gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()
 tpm_gene <- gene_pool_2019 %>% tbl('lsTPM_gene') %>% as_tibble() %>%  spread(sample_accession, value)
@@ -16,7 +16,7 @@ eye_samples <- core_tight %>%
   .[['sample_accession']]
 
 kept_samples <- core_tight %>%
-  filter(Kept == 'Kept', Tissue != 'Choroid Plexus') %>% 
+  filter(Kept == 'Kept', Tissue != 'Choroid Plexus', !grepl('MGS', Sub_Tissue)) %>% 
   pull(sample_accession)
 
 

### 2019-06-03 manuscript.Rmd
number updates

@@
 
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(Sub_Tissue)) %>% unique() %>% length()) - 1` total eye sub-tissue types
+## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept', study_accession != 'SRP151763') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r (core_tight_2019 %>% filter(Kept == 'Kept', study_accession == 'SRP151763') %>% pull(sample_accession)) %>% length()` new Retina AMD samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(Sub_Tissue)) %>% unique() %>% length())` and total eye sub-tissue types
 

### 2019-06-04 manuscript.Rmd
new text for gtex 2019 vs 2017, swaroop amd study text added, updated pca text

--- 2019-06-03 manuscript.Rmd
@@
 To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR R package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
+In a change from our previous eyeIntegration work [@Bryan_2018], we now correct our counts for mapping rate and tissue type with the limma batchEffects function.
+Importantly, these transformations are not done for the limma differential expression.
+The transformations are used for the box plot and t-SNE visualizations.
 
@@
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
+We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues (using mapping rate as a covarite), the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
@@
 The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov).
-The app can be deployed on a local computer with 35GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
+The app can be deployed on a local computer with 50GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".
 
@@
 ## scRNA-seq processing
-The eyeIntegration site, as of January 2019, hosts two large scRNA-seq datasets from Macosko et al.
+The eyeIntegration site, as of June 2019, hosts two large scRNA-seq datasets from Macosko et al.
 and Clark et al [@macoskoHighlyParallelGenomewide2015a; @clarkComprehensiveAnalysisRetinal2018].
@@
 
-## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sub-tissue types along with `r ((core_tight_2019 %>% filter(Kept == 'Kept') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(Sub_Tissue)) %>% unique() %>% length()) - 1` total eye sub-tissue types
+## EiaD 2019 contains `r ((core_tight_2019 %>% filter(Kept == 'Kept', study_accession != 'SRP151763') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r (core_tight_2019 %>% filter(Kept == 'Kept', study_accession == 'SRP151763') %>% pull(sample_accession)) %>% length()` new Retina AMD samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(Sub_Tissue)) %>% unique() %>% length())` and total eye sub-tissue types
 
@@
 # set color levels
-color <- c('#7fc97f','#beaed4','#ffff99','#fdc086','#386cb0','#f0027f', '#000000', '#000000', '#000000', '#000000')
-names(color) <- c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE', 'GTEx', 'Brain', 'Pituitary', 'Blood', 'Skin')
+color <- c('#7fc97f','#beaed4','#ffff99','#fdc086','#386cb0','#f0027f', '#000000', '#000000', '#000000', '#000000', '#000000')
+names(color) <- c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE', 'GTEx', 'Brain', 'Pituitary', 'Blood', 'Skin', 'Bone Marrow')
 custom_fill <- scale_fill_manual(values = color)
@@
 tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption =  paste0('Substantial increase in eye tissue count and type from 2017 (', core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow(), ', lighter color) to 2019 (', core_tight_2019 %>% filter(Kept=="Kept", study_accession != 'SRP012682') %>% nrow(),', darker color) EiaD.
-We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (orange) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.'))
+We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (orange) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.
+Counts for each bar plot given in the boxes.
+The y-axis is a log2 transformed sample count.
+'))
 
@@
   custom_fill
+
 ```
@@
 We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodogenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).
+
+We have also added the large retina AMD post-morten Ratnapriya cohort to EiaD 2019 [@ratnapriyaRetinalTranscriptomeEQTL2019].
+This cohort contains hundreds of samples ranging from non-AMD (Minnesota Grading System (MGS) 1) to severe AMD (MGS 4).
+While eyeIntegration is intended to be a source for normal tissues, we have made an exception for this study, as this is a large cohort and AMD is a common disease.
+We did find that we could not group the non-AMD Ratnapriya samples with our other retina samples.
+This may be related to the lower mapping rate of the Ratnapriya et al.
+data (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`).
+
+## `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length())` more GTEx samples and `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length())` new GTEx body sub-tissue types added to 2019 EiaD
+```{r echo = FALSE}
+powerCurve_cap <- supFig_cap(name='powerCurve', caption = 'Power curve to assess ability to detect >= 1 log2(FoldChange) in gene expression between n samples (x-axis) in each group.')
+```
+Our previous dataset for eyeIntegration 0.6 held about 20 samples per GTEx tissue type.
+We ran power calculations to assess our ability to detected >= 1 log2(Fold Change) in gene expression between two conditions to determine whether this is a sufficient number of samples (`r supFig_cap(name='powerCurve', display = 'cite')`).
+Our calculations suggest, for example.
+that we have 83% power to detect a difference in gene expression with two groups of twenty samples.
+To increase our ability to make eye to body comparisons, we added about 10 more samples per GTEx tissue types (which at 30 samples, would give about a 90% power).
+We also took this opportunity to add bladder, bone marrow, cervix uteri, fallopian tube, ovary, prostrate, testis, uterus, and vagina GTEx tissue samples (`r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite'))
 
@@
 We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from all other human body tissues.
-Adult retina is most similar to the brain tissue, pituitary is similar to adult RPE, and skin and blood are similar to cornea.
+Adult retina is most similar to the brain tissue.
+RPE and cornea are most similar to blood, bone marrow, and skin.
 
@@
 
-gtex_tissues <- core_tight_2019 %>% 
-  filter(study_accession == 'SRP012682', Kept == 'Kept') %>%
+gtex_tissues_2017 <- core_tight_2017 %>% 
+  filter(study_accession == 'SRP012682') %>%
+  select(Tissue, study_accession, sample_accession) %>% unique() %>% 
   group_by(Tissue, study_accession) %>% 
@@
 
-gtex <- cbind('GTEx', core_tight_2019 %>% 
+gtex_tissues_2019 <- core_tight_2019 %>% 
+  filter(study_accession == 'SRP012682', Kept == 'Kept') %>%
+  select(Tissue, study_accession, sample_accession) %>% unique() %>% 
+  group_by(Tissue, study_accession) %>% 
+  summarise(Count = n()) %>% 
+  ungroup() %>% 
+  mutate(Tissue = paste0(Tissue, ' (', Count, ')')) %>% 
+  pull(Tissue)
+
+gtex_2017 <- cbind('GTEx', '2017', core_tight_2017 %>% 
+                filter(study_accession == 'SRP012682') %>% nrow(),
+              core_tight_2017 %>% 
+                filter(study_accession == 'SRP012682') %>% nrow(),
+              paste(gtex_tissues_2017, collapse = ', ')) %>% as_tibble()
+
+gtex_2019 <- cbind('GTEx', '2019', core_tight_2019 %>% 
                 filter(study_accession == 'SRP012682') %>% nrow(),
@@
                 filter(study_accession == 'SRP012682', Kept == 'Kept') %>% nrow(),
-              paste(gtex_tissues, collapse = ', ')) %>% as_tibble()
+              paste(gtex_tissues_2019, collapse = ', ')) %>% as_tibble()
 
-colnames(gtex) <- c('Group','Pre QC Count', 'Count', 'Tissue Types (Count)')
+gtex <- rbind(gtex_2017, gtex_2019)
+
+colnames(gtex) <- c('Group', 'Version', 'Pre QC Count', 'Count', 'Tissue Types (Count)')
 # Output <div>
@@
 
+```{r powerCurve, echo = F, results='asis', fig.pos = 'H', fig.cap = powerCurve_cap, warning=F}
+
+ggplot(data = power_data, aes(x=V1, y=power)) + geom_line(size = 1.5) + theme_minimal() + 
+  xlab('Number of samples in each group') + ylab('Power') + scale_x_continuous(breaks = c(20,40,60,80,100))
+
+```
+
 ```{r allSamplePCA, echo = F, message = F, warning=F, fig.height = 8, fig.width = 10, fig.pos='H', fig.cap = pca_cap}
@@
 
-pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples suggests that the non-eye tissue most similar to adult retina is the brain, pituitary is the most similar to adult RPE, and skin and blood are similar to cornea.')
+pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples suggests that the non-eye tissue most similar to adult retina is the brain, RPE and cornea are most similar to bone marrow, blood, and skin')
 
@@
     geom_point(size=1, alpha=0.8, position='jitter', aes(shape=Origin)) +
-    xlab('PCA 1') + ylab('PCA 2') +
+    xlab('PC 1') + ylab('PC 2') +
     theme_bw() +
@@
 theme_set(theme_cowplot(font_size=10)) 
-plot_grid(make_pca_tissue('Brain'), make_pca_tissue('Pituitary'),
+plot_grid(make_pca_tissue('Brain'), make_pca_tissue('Bone Marrow'),
           make_pca_tissue('Blood'), make_pca_tissue('Skin'),
-          labels = c('A: Brain','B: Pituitary','C: Blood','D: Skin'),
+          labels = c('A: Brain','B: Bone Marrow','C: Blood','D: Skin'),
           align = 'hv')
@@
 regulartable(gtex) %>% 
-  width(width = c(0.5,1,0.5,3)) %>% 
+  width(width = c(0.5,0.5, 1,0.5,3)) %>% 
   flextable::align(align = "center", part = 'header') %>% 
   flextable::align(align='center') %>% 
+  flextable::fontsize(size =8) %>% 
   font(fontname = 'Linux Libertine O', part = 'all')

### 2019-06-04 2019_04_22_Reviewer_Comments.Rmd
massive update with text improvemetns, fixing sup tables (including order). updated viewing stats. added new sup figs for swaroop amd data.

@@
 ---
+
+Note to all reviewers:
+Reviewer 2 asked why we had not included the Ratnapriya et al.
+large AMD retina RNA-seq study.
+The reason was because the data was made available after our last data pull.
+While addressing the reviewer criticisms we decided to pull new GTEx samples along with all eye tissue datasets as of April 2019.
+This will change all sample counts in the document.
+We **have** corrected the incorrect code that generated the wrong sample numbers.
 
@@
 ```
-We have reduce the number of panels in Figure 4 to two and moved the remaining panels to a supplementary figure.
-
-Figure 5 has been split into two separate figures.
+We have reduced the number of panels in Figure 4 to two and moved the remaining panels to a new supplementary figure (Supplemental Figure 7).
+We would prefer to keep Figure 5 as one figure, especially as we envisioned this being a fairly short manuscript and having five figures in the main body is ...
+a lot.
 
@@
 
-Supplmental table 2 is unnecessarily confusing.
+Supplemental Table 2 is unnecessarily confusing.
 There are no differences between the "gene type" whether at the gene or transcript level.
-We have added the Gencode description of the biotypes to the table and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.
+We have added the Gencode description of the biotypes to the table (Supplementary Table 2) and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.
 
@@
 
-We originally had 63 tissues - several were removed as they had an insufficient number of samples after quality control. I've updated the text for this figure legend to reflect the final number (58).
+We originally had 63 tissues - several were removed as they had an insufficient number of samples after quality control.
+I've updated the text for this figure legend to reflect the number after removing samples for QC reasons.
 
@@
 ```
-Updated colors.
+The colors were poorly designed. We have updated the color scheme.
 
-There are 33 total clustesr. We have added a supplementary table summarizing the Sub Tissue composition of each one.
+There are 33 total clusters. We have added a supplementary table summarizing the Sub Tissue composition of each one.
 

### 2019-06-04 manuscript.Rmd
massive update with text improvemetns, fixing sup tables (including order). updated viewing stats. added new sup figs for swaroop amd data.

@@
 abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
-METHODS: We downloaded the raw sequnce data for 892 RNA-seq samples across 44 tissues in the GTEx project as a non-eye reference set.
+METHODS: We downloaded the raw sequnce data for 1,375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set.
 We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples.
-The 1311 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 450 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+The 2,291 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
 Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD.
-RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, and 58 tissues.
+RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, 19 eye tissues, and 54 body tissues.
 As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members.
 CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD.
-This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 44 body and 14 eye tissues."
+This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
 keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
@@
 # set up maps
-visitorMap_cap <- supFig_cap(name='visitorMap', caption = paste('As of January 2019, eyeIntegration has had usage across', cities %>% unique() %>% length(), 'cities and', geo_locate$countries %>% unique() %>% length(), 'countries.'))
+visitorMap_cap <- supFig_cap(name='visitorMap', caption = paste('As of June 2019, eyeIntegration has had usage across', cities %>% unique() %>% length(), 'cities and', geo_locate$countries %>% unique() %>% length(), 'countries.'))
 
@@
 To this end, a wide variety of human tissue sources have been used to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids.
-These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014b; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014b; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015a; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
+These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].
 
@@
 have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link.
-This data, as of May 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
+This data, as of June 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
 The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018].
@@
 The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provides consistent sample quality control, and improves efficiency for future sample updates.
-The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling.
+The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over hundreds of new samples and improved sample labeling.
 The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data table searching, easy download of core datasets, and local install of the entire app with three commands.
@@
 We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, induced pluripotent stem cell (iPSC)-derived tissue, stem cell derived organoids, and immortalized cell lines.
-In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project.
+In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues from the GTEx project.
 Using SRA metadata from each study we extracted sample and run accessions, library type, tissue of origin, and sub-tissue of origin.
@@
 
-We generated sample level two-dimensional groupings by with the t-Stochastic Neighbor Embedding (t-SNE) algorithm [@maatenAcceleratingTSNEUsing2014].
-We then calculated the center of the resulting sub-tissue clusters by finding the average two-dimensional coordinates of all samples of a given sub-tissue type (p1).
-We used these sub-tissues groupings to identify outlier samples, or samples that failed to cluster with other samples of the same sub-tissue type, by measuring the euclidean distance of each sample (p2) from its sub-tissue cluster center and removed samples that had distance 4 standard deviations from the average distance with the equation $dist = \sqrt\sum(p1-p2)^{2}$.
+To identify outliers we followed an approach similar to a method in Wright et al [@wrightHeritabilityGenomicsGene2014].
+Briefly, we first selected the 3000 genes with the highest variance across all samples and then for each sub-tissue type $T$ and each sample $i$ in $T$, we first calculated $r_{i}$, the average correlation between $i$ and all other samples in $T$.
+Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all r in T.
+We removed samples with $D_{i} < -17.5$; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.
+The ($ max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold.
 
@@
 The eyeIntegration site, as of June 2019, hosts two large scRNA-seq datasets from Macosko et al.
-and Clark et al [@macoskoHighlyParallelGenomewide2015a; @clarkComprehensiveAnalysisRetinal2018].
-We use the processed gene count data from each group, as well as their cluster assignments which specify what cell type each individual cell is.
+and Clark et al [@macoskoHighlyParallelGenomewide2015a; @clarkSingleCellRNASeqAnalysis2019].
+We use the processed gene count data directly from each group, as well as their cluster assignments which specify what cell type each individual cell is.
 The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration.
@@
 
-## Manuscript as code
+## Power Calculation
+
+We use the ssizeRNA package to calculate power (p) across samples (n) at an FDR of 0.05 [@biSampleSizeCalculation2016].
+Important parameters for ssize RNA include the variability (dispersions for the samples and genes), which were calculated directly from our EiaD length scaled TPM values by the edgeR packages estimateCommonDisp and estimateTagwiseDisp.
+The code to calculate the power is given as 'power_calc.R'.
+
+## Manuscript as code and reproducibility
 This manuscript's figures, tables, and most numbers, are all created and laid out in a R markdown document that interweaves code and text.
@@
 The code that generates this manuscript can be found at https://github.com/davemcg/eyeIntegration_v1_app_manuscript.
+
+The relevant code-bases (https://github.com/davemcg/eyeIntegration_v1_app_manuscript, https://github.com/davemcg/EiaD_build) and the EiaD dataset itself has been deposited into Zenodo with accession 10.5281/zenodo.3238677 to ensure the data can be accessed in the future, even should eyeIntegration and GitHub become inaccessible in the future.
 
@@
 While eyeIntegration is intended to be a source for normal tissues, we have made an exception for this study, as this is a large cohort and AMD is a common disease.
-We did find that we could not group the non-AMD Ratnapriya samples with our other retina samples.
+We found our corrections methods did not group the non-AMD Ratnapriya et al.
+samples with our other collected retina samples (see Retina MGS in `r supFig_cap(name = 'tSNE_sup', display = 'cite')`).
 This may be related to the lower mapping rate of the Ratnapriya et al.
-data (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`).
+data (see Retina MGS in `r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`).
 
-## `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length())` more GTEx samples and `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length())` new GTEx body sub-tissue types added to 2019 EiaD
+## `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% length())` more GTEx samples and `r (core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()) - (core_tight_2017 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()) - 1` new GTEx body sub-tissue types added to 2019 EiaD
 ```{r echo = FALSE}
@@
 To increase our ability to make eye to body comparisons, we added about 10 more samples per GTEx tissue types (which at 30 samples, would give about a 90% power).
-We also took this opportunity to add bladder, bone marrow, cervix uteri, fallopian tube, ovary, prostrate, testis, uterus, and vagina GTEx tissue samples (`r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite'))
+We also took this opportunity to add bladder, bone marrow, cervix uteri, fallopian tube, ovary, prostrate, testis, uterus, and vagina GTEx tissue samples (`r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`)
 
@@
 We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data (`r fig_cap(name='workflowDoodle', display = 'cite')`, `r supFig_cap(name='snakemakeDAG', display = 'cite')`)[@kosterSnakemakeScalableBioinformatics2012].
-To create a high quality final dataset across the `r core_tight_2019 %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`) and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
-To identify tube mislabeling or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
+To create a high quality final dataset across the `r core_tight_2019 %>% nrow()` initial samples (`r supTab_cap(name = 'metadataTable', display = 'cite')`) and 67,315,523,736 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).
+To identify tube mislabeling or sample extraction issues, we used sample-level gene correlation metrics (see methods) to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).
 After these steps `r core_tight_2019 %>% filter(Kept != 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` eye samples and `r core_tight_2019 %>% filter(Kept != 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples were removed.
 
-After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')`).
+After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')` and `r supTab_cap(name = 'tSNE_table', display = 'cite')`)).
 For example, in the retina group, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina samples cluster together.
@@
 ## Murine scRNA-seq enables testing of retina cell type specific expression
-We incorporated scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a].
+We incorporated scRNA-seq data from murine retina across two studies [@clarkSingleCellRNASeqAnalysis2019; @macoskoHighlyParallelGenomewide2015a].
 This allows researchers to quickly examine gene expression across individual cell types in the retina.
@@
 mapping_rate_cap <- supFig_cap(name='mapping_rate_cap', caption = paste0('Salmon mapping rate for each sample, grouped by tissue type.
-1st quartile mapping rate is ', mapping_stats$q1, '%, median is ', mapping_stats$median, '%, mean is ', mapping_stats$mean, '%, and third quartile is ', mapping_stats$q3, '%.'))
+1st quartile mapping rate is ', round(mapping_stats$q1, 1), '%, median is ', round(mapping_stats$median,1), '%, mean is ', round(mapping_stats$mean,1), '%, and third quartile is ', round(mapping_stats$q3,1), '%.'))
 core_tight_2019 %>% 
@@
   ggplot(aes(x=Tissue, y = `Mapping Rate`)) + 
-  ggbeeswarm::geom_quasirandom() + 
+  ggbeeswarm::geom_quasirandom(size = 0.5) + 
   theme_minimal() + 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + coord_cartesian(ylim=c(0,100))
+```
+
+```{r tSNE_sup, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = tSNE_sup_cap}
+tSNE_sup_cap <- supFig_cap(name='tSNE_sup', caption = 'Adult retina samples from Ratnapriya et al.
+(MGS 1 - 4, not-AMD is MGS 1, AMD is MGS 2 through 4) cluster independently from all other adult retina samples collected.')
+
+knitr::include_graphics('figures_and_tables/zoom_heatmap_retina_supplemental.png')
+```
+
+```{r powerCurve, echo = F, results='asis', fig.pos = 'H', fig.cap = powerCurve_cap, warning=F}
+
+ggplot(data = power_data, aes(x=V1, y=power)) + geom_line(size = 1.5) + theme_minimal() + 
+  xlab('Number of samples in each group') + ylab('Power') + scale_x_continuous(breaks = c(20,40,60,80,100))
 
@@
 knitr::include_graphics('figures_and_tables/2019_workflow.png')
-```
-
-```{r powerCurve, echo = F, results='asis', fig.pos = 'H', fig.cap = powerCurve_cap, warning=F}
-
-ggplot(data = power_data, aes(x=V1, y=power)) + geom_line(size = 1.5) + theme_minimal() + 
-  xlab('Number of samples in each group') + ylab('Power') + scale_x_continuous(breaks = c(20,40,60,80,100))
-
 ```
@@
 regulartable(gtex) %>% 
-  width(width = c(0.5,0.5, 1,0.5,3)) %>% 
+  width(width = c(0.9,0.9, 1,0.9,3)) %>% 
   flextable::align(align = "center", part = 'header') %>% 
@@
 regulartable(table %>% arrange(Definition)) %>% 
-  width(width = c(2,1,1,4)) %>% 
+  width(width = c(2,0.8,1,4)) %>% 
   flextable::align(align = "center", part = 'header') %>% 
   flextable::align(align='right') %>% 
+   flextable::fontsize(size =6) %>% 
   font(fontname = 'Linux Libertine O', part = 'all')
@@
 ```{r metadataTable, echo = F,  results='asis', fig.pos = 'H', fig.cap = metadata_cap}
-metadata_cap <- supTab_cap(name='metadataTable', caption = 'Full metadata for 20 random eye samples. Full metadata available as supplementary file.')
+metadata_cap <- supTab_cap(name='metadataTable', caption = 'Full metadata for 10 random eye samples. Full metadata available as supplementary file "metadata.csv."')
 set.seed(1534)
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(20)) %>% 
+flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(10)) %>% 
   #left_join(mapping_rate, by = 'sample_accession') %>% 
@@
   flextable::fontsize(size = 5) %>% 
-  width(width = c(0.5,0.5,0.5,2,1,0.5,0.5,0.5,0.5,0.5,0.5))
+  width(width = c(0.4,0.4,0.4,0.4,0.4,0.4,0.4,1,3,0.4,0.4)) %>% 
+  flextable::height_all(height = 0.1)
 #Output Caption (* * for italics)
@@
 
+```{r tSNE_table, echo = F, results='asis', warning = F, fig.cap = tSNE_table_cap}
+# Output <div>
+cat("<div id=\"tbl:tSNE_table\">")
+tSNE_table_cap <- supTab_cap(name = 'tSNE_table', caption = 'Counts of Sub Tissues in each tSNE - dbscan based cluster group')
+tsne_50_prep %>% 
+  select(Sub_Tissue, sample_accession, Cluster) %>% 
+  mutate(`Sub Tissue` = trimws(Sub_Tissue)) %>% 
+  group_by(`Sub Tissue`, Cluster) %>% 
+  count(name = 'Count') %>% 
+  arrange(Cluster) %>% 
+  regulartable() %>% 
+  width(width = c(2, 1,1)) %>% 
+  flextable::align(align = "center", part = 'header') %>% 
+  flextable::align(align='center') %>% 
+  flextable::fontsize(size =8) %>% 
+  font(fontname = 'Linux Libertine O', part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",tSNE_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
 ```{r GOterm_fetal_retina_organoid, echo = F, results='asis', warning = F, fig.cap = GOterm_cap, fig.pos = 'H'}
-GOterm_cap <- supTab_cap(name='GOterm_fetal_retina_organoid', caption = 'Top GO terms enriched between fetal retina and organoid retina, after paring of redundant terms with REVIGO')
+GOterm_cap <- supTab_cap(name='GOterm_fetal_retina_organoid', caption = 'Top GO terms enriched between fetal retina and organoid retina, after paring of redundant terms with REVIGO [@supekREVIGOSummarizesVisualizes2011]')
 # Output <div>
@@
   font(fontname = 'Linux Libertine O', part = 'all')
-#
+
 #Output Caption (* * for italics)
@@
 Tiziana Cogliati was especially helpful in the editing of this manuscript.
-We also would like to thank the two anonymous reviewers for their constructive criticisms.
+We also would like to thank the two anonymous reviewers for their careful reading and constructive criticisms.
 Finally, this work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).

### 2019-06-04 src/viewing_stats.Rmd
massive update with text improvemetns, fixing sup tables (including order). updated viewing stats. added new sup figs for swaroop amd data.

@@
 ```{r, fig.width = 6, fig.height=2}
+world <- map_data("world")
+world <- world[world$region != "Antarctica",]
 ggplot(data = geo_locate, aes(x = longitudes, y = latitudes, colour = countries)) + 

### 2019-06-04 src/build_figs.sh

@@ -1,0 +1,8 @@
+# makes complicated SVG figs that rely on access to the EiaD sqlite file
+# have to hand convert SVG to png for manuscript
+
+cd ~/git/eyeIntegration_v1_app_manuscript
+Rscript src/data_pull.R
+Rscript src/retina_developmental_heatmaps.R
+Rscript src/tsne_and_pca_calcs.R
+Rscript figures_and_tables/zoom_heatmap_figure.R

### 2019-06-05 manuscript.Rmd
fixed fig 1 legend issue

@@
   xlab('') + 
-  geom_label_repel(stat = 'identity', position = position_dodge(width = 0.9), size = 4, ylim =c(3, NA)) + 
+  geom_label_repel(stat = 'identity', position = position_dodge(width = 0.9), size = 4, ylim =c(3, NA), show.legend = FALSEx) + 
   ylab('log2(Count of Samples)') + scale_alpha_manual(values = c(0.5,1)) +

### 2019-06-06 manuscript.Rmd
text fixes

--- 2019-06-04 manuscript.Rmd
@@
 Recently Ratnapriya et al.
-have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link.
+have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration (AMD) and the GTEx project is providing the data as a download link.
 This data, as of June 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019].
@@
 The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provides consistent sample quality control, and improves efficiency for future sample updates.
-The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over hundreds of new samples and improved sample labeling.
-The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data table searching, easy download of core datasets, and local install of the entire app with three commands.
+The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with hundreds of new samples and improved sample labeling.
+The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data table searching, easy download of core datasets, and local install of the entire interactive resource with three commands.
 Additionally, we are prototyping new tools to display single cell RNA-seq (scRNA-seq) data to provide researchers access to cell type specific information about gene expression across murine retinal development.
@@
 We exhaustively searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'SRA' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a].
+As the SRAdb is being deprecated we also ran searches on the SRA and Gene Expression Omnibus (GEO) web pages with as follows: ((“Homo sapiens"[orgn:__txid9606])  AND (transcriptomic[Source]) AND ("2019/01/01"[Publication Date] : "3000"[Publication Date]) AND (retina[Text Word] OR RPE[Text Word] OR macula[Text Word] OR fovea[Text Word] OR choroid[Text Word] OR sclera[Text Word] OR iris[Text Word] OR lens[Text Word] OR cornea[Text Word] OR ‘trabecular meshwork’[Text Word] OR ‘canals of schlemm’[Text Word] OR ‘cillary body’[Text Word] OR ‘optic nerve’[Text Word] OR ‘laminar cribosa’[Text Word] OR retina[Title] OR RPE[Title] OR macula[Title] OR fovea[Title] OR choroid[Title] OR sclera[Title] OR iris[Title] OR lens[Title] OR cornea[Title] OR ‘trabecular meshwork’[Title] OR ‘canals of schlemm’[Title] OR ‘cillary body’[Title] OR ‘optic nerve’[Title] OR ‘laminar cribosa’[Title] ))".
 We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, induced pluripotent stem cell (iPSC)-derived tissue, stem cell derived organoids, and immortalized cell lines.
@@
 ## Power Calculation
-
-We use the ssizeRNA package to calculate power (p) across samples (n) at an FDR of 0.05 [@biSampleSizeCalculation2016].
+We use the ssizeRNA R package to calculate power (p) across samples (n) at an FDR of 0.05 [@biSampleSizeCalculation2016].
 Important parameters for ssize RNA include the variability (dispersions for the samples and genes), which were calculated directly from our EiaD length scaled TPM values by the edgeR packages estimateCommonDisp and estimateTagwiseDisp.
@@
 Counts for each bar plot given in the boxes.
-The y-axis is a log2 transformed sample count.
+The y-axis is a log2 transformed count of samples pasing our QC filters.
 '))
@@
                       Sub_Tissue = case_when(grepl('MGS', Sub_Tissue) ~ 'Retina - Adult AMD Tissue', TRUE ~ Sub_Tissue)) %>% 
-  filter(grepl('GTEx|Retina|Cornea|Lens|RPE', Tissue)) %>% unique() %>% 
+  filter(grepl('GTEx|Retina|Cornea|Lens|RPE|ESC', Tissue)) %>% unique() %>% 
   #mutate(Sub_Tissue = case_when(sample_accession == 'SRS1955479' ~ 'Retina - Adult Tissue', TRUE ~ Sub_Tissue)) %>%
@@
   xlab('') + 
-  geom_label_repel(stat = 'identity', position = position_dodge(width = 0.9), size = 4, ylim =c(3, NA)) + 
+  geom_label_repel(stat = 'identity', position = position_dodge(width = 0.9), size = 4, ylim =c(3, NA), show.legend = FALSE) + 
   ylab('log2(Count of Samples)') + scale_alpha_manual(values = c(0.5,1)) +
@@
   theme(text = element_text(size = 10)) +
+  scale_y_continuous(breaks = c(0,3,6,9)) +
   custom_fill
@@
 
-We have also added the large retina AMD post-morten Ratnapriya cohort to EiaD 2019 [@ratnapriyaRetinalTranscriptomeEQTL2019].
+We have also added the large retina AMD post-morten Ratnapriya et al.
+cohort to EiaD 2019 [@ratnapriyaRetinalTranscriptomeEQTL2019].
 This cohort contains hundreds of samples ranging from non-AMD (Minnesota Grading System (MGS) 1) to severe AMD (MGS 4).
@@
 ```
-Our previous dataset for eyeIntegration 0.6 held about 20 samples per GTEx tissue type.
+Our previous dataset for eyeIntegration version 0.6 held about 20 samples per GTEx tissue type.
 We ran power calculations to assess our ability to detected >= 1 log2(Fold Change) in gene expression between two conditions to determine whether this is a sufficient number of samples (`r supFig_cap(name='powerCurve', display = 'cite')`).
-Our calculations suggest, for example.
-that we have 83% power to detect a difference in gene expression with two groups of twenty samples.
-To increase our ability to make eye to body comparisons, we added about 10 more samples per GTEx tissue types (which at 30 samples, would give about a 90% power).
-We also took this opportunity to add bladder, bone marrow, cervix uteri, fallopian tube, ovary, prostrate, testis, uterus, and vagina GTEx tissue samples (`r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`)
+Our calculations suggest, for example, that we have 83% power to detect a 1 log2(Fold Change) difference in gene expression with two groups of twenty samples.
+To increase our power to make significant eye to body comparisons, we added about 10 more samples per GTEx tissue types (which at 30 samples, would give about 90% power).
+We also took this opportunity to add bladder, bone marrow, cervix uteri, fallopian tube, ovary, prostrate, testis, uterus, and vagina GTEx tissue samples (`r supTab_cap(name = 'tableSampleCountsGTEx', display = 'cite')`).
 

### 2019-06-06 manuscript.Rmd
text updates

@@
 To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR R package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].
-In a change from our previous eyeIntegration work [@Bryan_2018], we now correct our counts for mapping rate and tissue type with the limma batchEffects function.
+In a change from our previous eyeIntegration work [@Bryan_2018], we now correct our counts for mapping rate and tissue type with the limma batchEffects function [@ritchieLimmaPowersDifferential2015].
 Importantly, these transformations are not done for the limma differential expression.
@@
 Briefly, we first selected the 3000 genes with the highest variance across all samples and then for each sub-tissue type $T$ and each sample $i$ in $T$, we first calculated $r_{i}$, the average correlation between $i$ and all other samples in $T$.
-Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all r in T.
+Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all $r_{i}$ for $i$ in $T$.
 We removed samples with $D_{i} < -17.5$; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.
-The ($ max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold.
+The ($max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold.
 
@@
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
-We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues (using mapping rate as a covarite), the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
+We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues (using mapping rate as a covariate), the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchieLimmaPowersDifferential2015; @lawVoomPrecisionWeights2014].
 We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment.
-The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set f tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
+The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].
 
@@
 
-## EiaD 2019 contains `r ((core_tight_2019 %>% filter(Kept == 'Kept', study_accession != 'SRP151763') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r (core_tight_2019 %>% filter(Kept == 'Kept', study_accession == 'SRP151763') %>% pull(sample_accession)) %>% length()` new Retina AMD samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(Sub_Tissue)) %>% unique() %>% length())` and total eye sub-tissue types
+## EiaD 2019 contains `r ((core_tight_2019 %>% filter(Kept == 'Kept', study_accession != 'SRP151763') %>% pull(study_accession) %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new human eye RNA-seq studies, `r (core_tight_2019 %>% filter(Kept == 'Kept', study_accession == 'SRP151763') %>% pull(sample_accession)) %>% length()` new Retina AMD samples, `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new eye samples, and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682', study_accession != 'SRP151763') %>% pull(Sub_Tissue)) %>% unique() %>% length())`  total eye sub-tissue types
 
@@
 
-Our query on January 7th 2019 to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies.
+Our query on May 8th 2019 to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies.
 We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset.

### 2019-06-06 2019_04_22_Reviewer_Comments.Rmd
updating reviewer comments and making a few corresponding changes to manuscript

@@
 The reason was because the data was made available after our last data pull.
-While addressing the reviewer criticisms we decided to pull new GTEx samples along with all eye tissue datasets as of April 2019.
+While addressing the reviewer criticisms we decided to pull new GTEx samples along with all eye tissue datasets as of May 2019.
 This will change all sample counts in the document.
@@
 
-Supplemental Table 2 is unnecessarily confusing.
+The original design of Supplemental Table 2 was unnecessarily confusing.
 There are no differences between the "gene type" whether at the gene or transcript level.
-We have added the Gencode description of the biotypes to the table (Supplementary Table 2) and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.
+We have reworked the table, adding the Gencode description of the biotypes to the table (Supplementary Table 2, line 400) and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.
 
@@
 I've updated the text for this figure legend to reflect the number after removing samples for QC reasons.
+The new numbers are 19 eye and 54 body tissues (which again, are post-QC).
+There are a few tissue types which entirely drop out after QC.
+For example, we have a small number of eye lid tissues that all failed QC.
 
@@
 ```
-1311 went into the pipeline.
-30 GTEx and 62 eye samples were removed for a variety of QC reasons (see sup table).
-371 + 848 + 30 + 62 = 1311.
-This was made more clear in the text.
-** PUT IN NEW QUOTE AFTER KNITTING WORD DOCUMENTS**
+916 eye and 1,375 GTEx samples now go into the pipeline.
+Table 1 (line 235) gives the counts pre and post-QC for the cornea, ESC, lens, retina, and RPE tissues.
+Eye lid and retinal endothelium were initially included, but all samples failed QC.
+For GTEx were go from 1,375 to 1,314 after QC (Supplemental Table, line 397)
+
 ```
@@
 ```
-The colors were poorly designed. We have updated the color scheme.
+You are correct - the originals colors were poorly designed. We have updated the color scheme. We hope the update is more clear.
 
-There are 33 total clusters. We have added a supplementary table summarizing the Sub Tissue composition of each one.
+There are 37 total clusters (up from 33 in the last submission, as we have some more GTEx tissue types and the AMD retina study).
+We have added a supplementary table (line 404, Supplemental Table 4) summarizing the Sub Tissue composition of each one.
 
-tsne_50 %>% group_by(Cluster, Sub_Tissue) %>% summarise(Count = n()) %>%  mutate(Sub_Tissue = trimws(Sub_Tissue))  %>% rename(`Sub Tissue` = Sub_Tissue)  %>% flextable::regulartable()
 ```
@@
 ```
-HGNC gene names
+Line 208 now reads "The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then Hugo Gene Nomeclature Committee (HGNC) genes names (or ENSEMBL transcripts), then tissues" (it is now line 247).
+We define decile, log2(TPM + 1) in the Figure 4 text (lines 255-258) as follows "Figure 4: Screenshots from eyeIntegration web app.
+A.
+Pan-tissue gene expression box plots with accompanying data tables.
+The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the mean gene’s log2(TPM + 1) score for each sub tissue.
+B.
+Heatmap visualization."
 ```
@@
 
-Add section to suggest ways researchers can faciliate addition of their data to eyeIntegration.
-
-1. Contact us directly (mcgaugheyd@mail.nih.gov) 
-2. Deposit data on SRA
-3. Clear, descriptive, consistent, metadata for each sample (in the sample_attribute section)
+We added a short section to the discussion giving some tips on how to help with future updates to EiaD from lines
 
@@
 7) Many typos and some unfished edit should be corrected, e.g. Line 285 "citation" should be added!
+```
+Fixed the "CITATION" that was missing. We re-read several times and have fixed many typos.
 
+```
 Reviewer #2 (Comments for the Author (Required)):

### 2019-06-06 manuscript.Rmd
updating reviewer comments and making a few corresponding changes to manuscript

@@
 We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples.
-The 2,291 total samples we found were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+The 916 eye and 1,375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
 Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD.
-RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date, by database version, gene/transcript, 19 eye tissues, and 54 body tissues.
+RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date by database version, gene/transcript, 19 eye tissues, and 54 body tissues.
 As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members.
 CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD.
-This offers the eye community a powerful and quick mean to testing hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
+This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
 keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
@@
 In a change from our previous eyeIntegration work [@Bryan_2018], we now correct our counts for mapping rate and tissue type with the limma batchEffects function [@ritchieLimmaPowersDifferential2015].
-Importantly, these transformations are not done for the limma differential expression.
-The transformations are used for the box plot and t-SNE visualizations.
+The transformed values are used for the box plot and t-SNE visualizations.
 
@@
 Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all $r_{i}$ for $i$ in $T$.
-We removed samples with $D_{i} < -17.5$; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.
+We removed samples with $D_{i} <$ -17.5; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.
 The ($max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold.
@@
 ## Differential Gene Expression Analysis and GO term enrichment
-We used the processed data to determine differential gene expression between different sub-tissue types.
+We used the non-transformed length scaled TPM values to determine differential gene expression between different sub-tissue types.
 First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.
@@
 tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption =  paste0('Substantial increase in eye tissue count and type from 2017 (', core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow(), ', lighter color) to 2019 (', core_tight_2019 %>% filter(Kept=="Kept", study_accession != 'SRP012682') %>% nrow(),', darker color) EiaD.
-We also improved the metadata labelling, the cornea samples (yellow) now delineates endothelial and epithelial tissues and the retina samples (orange) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.
+We also improved the metadata labelling, the cornea samples (green) now delineates endothelial and epithelial tissues and the retina samples (orange) distinguish retina organoid and retinal ganglion cell (RGC) from stem cells.
 Counts for each bar plot given in the boxes.
-The y-axis is a log2 transformed count of samples pasing our QC filters.
+The y-axis is a log2 transformed count of samples passing our QC filters.
 '))
@@
   xlab('t-SNE 1') + ylab('t-SNE 2') +
-  geom_label_repel(size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines")) + 
+  geom_label_repel(size = 7, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 5) + 
   theme_bw() +
@@
 ```{r tableSampleCounts, echo = F, results='asis', warning = F, fig.cap = tableSampleCounts_cap}
-tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD contain a huge set of diverse eye tissues')
+tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD contains a large set of diverse eye tissues, including embryonic stem cells (ESC).
+Eye lid and retina endothelium samples were included, but all failed to pass our QC filters.')
 table <- left_join(core_tight_2019  %>% 
-                     filter(study_accession != 'SRP012682',Tissue %in% c('RPE','Lens','Cornea', 'Retina')) %>% 
+                     filter(study_accession != 'SRP012682') %>% 
                      group_by(Tissue) %>% 
@@
                    core_tight_2019  %>%
-                     filter(study_accession != 'SRP012682', Tissue %in% c('RPE','Lens','Cornea', 'Retina'), Kept == 'Kept') %>% 
+                     filter(study_accession != 'SRP012682', Kept == 'Kept') %>% 
                      rowwise() %>% 
@@
                      summarise(`Count` = sum(Count), `Sub-Tissue Types (Count)` = paste(sort(unique(trimws(Sub))), collapse=', ')), by = 'Tissue') %>% 
-  as_tibble()
+  as_tibble() %>% 
+  mutate(Tissue = case_when(Tissue == 'EyeLid' ~ 'Eye lid', TRUE ~ Tissue)) %>% 
+  mutate(Count = case_when(is.na(Count) ~ 0L, TRUE ~ Count))
 
@@
 Pan-tissue gene expression box plots with accompanying data tables.
+The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the mean gene\'s log2(TPM + 1) score for each sub tissue.
 B.
@@
 There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell derived organoids.
-We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.
+We looked at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.
 
@@
 To identify specific changes between retinal organoid and fetal retina tissue, we performed differential gene expression and GO term enrichment analyses.
-The GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets in retinal organoids As there is some evidence suggesting that protocadherins influence RGC viability CITATION and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression [@lefebvreGProtocadherinsRegulateNeuronal2008].
+The GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets in retinal organoids As there is some evidence suggesting that protocadherins influence RGC viability and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression [@lefebvreGProtocadherinsRegulateNeuronal2008].
 
@@
 Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].
+
+If you wish to have your data added to EiaD in the future, we suggest you 1.
+deposit data into GEO/SRA, 2.
+use clear, descriptive, consistent, and detailed metadata for each sample, and 3.
+(optional) contact the corresponding author.
+Contacting the corresponding author is only necessary if you feel your data should be included in EiaD and was deposited into the SRA before May 8th, 2019.
 
@@
 cat("<div id=\"tbl:metadataTable\">")
-flextable(core_tight_2019 %>% filter(study_accession != 'SRP012682') %>% sample_n(10)) %>% 
+table_ct_2019 <- core_tight_2019
+colnames(table_ct_2019) <- gsub('_', ' ', colnames(table_ct_2019))
+flextable(table_ct_2019 %>% filter(`study accession` != 'SRP012682') %>% sample_n(10)) %>% 
   #left_join(mapping_rate, by = 'sample_accession') %>% 
@@
   flextable::fontsize(size = 5) %>% 
-  width(width = c(0.4,0.4,0.4,0.4,0.4,0.4,0.4,1,3,0.4,0.4)) %>% 
+   flextable::fontsize(size = 5, part = 'header') %>% 
+  width(width = c(0.5,0.4,0.4,0.4,0.3,0.3,0.6,1,3,0.45,0.4)) %>% 
   flextable::height_all(height = 0.1)

### 2019-06-07 2019_04_22_Reviewer_Comments.Rmd
updated comments, added some sections to the manuscript, new citations

@@
 ```
+
 Line 208 now reads "The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then Hugo Gene Nomeclature Committee (HGNC) genes names (or ENSEMBL transcripts), then tissues" (it is now line 247).
@@
 
-We added a short section to the discussion giving some tips on how to help with future updates to EiaD from lines
+We added a short section to the discussion giving some tips on how to help with future updates to EiaD from lines 348-351: "If you wish to have your data added to EiaD in the future, we suggest you 1.
+deposit data into GEO/SRA, 2.
+use clear, descriptive, consistent, and detailed metadata for each sample, and 3.
+(optional) contact the corresponding author.
+Contacting the corresponding author is only necessary if you feel your data should be included in EiaD and was deposited into the SRA before May 8th, 2019."
 
@@
 
-Compare to GTEx (their counts against our counts)
+To address this concern about the accuracy of our process we directly compared the GTEx provided count table against the counts we calculated for the identical GTEx samples.
+The code below can be run in R to re-produce our finding: which is that we have a pearson (R^2) correlation of 0.89
 
+```
+# R code
+# need about 64GB of memory!
 library(tidyverse)
 gtex <- read_tsv('https://storage.googleapis.com/gtex_analysis_v7/rna_seq_data/GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz', skip = 2)
-tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
-metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')
+tpm <- read_tsv('/data/mcgaugheyd/datashare/eyeIntegration/2019_gene_TPM_04.tsv.gz')
+metadata <- read_tsv('/data/mcgaugheyd/datashare/eyeIntegration/2019_metadata_04.tsv.gz')
 metadata <- metadata %>% mutate(si = str_extract(sample_attribute, "GTEX\\S+"))
@@
 genes <- tpm$ID[tpm$ID %in% gtex$Description]
+# make data long to join GTEx and EiaD results
+long_tpm <- tpm %>% gather(Sample, TPM, -ID)
+long_gtex <- gtex %>% select(-Name) %>% gather(Sample, GTEX_TPM, -Description)
+# filter EiaD tpm down to just GTEx samples
+long_tpm2 <- left_join(long_tpm, metadata %>% select(sample_accession, si) %>% unique(), by = c("Sample" = "sample_accession")) %>% filter(si %in% samples)
+# merge EiaD and GTEx
+both <- left_join(long_tpm2 %>% select(-Sample) %>% rename(Sample = si), long_gtex %>% filter(Sample %in% samples) %>% rename(ID=Description), by = c('ID','Sample'))
+# remove NA
+both2 <- both %>% filter(!is.na(TPM), !is.na(GTEX_TPM))
+# calculate correlations with log2(TPM + 0.01 pseudocount) as per Zhang et al. 
+cor(log2(both2 %>% pull(TPM) + 0.01), log2(both2 %>% pull(GTEX_TPM) + 0.01), method = 'spearman')
+# spearman 0.91
+cor(log2(both2 %>% pull(TPM) + 0.01), log2(both2 %>% pull(GTEX_TPM) + 0.01), method = 'pearson')
+# pearson 0.89
+```
 
-long_tpm <- tpm %>% gather(Sample, TPM, -ID)
-long_gtex <- gtex %>% select(-Name) %>% gather(Sample, Count, -Description)
+Zhang et al.
+(https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4002-1) report a pearson cor R^2 with TPM of 0.89 and 0.93 between alignment free (salmon/kallisto) methods compared to alignment-based methods (STAR/REM) [@zhangEvaluationComparisonComputational2017].
+As GTEx used an alignment-based approach (and a different reference genome - we use GRCh38, GTEx uses GRCh37) and we length scale our TPM values (GTEx does note) we feel that this is a reasonable result.
 
-long_tpm2 <- left_join(long_tpm, metadata %>% select(sample_accession, si), by = c("Sample" = "sample_accession")) %>% filter(si %in% samples)
+We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-233).
 
-both <- left_join(long_tpm2 %>% select(-Sample) %>% rename(Sample = si), long_gtex %>% filter(Sample %in% samples) %>% rename(ID=Description), by = c('ID','Sample'))
-
-cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'spearman')
-# spearman 0.91
-cor(both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(TPM), both %>% filter(!is.na(TPM), !is.na(Count)) %>% pull(Count), method = 'pearson')
-# pearson 0.72
-save(both, file = 'gtex_eyeIntegration_compare.Rdata')
-
-# reports pearson cor of TPM of 0.68 to 0.72 between alignment free (salmon/kallisto) methods compared to alignment-based methods (HISAT2)
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6042521/
-
-Compare retina / RPE / cornea markers?
 ```
@@
 
-The single cell expression data is in active development.
+The single cell expression section of eyeIntegration is still in active development.
 Right now we are using the counts quantified by the original research groups.
 As more scRNA-seq datasets come available we will be re-processing the data ourself.
+The high rhodopsin expression is most likely because (some) rods are being incorrectly labelled as cones by the original authors.
 
-** UPDATE manuscript to make this more explicit. **
+The methods section for scRNA-seq procesesing reads "We use the processed gene count data directly from each group, as well as their cluster assignments which specify what cell type each individual cell is.
+The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration."
 
@@
 
-** UPDATE manuscript to list these limitations **
+We have added a "limitations" section to the resulst section of the manuscript:
+
+"Salmon quantification, while highly performant and accurate, has a higher variance when with lower read depth and with shorter transcripts [@zhangEvaluationComparisonComputational2017].
+Extra care should be taken with comparisons with lower counts of samples (cornea, RGC) as smaller sample numbers decrease the confidence in differential expression.
+We recommend that if you wish to directly compare your RNA-seq sample quantifications with EiaD, you should not directly compare our TPM values with yours as there are many important variables that will differ.
+Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build) and add your samples.
+Finally, we would like to remind any users that RNA-seq methods measure mRNA levels, but the functional unit is the protein; westerns are still the gold standard with which to evaluate expression and localization."
 
@@
 Retina/RPE/cornea).
-We demonstrate the effect of this tool in **LAST PAPER**
 
-While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).
+While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (see https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/) [@nygaardMethodsThatRemove2016].
 
-In these situations, I prefer to do less and let the data "speak" for themselves.
-This is why the primary output of eyeIntegration is a boxplot with the individual data points drawn; this allows the user to get a quick sense of the variance and skew of the data across many datasets.
+We have made a significant change to how our TPM values are computed, using limma's batchEffects() tool to correct the TPM values for mapping rate and sub tissue.
+These two changes clean up the t-SNE plotting for the eye tissues.
+The major difference we see within a single sub tissue type is adult retina - for some reason we don't understand the Ratnapriya / Swaroop retina AMD samples (both non-AMD and AMD) cluster apart from all of the other adult retina tissue samples we have collected.
+This may be related to the lower mapping rate (and RIN scores) from this cohort.
+
+We do not use the tranformed TPM values for the limma / voom differential testing. These transformed values are only used the the t-SNE, boxplots, and the heatmap visualizations.
 
@@
 ```
-Parameters selected by checking distributions and picking the parameter that removes the left tail. We have added mapping rate as a supplementary figure and edited the text in the methods accordingly.
+Parameters were selected by checking distributions and picking the parameter that removes the left tail.
+We have added mapping rate as a supplementary figure and edited the text in the methods accordingly.
+Line 111-112: "We first removed samples with a Salmon calculated mapping rate less than 40%.
+This value was selected as being the far left tail of the distribution of mapping rates across samples (Supplemental Figure 2)."
 
@@
 
-I wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position.
+I first wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position.
 While this t-SNE / euclidean distance cutoff approach works on a empirical level, it is a poor idea.
 
-We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014).
+We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014) (lines 119-125):
+
+"To identify outliers we followed an approach similar to a method in Wright et al [@wrightHeritabilityGenomicsGene2014].
+Briefly, we first selected the 3000 genes with the highest variance across all samples and then for each sub-tissue type $T$ and each sample $i$ in $T$, we first calculated $r_{i}$, the average correlation between $i$ and all other samples in $T$.
+Next, we calculated $D_{i}$, where $D_{i} = \frac{(r_{i} - r\bar{\bar{}})}{median(r_{i} - r\bar{\bar{}})}$ and  $r\bar{\bar{}}$ is the grand mean of all $r_{i}$ for $i$ in $T$.
+We removed samples with $D_{i} <$ -17.5; we determined this threshold by generating a tSNE plot of our samples, and visually identifying outliers in adult retina tissue.
+The ($max(D_{i})$) amongst these outliers was -17.58 and from this we chose -17.5 as our outlier threshold."
 ```
@@
 
-We have added the power calculation as supplemental figure.
+We have added the power calculation curve as supplemental figure.
 
-While it would be nice in *principal* to just use the entire GTEx dataset, there are very practical limiations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on.
-Also this amount of storage space brushes against the limits of our compute cluster.
-
-As for the Ratnapriya/Swaroop data, we mention it briefly in our introduction, but we hadn't added the data to eyeIntegration as it became available shortly after our last data pull.
-
-We have surveyed the SRA / GEO again and have added new studies (including the Ratnapriya/Swaroop) project data to eyeIntegration.
-
-As the AMD data they collected are likely to be of interest, we have made an exception to our "normal tissue" policy and have added the different AMD stages as selectable tissue options.
+While it would be nice in principle to just use the entire GTEx dataset, there are very practical limitations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on.
+Also this amount of storage space brushes against the limits of our allocation in the compute cluster.
+As for the Ratnapriya/Swaroop data, we have now added it to EiaD along with a few other smaller studies which were made available after our last EiaD update in January 2019.

### 2019-06-07 manuscript.Rmd
updated comments, added some sections to the manuscript, new citations

@@
 abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
-METHODS: We downloaded the raw sequnce data for 1,375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set.
+METHODS: We downloaded the raw sequnce data for 1375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set.
 We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples.
-The 916 eye and 1,375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
+The 916 eye and 1375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset.
 Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD.
@@
 
+To calculate pearson correlation ($R^2$) between GTEx-calculated TPM gene values and our GTEx TPM gene values, we downloaded "GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_tpm.gct.gz" and matched against our GTEx TPM values, running the pearson correlation with $log2(TPM + 0.01)$ values as per Zhang et al.
+with the cor function in R [@zhangEvaluationComparisonComputational2017].
+
 ## Differential Gene Expression Analysis and GO term enrichment
@@
 
-After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')` and `r supTab_cap(name = 'tSNE_table', display = 'cite')`)).
+To ensure there are no substantial differences in quantification of the gene TPM, we calculated the $R^2$ between the GTEX-calculated TPM values and our length scaled and transformed TPM values we calculate for our GTEx samples (see methods).
+We got an $R^2$ of 0.89.
+Zhang et al.
+report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a pearson correlation ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017].
+As Zhang et al.
+compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.
+
+After our quality control and processing workflow, we found that samples of the same tissue type and origin cluster well together (`r fig_cap(name = 'tSNE', display = 'cite')` and `r supTab_cap(name = 'tSNE_table', display = 'cite')`).
 For example, in the retina group, primary adult tissue clusters tightly and distinctly from other cell types, and retinal organoids and fetal retina samples cluster together.
@@
 We also plotted the differential expression values between all organoids and all fetal retina samples; all genes across all three groups are significantly differentially expressed with an FDR corrected p value < 0.01.
+
+## Limitations of the RNA-seq quantification in eyeIntegration
+
+Salmon quantification, while highly performant and accurate, has a higher variance when with lower read depth and with shorter transcripts [@zhangEvaluationComparisonComputational2017].
+Extra care should be taken with comparisons with lower counts of samples (cornea, RGC) as smaller sample numbers decrease the confidence in differential expression.
+We recommend that if you wish to directly compare your RNA-seq sample quantifications with EiaD, you should not directly compare our TPM values with yours as there are many important variables that will differ.
+Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build) and add your samples.
+Finally, we would like to remind any users that RNA-seq methods measure mRNA levels, but the functional unit is the protein; westerns are still the gold standard with which to evaluate expression and localization.
 

### 2019-06-08 manuscript.Rmd
fixed vinny comments

@@
 
-To ensure there are no substantial differences in quantification of the gene TPM, we calculated the $R^2$ between the GTEX-calculated TPM values and our length scaled and transformed TPM values we calculate for our GTEx samples (see methods).
-We got an $R^2$ of 0.89.
-Zhang et al.
-report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a pearson correlation ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017].
+To ensure there are no substantial differences in quantification of gene TPM values, we determined the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89.
+“ Zhang et al.
+report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017].
 As Zhang et al.
@@
 Pan-tissue gene expression box plots with accompanying data tables.
-The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the mean gene\'s log2(TPM + 1) score for each sub tissue.
+The data tables display the rank (lower is more highly expressed) of each gene in each sub tissue, the decile of the rank (10 is the highest decile of expression), and the gene\'s mean log2(TPM + 1) score for each sub tissue.
 B.
@@
 
-Salmon quantification, while highly performant and accurate, has a higher variance when with lower read depth and with shorter transcripts [@zhangEvaluationComparisonComputational2017].
+Salmon quantification, while highly performant and accurate, has a higher variance for lower read depth samples and shorter transcripts [@zhangEvaluationComparisonComputational2017].
 Extra care should be taken with comparisons with lower counts of samples (cornea, RGC) as smaller sample numbers decrease the confidence in differential expression.
-We recommend that if you wish to directly compare your RNA-seq sample quantifications with EiaD, you should not directly compare our TPM values with yours as there are many important variables that will differ.
-Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build) and add your samples.
+We do not recommend you directly compare our TPM values with your counts data as there are many important variables that will differ.
+Instead run our Snakemake pipeline (https://www.github.com/davemcg/EiaD_build), adding your samples.
 Finally, we would like to remind any users that RNA-seq methods measure mRNA levels, but the functional unit is the protein; westerns are still the gold standard with which to evaluate expression and localization.
@@
 All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’).
-The code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.
+The code for the EiaD data processing pipeline can be found at https://github.com/davemcg/EiaD_build.
 

### 2019-06-10 2019_04_22_Reviewer_Comments.Rmd
re-submission commit!

@@
 Reviewer 2 asked why we had not included the Ratnapriya et al.
-large AMD retina RNA-seq study.
+large AMD retina RNA-seq study [@ratnapriyaRetinalTranscriptomeEQTL2019].
 The reason was because the data was made available after our last data pull.
@@
 
+To ensure there are no substantial differences in quantification of gene TPM values, we calculated the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89.
 Zhang et al.
-(https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4002-1) report a pearson cor R^2 with TPM of 0.89 and 0.93 between alignment free (salmon/kallisto) methods compared to alignment-based methods (STAR/REM) [@zhangEvaluationComparisonComputational2017].
-As GTEx used an alignment-based approach (and a different reference genome - we use GRCh38, GTEx uses GRCh37) and we length scale our TPM values (GTEx does note) we feel that this is a reasonable result.
+report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017].
+As Zhang et al.
+compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.
 
-We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-233).
+We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-232).
 
@@
 
-We have added a "limitations" section to the resulst section of the manuscript:
+We have added a "limitations" section to the results section of the manuscript:
 

### 2019-06-10 manuscript.Rmd
re-submission commit!

@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
+abstract: "PURPOSE: To develop an accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.
 METHODS: We downloaded the raw sequnce data for 1375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set.
@@
 
-To ensure there are no substantial differences in quantification of gene TPM values, we determined the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89.
-“ Zhang et al.
+To ensure there are no substantial differences in quantification of gene TPM values, we calculated the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89.
+Zhang et al.
 report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017].

### 2019-06-10 src/age_prediction.R

@@ -1,79 +1,0 @@
-
-library(tidyverse)
-
-library(pool)
-library(RSQLite)
-gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/Arges/eyeIntegration_app/www/2019/EiaD_human_expression_2019_03.sqlite')
-
-
-###################
-
-# build model to predict fetal tissue age
-# apply to organoid 
-# works OK on fetal tissue, but doesn't 
-# work well with organoid
-
-###################
-tissue <- bind_rows(ESC, fetal_tissue, adult_tissue)
-
-query = 'select * from lsTPM_gene'
-p <- dbGetQuery(gene_pool_2019, query) %>% left_join(.,core_tight_2019) %>% 
-  left_join(., gene_pool_2019 %>% tbl('gene_IDs') %>% as_tibble()) %>% 
-  as_tibble()
-fetalTissue_allGene <- p %>% 
-  filter(gene_type == 'protein_coding') %>% 
-  filter(Sub_Tissue == 'Retina - Fetal Tissue' | study_accession == 'SRP159246') %>% 
-  select(ID, sample_accession, Age_Days, value) %>% 
-  mutate(age_accession = paste0(Age_Days, sep='_', sample_accession)) %>% 
-  select(-sample_accession, -Age_Days) %>% 
-  unique()
-
-# set up train test
-age_access <- fetalTissue_allGene$age_accession %>% unique() %>% sort()
-train_samp <- age_access[seq(1, length(age_access), 2)]
-train <- fetalTissue_allGene %>% filter(age_accession %in% train_samp)
-test <- fetalTissue_allGene %>% filter(!age_accession %in% train_samp)
-
-# mean summarise multiple overlappping time points
-train$age <- as.numeric(gsub('_.*','', train$age_accession))
-train <- train %>% group_by(ID, age) %>% summarise(value = mean(value))
-
-trainer <- train %>% spread(ID, value) 
-tester <- test %>% spread(ID, value)
-
-# caret
-library(caret)
-library(xgboost)
-set.seed(96)
-
-# rf_fit <- train(x = trainer %>% select(-age) %>% as.matrix(), 
-#                 y = trainer$age,
-#                 method = "rf")
-xgb <- xgboost(data = trainer %>% select(-age) %>% as.matrix(), 
-               label = trainer$age,
-               nrounds = 50)
-
-# predict on test
-## rf
[…]

### 2019-10-07 src/viewing_stats.Rmd
updated plot, now collects info by date

--- 2019-01-23 src/viewing_stats.Rmd
@@
 library(ipstack)
-ipaddresses <- read_delim('~/Desktop/stats/eyeIntegration.stats', col_names = F, delim = '\\s+') %>% 
+ipaddresses <- read_delim('~/Desktop/email_eyeintegration_//eyeIntegration.stats', col_names = F, delim = '\\s+') %>% 
   separate(col = X1, into = c('IP Address', 'Page Views'), sep='\\s+') %>% 
@@
 
-Total page view from May 2017 to November 2017
 ```{r}
-sum(ipaddresses$`Page Views`)
+files <- list.files('~/Desktop/email_eyeintegration_/', pattern = '*eml', full.names = TRUE)
+eml <- files %>% map(read_lines) 
+dates <- eml %>% 
+  map(grep, pattern = 'Date:', value =TRUE) %>% 
+  map(gsub, pattern = 'Date: \\w+, | \\d+:.*', replacement = '') %>% 
+  map(as.Date, format = "%d %b %Y") %>% 
+  map(as.character) %>% 
+  unlist() %>% 
+  as.Date()
+ip <- eml %>% 
+  map(grep, pattern = '^[[:digit:]]', value = TRUE) %>% 
+  map(str_split, pattern = '\\s+') %>% 
+  set_names(dates)
+
+empty_list <- ip %>% map(length) %>% unlist()
+empty_list <- empty_list != 0
+ip <- ip[empty_list]
+
+ip <- ip %>% map(enframe) %>% map(unnest_wider, value) %>% bind_rows(.id = 'Date')
+colnames(ip) <- c('Date', 'Name', 'IP', 'Pages')
+ip$Pages <- as.numeric(ip$Pages)
+```
+
+Total page views from May 2017 to Today
+```{r}
+sum(ip$Pages)
 ```
@@
 ```{r}
-unique_IP <- ipaddresses$`IP Address` %>% unique()
+unique_IP <- ip$IP %>% unique()
 ip_info <- list()
@@
 countries <- ''
+hostname <- ''
 for (i in unique_IP){
@@
   ips <- c(ips, ip_info[[i]]$ip)
+  hostname <- c(hostname, ip_info[[i]]$hostname)
   latitudes <- c(latitudes, ip_info[[i]]$latitude)
@@
   scale_color_viridis_d()
+```
 
+Page View over Time
+```{r}
+ip$Date <- as.Date(ip$Date)
+ip %>% 
+  filter(Pages > 11) %>% 
+  arrange(Date) %>% 
+  ggplot(aes(x=Date, y = cumsum(Pages)/10)) + 
+  geom_line() + cowplot::theme_cowplot()  + ylab("Cumulative Queries")
 ```
